---
title: "Hypoxia Peak Reduced Report"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(ggplot2)
library(dplyr)
library(scales)
library(knitr)

report_file <- knitr::current_input(dir = TRUE)
report_dir <- if (nzchar(report_file)) dirname(report_file) else file.path("dev", "hypoxia_peak_reduced")
out_dir <- file.path(report_dir, "output")
project_root <- normalizePath(file.path(report_dir, "..", ".."), winslash = "/", mustWork = TRUE)

summary_tbl <- read.csv(
  file.path(out_dir, "sample_peak_summary.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

input_tbl <- read.csv(
  file.path(out_dir, "prepared_input_table.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

ablation_tbl <- read.csv(
  file.path(out_dir, "ablations", "ablation_comparison.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

extension_tbl <- read.csv(
  file.path(out_dir, "extensions", "extension_comparison.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

karyotype_lookup_tbl <- read.csv(
  file.path(out_dir, "karyotype_flow_lookup.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

karyotype_ploidy_tbl <- read.csv(
  file.path(out_dir, "karyotype_ploidy_summary.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

adjacent_8n_tbl <- read.csv(
  file.path(out_dir, "extensions", "adjacent_8n", "sample_peak_summary.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

adjacent_8n_delta_tbl <- read.csv(
  file.path(out_dir, "extensions", "adjacent_8n_delta_dna", "sample_peak_summary.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

dna_vectors <- readRDS(
  file.path(project_root, "processed_data", "hypoxia-sum159", "filtered_dna_area_vectors.rds")
)$raw

sample_order_tbl <- summary_tbl %>%
  arrange(as.Date(latest_match_date), sample_name) %>%
  transmute(
    sample_name,
    latest_match_date,
    sample_label = sprintf(
      "%s\n%s",
      sub("^Sample_SUM159_", "", sample_name),
      latest_match_date
    )
  )

sample_levels <- sample_order_tbl$sample_name
sample_label_map <- setNames(sample_order_tbl$sample_label, sample_order_tbl$sample_name)
outlier_palette <- c("Primary outlier" = "#cb181d", "Not primary outlier" = "#2171b5")

state_flip_tbl <- summary_tbl %>%
  select(sample_name, condition, latest_match_date, relative_day, observed_g1_peak, baseline_state = assigned_state) %>%
  inner_join(
    adjacent_8n_tbl %>%
      select(sample_name, adjacent_8n_state = assigned_state, adjacent_8n_counterfactual = phi1_counterfactual_peak),
    by = "sample_name"
  ) %>%
  inner_join(
    adjacent_8n_delta_tbl %>%
      select(sample_name, adjacent_8n_delta_state = assigned_state, adjacent_8n_delta_counterfactual = phi1_counterfactual_peak),
    by = "sample_name"
  ) %>%
  filter(
    baseline_state != adjacent_8n_state |
      baseline_state != adjacent_8n_delta_state
  ) %>%
  arrange(relative_day, sample_name)

plot_burden_vs_peak <- function(data, y_var, y_label, title) {
  ggplot(data, aes(x = ratio_above_below, y = .data[[y_var]])) +
    geom_point(aes(color = latest_match_date), size = 2.4, alpha = 0.9) +
    labs(
      title = title,
      x = "Burden proxy: ratio above / below",
      y = y_label,
      color = "Sample date"
    ) +
    theme_bw(base_size = 10) +
    theme(panel.grid.minor = element_blank())
}

plot_peak_ratio_by_sample <- function(summary_tbl, input_tbl) {
  ratio_tbl <- input_tbl %>%
    select(sample_name, tumor_peak_1, tumor_peak_2) %>%
    inner_join(summary_tbl %>% select(sample_name, is_primary_outlier), by = "sample_name") %>%
    filter(!is_primary_outlier) %>%
    inner_join(sample_order_tbl, by = "sample_name") %>%
    mutate(
      sample_name = factor(sample_name, levels = sample_levels),
      secondary_primary_ratio = tumor_peak_2 / tumor_peak_1
    ) %>%
    filter(is.finite(secondary_primary_ratio))

  ggplot(ratio_tbl, aes(x = sample_name, y = secondary_primary_ratio, fill = latest_match_date)) +
    geom_col(width = 0.72, color = "white", linewidth = 0.2) +
    geom_hline(yintercept = 2, linetype = "dashed", color = "grey35", linewidth = 0.4) +
    coord_flip() +
    scale_x_discrete(labels = sample_label_map) +
    labs(
      title = "Secondary-to-primary peak ratio by sample",
      subtitle = "Primary outliers are excluded. Dashed line marks a ratio of 2.",
      x = NULL,
      y = "Observed secondary / primary peak ratio",
      fill = "Sample date"
    ) +
    theme_bw(base_size = 9) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "bottom"
    )
}

build_histogram_plot_data <- function(summary_tbl, input_tbl, dna_vectors) {
  hist_tbl <- bind_rows(lapply(names(dna_vectors), function(sample_name) {
    tibble(
      sample_name = sample_name,
      dna_a = dna_vectors[[sample_name]]
    )
  })) %>%
    filter(is.finite(dna_a)) %>%
    inner_join(sample_order_tbl, by = "sample_name") %>%
    mutate(sample_name = factor(sample_name, levels = sample_levels))

  base_peak_tbl <- input_tbl %>%
    select(sample_name, cen_peak, tumor_peak_1, tumor_peak_2) %>%
    inner_join(summary_tbl %>% select(sample_name, is_primary_outlier), by = "sample_name") %>%
    inner_join(sample_order_tbl, by = "sample_name") %>%
    mutate(
      sample_name = factor(sample_name, levels = sample_levels),
      outlier_status = if_else(is_primary_outlier, "Primary outlier", "Not primary outlier")
    ) %>%
    select(sample_name, sample_label, outlier_status, cen_peak, tumor_peak_1, tumor_peak_2)

  peak_tbl <- bind_rows(
    base_peak_tbl %>%
      transmute(sample_name, sample_label, outlier_status, peak_type = "CEN peak", peak_location = cen_peak),
    base_peak_tbl %>%
      transmute(sample_name, sample_label, outlier_status, peak_type = "Tumor peak 1", peak_location = tumor_peak_1),
    base_peak_tbl %>%
      transmute(sample_name, sample_label, outlier_status, peak_type = "Tumor peak 2", peak_location = tumor_peak_2)
  ) %>%
    filter(is.finite(peak_location))

  list(hist_tbl = hist_tbl, peak_tbl = peak_tbl)
}

plot_sample_dna_histograms <- function(summary_tbl, input_tbl, dna_vectors) {
  plot_parts <- build_histogram_plot_data(summary_tbl, input_tbl, dna_vectors)
  hist_tbl <- plot_parts$hist_tbl
  peak_tbl <- plot_parts$peak_tbl
  binwidth <- max(1000, diff(range(hist_tbl$dna_a, na.rm = TRUE)) / 90)

  ggplot(hist_tbl, aes(x = dna_a)) +
    geom_histogram(binwidth = binwidth, fill = "grey82", color = "white", linewidth = 0.15) +
    geom_vline(
      aes(xintercept = peak_location, color = outlier_status, linetype = peak_type),
      data = peak_tbl,
      linewidth = 0.5,
      alpha = 0.9,
      show.legend = TRUE
    ) +
    facet_wrap(~ sample_name, scales = "free_y", ncol = 4, labeller = as_labeller(sample_label_map)) +
    scale_color_manual(values = outlier_palette) +
    scale_linetype_manual(
      values = c(
        "CEN peak" = "dotted",
        "Tumor peak 1" = "solid",
        "Tumor peak 2" = "longdash"
      )
    ) +
    scale_x_continuous(labels = label_comma()) +
    labs(
      title = "Per-sample raw DNA-A histograms with peak annotations",
      subtitle = "Peak markers are colored by each sample's primary-outlier status from the peak summary table.",
      x = "Raw DNA-A signal",
      y = "Cell count",
      color = NULL,
      linetype = NULL
    ) +
    theme_bw(base_size = 9) +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(size = 8),
      legend.position = "bottom"
    )
}
```

# Data

This report uses the current checkpoint outputs under `output/`, including the
baseline sample summary, ablation comparison, extension comparison, and the
karyotype lookup tables, plus the raw DNA-A vectors in
`processed_data/hypoxia-sum159/filtered_dna_area_vectors.rds`.

# Current Takeaway

The current checkpoint interpretation is narrower than the earlier
`phi = 1`-focused story.

The strongest ablation result is that the date effect is real and needed:
removing it causes a large degradation in lower-peak fit. By contrast, removing
the burden effect changes the fit much less, and fixing `rho = 1` is nearly
irrelevant at this checkpoint.

Among the extensions, a pure adjacent-state reinterpretation appears too
aggressive because it relabels some late 4N samples as 2N. The combined
`adjacent_8n_delta_dna` model is more plausible because it keeps the late 4N
samples as 4N while reassigning only a smaller subset of ambiguous O2 samples.

# Ablations

```{r ablation-table}
ablation_tbl %>%
  mutate(
    objective_lp = round(objective_lp, 2),
    rmse_cen = round(rmse_cen, 1),
    rmse_g1 = round(rmse_g1, 1),
    cor_counterfactual_vs_burden = round(cor_counterfactual_vs_burden, 3),
    mean_phi_date = round(mean_phi_date, 3),
    sd_phi_date = round(sd_phi_date, 3)
  ) %>%
  arrange(desc(include_date_effect), desc(include_x_ratio_effect), rho_fixed_one) %>%
  kable()
```

# Extensions

```{r extension-table}
extension_tbl %>%
  mutate(
    objective_lp = round(objective_lp, 2),
    rmse_cen = round(rmse_cen, 1),
    rmse_g1 = round(rmse_g1, 1),
    cor_counterfactual_vs_burden = round(cor_counterfactual_vs_burden, 3),
    mean_phi_date = round(mean_phi_date, 3),
    sd_phi_date = round(sd_phi_date, 3)
  ) %>%
  kable()
```

# State Reassignments

The table below isolates the samples whose lower-peak state assignment changes
relative to the baseline model.

```{r state-flips}
state_flip_tbl %>%
  mutate(
    observed_g1_peak = round(observed_g1_peak, 1),
    adjacent_8n_counterfactual = round(adjacent_8n_counterfactual, 1),
    adjacent_8n_delta_counterfactual = round(adjacent_8n_delta_counterfactual, 1)
  ) %>%
  kable()
```

This is the main reason the combined model is easier to defend than
`adjacent_8n` alone: the plain adjacent-state model also relabels the late
`4N_O1_A22` and `4N_O2_A22` samples as 2N, whereas the combined model leaves
those late 4N samples alone and only flips the two more ambiguous O2 samples
(`2N_O2_A18` and `2N_O2_A23`).

# Karyotype Link

These karyotype summaries do not prove the flow-state interpretation, but they
do make the O2 ambiguity biologically more interesting rather than purely a
fitting artifact.

```{r karyotype-summary}
karyotype_ploidy_tbl %>%
  mutate(
    mean_dna_ploidy_using_supplied_unidentified = round(mean_dna_ploidy_using_supplied_unidentified, 3),
    sd_dna_ploidy_using_supplied_unidentified = round(sd_dna_ploidy_using_supplied_unidentified, 3),
    mean_dna_ploidy_using_average_unidentified = round(mean_dna_ploidy_using_average_unidentified, 3),
    sd_dna_ploidy_using_average_unidentified = round(sd_dna_ploidy_using_average_unidentified, 3)
  ) %>%
  kable()
```

```{r karyotype-lookup}
karyotype_lookup_tbl %>%
  select(
    karyo_id,
    flow_sample_name,
    flow_condition,
    mapping_note,
    mean_dna_ploidy_using_supplied_unidentified
  ) %>%
  mutate(mean_dna_ploidy_using_supplied_unidentified = round(mean_dna_ploidy_using_supplied_unidentified, 3)) %>%
  kable()
```

# Peak Histograms

```{r raw-dna-histograms-with-peaks, fig.width=14, fig.height=12}
plot_sample_dna_histograms(summary_tbl, input_tbl, dna_vectors)
```

# Peak Ratio

```{r peak-ratio-by-sample, fig.width=10, fig.height=7}
plot_peak_ratio_by_sample(summary_tbl, input_tbl)
```

# Burden Proxy Plots

```{r observed-g1-vs-burden, fig.width=7, fig.height=4.5}
plot_burden_vs_peak(
  summary_tbl,
  y_var = "observed_g1_peak",
  y_label = "Observed G0/G1 peak",
  title = "Observed G0/G1 peak vs burden proxy"
)
```

```{r observed-cen-vs-burden, fig.width=7, fig.height=4.5}
plot_burden_vs_peak(
  summary_tbl,
  y_var = "cen_peak",
  y_label = "Observed CEN peak",
  title = "Observed CEN peak vs burden proxy"
)
```

```{r counterfactual-g1-vs-burden, fig.width=7, fig.height=4.5}
plot_burden_vs_peak(
  summary_tbl,
  y_var = "phi1_counterfactual_peak",
  y_label = "Counterfactual G0/G1 peak",
  title = "Counterfactual G0/G1 peak vs burden proxy"
)
```

The burden plots remain useful descriptively, but they are no longer the main
interpretive result of this checkpoint. The current comparison tables argue more
strongly for a persistent date effect and for a narrower adjacent-state-plus-
`delta_dna` explanation in a subset of ambiguous samples.
