---
title: "Flow Gating Workflow"
output:
  html_document:
    toc: true
    toc_float: true
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
---

```{r setup, include=FALSE}
input_path <- knitr::current_input()
workflow_dir <- if (length(input_path) == 0) getwd() else dirname(input_path)
project_root <- normalizePath(file.path(workflow_dir, ".."), winslash = "/", mustWork = TRUE)
if (!dir.exists(file.path(project_root, "data")) && dir.exists(file.path(getwd(), "data"))) {
  project_root <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
}
knitr::opts_knit$set(root.dir = project_root)
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

# Goal

This document runs the gating workflow independently for each dataset directory
under `data/`. Each dataset gets its own cache under `processed_data/` and its
own figure namespace under `figure/`, so additional datasets can be added
without colliding with existing outputs. The current inventory includes the
`Polyploidization_EthanolFixation` dataset alongside the anoxia and hypoxia
collections.

```{r load-js, results='asis'}
cat('
<script src="https://unpkg.com/panzoom@9.4.0/dist/panzoom.min.js"></script>
<style>
.plot-gallery {
  margin: 1rem 0 2rem 0;
}
.plot-gallery-controls {
  display: flex;
  align-items: center;
  gap: 0.75rem;
  margin-bottom: 0.75rem;
}
.plot-gallery-controls button {
  padding: 0.3rem 0.7rem;
}
.plot-gallery-status {
  font-size: 0.95rem;
  color: #444;
}
.plot-slide {
  max-width: 100%;
}
.plot-caption {
  margin-top: 0.5rem;
  font-size: 0.95rem;
  color: #444;
  text-align: center;
}
</style>
')
```

```{r libraries}
library(flowCore)
library(flowWorkspace)
library(PeacoQC)
library(MASS)
library(ggcyto)
library(scales)
library(knitr)
library(tibble)
library(dplyr)
library(patchwork)

source(file.path(project_root, "R", "metadata-utils.R"))
```

```{r config}
data_root <- "data"
processed_root <- "processed_data"
figure_root <- "figure"

reuse_existing_figures <- TRUE
force_rebuild_figures <- TRUE

default_qc_channels <- c(
  "FSC-A", "FSC-H", "FSC-W",
  "SSC-A", "SSC-H", "SSC-W",
  "450-Violet C-A", "450-Violet C-H"
)
default_dna_channels <- c("450-Violet C-A", "450-Violet C-H")
default_dna_cofactor <- 500
default_dna_fraction_threshold <- 20000
default_raw_dna_reference_threshold <- 12000

sanitize_dataset_id <- function(x) {
  x <- tolower(gsub("[^A-Za-z0-9]+", "-", x))
  x <- gsub("(^-+|-+$)", "", x)
  if (!nzchar(x)) {
    stop("Could not derive a dataset id from the supplied name.")
  }
  x
}

discover_dataset_dirs <- function(root_path) {
  dirs <- list.dirs(root_path, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[!startsWith(basename(dirs), ".")]
  dirs <- dirs[basename(dirs) != "cache"]
  dirs <- dirs[basename(dirs) != "__MACOSX"]
  dirs <- dirs[vapply(dirs, function(path) {
    length(discover_flow_files(path)) > 0
  }, logical(1))]
  sort(normalizePath(dirs, winslash = "/", mustWork = TRUE))
}

discover_flow_files <- function(dataset_dir) {
  files <- list.files(
    dataset_dir,
    pattern = "\\.fcs$",
    full.names = TRUE,
    recursive = TRUE,
    ignore.case = TRUE
  )
  files <- files[!grepl("(^|[\\\\/])__MACOSX([\\\\/]|$)", files)]
  files <- files[!startsWith(basename(files), "._")]
  sort(normalizePath(files, winslash = "/", mustWork = TRUE))
}

make_plot_paths <- function(figure_dir) {
  list(
    fsc_ssc = file.path(figure_dir, "fsc-ssc-plots-1.png"),
    fsc_time = file.path(figure_dir, "fsc-time-plots-1.png"),
    dna_scatter = file.path(figure_dir, "dna-scatter-plots-1.png"),
    dna_density = file.path(figure_dir, "dna-density-plots-1.png"),
    fluorescence_area_histograms = file.path(figure_dir, "fluorescence-area-histograms-1.png"),
    fluorescence_vs_dna = file.path(figure_dir, "fluorescence-vs-dna-1.png"),
    dna_fraction_summary = file.path(figure_dir, "dna-fraction-summary-1.png"),
    raw_dna_peak_locations = file.path(figure_dir, "raw-dna-peak-locations-1.png"),
    raw_dna_peak_location_covariation = file.path(figure_dir, "raw-dna-peak-location-covariation-1.png"),
    raw_dna_peak_width_covariation = file.path(figure_dir, "raw-dna-peak-width-covariation-1.png"),
    raw_dna_peak_vs_fraction_above_threshold = file.path(figure_dir, "raw-dna-peak-vs-fraction-above-threshold-1.png")
  )
}

discover_time_step_seconds <- function(flow_files, dataset_name) {
  kw <- keyword(read.FCS(
    flow_files[[1]],
    transformation = FALSE,
    truncate_max_range = FALSE
  ))
  time_step <- suppressWarnings(as.numeric(kw[["$TIMESTEP"]]))

  if (!is.finite(time_step) || time_step <= 0) {
    stop(sprintf(
      "Dataset '%s' is missing a valid $TIMESTEP keyword needed for FSC-A vs Time plotting.",
      dataset_name
    ))
  }

  time_step
}

dir.create(processed_root, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_root, recursive = TRUE, showWarnings = FALSE)
stopifnot(dir.exists(data_root))

dataset_dirs <- discover_dataset_dirs(data_root)
stopifnot(length(dataset_dirs) > 0)

dataset_tbl <- tibble(
  dataset_name = basename(dataset_dirs),
  dataset_dir = dataset_dirs,
  dataset_id = vapply(basename(dataset_dirs), sanitize_dataset_id, character(1)),
  enabled = TRUE
)

dataset_tbl$qc_channels <- rep(list(default_qc_channels), nrow(dataset_tbl))
dataset_tbl$dna_channels <- rep(list(default_dna_channels), nrow(dataset_tbl))
dataset_tbl$dna_cofactor <- rep(default_dna_cofactor, nrow(dataset_tbl))
dataset_tbl$dna_fraction_threshold <- rep(default_dna_fraction_threshold, nrow(dataset_tbl))
dataset_tbl$raw_dna_reference_threshold <- rep(default_raw_dna_reference_threshold, nrow(dataset_tbl))

ethanol_idx <- which(dataset_tbl$dataset_id == "polyploidization-ethanolfixation")
if (length(ethanol_idx) > 0) {
  dataset_tbl$qc_channels[ethanol_idx] <- list(c(
    "FSC-A", "FSC-H", "FSC-W",
    "SSC-A", "SSC-H", "SSC-W",
    "450/50 Violet B-A", "450/50 Violet B-H", "450/50 Violet B-W"
  ))
  dataset_tbl$dna_channels[ethanol_idx] <- list(c("450/50 Violet B-A", "450/50 Violet B-H"))
}

dataset_tbl <- dataset_tbl %>%
  filter(enabled) %>%
  mutate(
    cache_dir = file.path(processed_root, dataset_id),
    gs_cache_path = file.path(cache_dir, "gating_gs"),
    figure_dir = file.path(figure_root, dataset_id),
    sample_metadata_export_path = file.path(cache_dir, "sample_metadata.csv"),
    dna_vector_export_path = file.path(cache_dir, "filtered_dna_area_vectors.rds")
  )

dataset_tbl$flow_files <- lapply(dataset_tbl$dataset_dir, discover_flow_files)
dataset_tbl$n_files <- lengths(dataset_tbl$flow_files)
stopifnot(all(dataset_tbl$n_files > 0))
dataset_tbl$time_step_seconds <- mapply(
  discover_time_step_seconds,
  dataset_tbl$flow_files,
  dataset_tbl$dataset_name,
  SIMPLIFY = TRUE
)

dataset_tbl$plot_paths <- lapply(dataset_tbl$figure_dir, make_plot_paths)
dataset_tbl$has_cache <- dir.exists(dataset_tbl$gs_cache_path)
dataset_tbl$has_all_reusable_plots <- vapply(
  dataset_tbl$plot_paths,
  function(x) all(file.exists(unlist(x))),
  logical(1)
)

dataset_configs <- lapply(seq_len(nrow(dataset_tbl)), function(i) {
  list(
    dataset_name = dataset_tbl$dataset_name[[i]],
    dataset_dir = dataset_tbl$dataset_dir[[i]],
    dataset_id = dataset_tbl$dataset_id[[i]],
    qc_channels = dataset_tbl$qc_channels[[i]],
    dna_channels = dataset_tbl$dna_channels[[i]],
    dna_cofactor = dataset_tbl$dna_cofactor[[i]],
    dna_fraction_threshold = dataset_tbl$dna_fraction_threshold[[i]],
    raw_dna_reference_threshold = dataset_tbl$raw_dna_reference_threshold[[i]],
    time_step_seconds = dataset_tbl$time_step_seconds[[i]],
    cache_dir = dataset_tbl$cache_dir[[i]],
    gs_cache_path = dataset_tbl$gs_cache_path[[i]],
    figure_dir = dataset_tbl$figure_dir[[i]],
    sample_metadata_export_path = dataset_tbl$sample_metadata_export_path[[i]],
    dna_vector_export_path = dataset_tbl$dna_vector_export_path[[i]],
    flow_files = dataset_tbl$flow_files[[i]],
    plot_paths = dataset_tbl$plot_paths[[i]],
    has_cache = dataset_tbl$has_cache[[i]],
    has_all_reusable_plots = dataset_tbl$has_all_reusable_plots[[i]]
  )
})
names(dataset_configs) <- vapply(dataset_configs, `[[`, character(1), "dataset_id")
```

# Dataset Inventory

```{r dataset-inventory}
dataset_inventory <- dataset_tbl %>%
  transmute(
    dataset_name,
    dataset_id,
    n_files,
    has_cache,
    has_all_reusable_plots
  )

if (nrow(dataset_inventory) <= 12 && ncol(dataset_inventory) <= 6) {
  kable(dataset_inventory)
}
```

# Hypoxia Passaging Days

```{r hypoxia-passaging-days}
passaging_path <- file.path(data_root, "passaging.csv")
passaging_tbl <- read.csv(passaging_path, stringsAsFactors = FALSE, check.names = FALSE)

hypoxia_cfg <- dataset_tbl %>%
  filter(dataset_id == "hypoxia-sum159") %>%
  slice(1)

hypoxia_passaging_day_tbl <- if (nrow(hypoxia_cfg) == 1) {
  build_sum159_hypoxia_match_table(
    sample_names = hypoxia_cfg$flow_files[[1]],
    passaging_tbl = passaging_tbl
  ) %>%
    transmute(
      sample_name,
      latest_matching_id,
      latest_match_date,
      relative_day
    ) %>%
    distinct(sample_name, .keep_all = TRUE)
} else {
  tibble(
    sample_name = character(),
    latest_matching_id = character(),
    latest_match_date = as.Date(character()),
    relative_day = integer()
  )
}

kable(hypoxia_passaging_day_tbl)
```

# Helper Functions

```{r helpers}
first_flowFrame <- function(x) {
  if (inherits(x, "flowFrame")) return(x)
  if (is.list(x)) {
    hits <- which(vapply(x, function(z) inherits(z, "flowFrame"), logical(1)))
    if (length(hits) > 0) return(x[[hits[1]]])
  }
  stop("Could not find a flowFrame in RemoveMargins output.")
}

validate_dna_channels <- function(dna_channels, dataset_name) {
  if (length(dna_channels) != 2) {
    stop(sprintf(
      "Dataset '%s' must define exactly two DNA channels; got %d.",
      dataset_name,
      length(dna_channels)
    ))
  }
}

validate_flowset_channels <- function(fs, required_channels, dataset_name) {
  missing_by_sample <- lapply(sampleNames(fs), function(sn) {
    setdiff(required_channels, colnames(fs[[sn]]))
  })
  names(missing_by_sample) <- sampleNames(fs)

  bad_samples <- names(missing_by_sample)[lengths(missing_by_sample) > 0]
  if (length(bad_samples) > 0) {
    detail <- paste(
      sprintf(
        "%s [%s]",
        bad_samples,
        vapply(
          missing_by_sample[bad_samples],
          function(x) paste(x, collapse = ", "),
          character(1)
        )
      ),
      collapse = "; "
    )
    stop(sprintf(
      "Dataset '%s' is missing required channels in one or more samples: %s",
      dataset_name,
      detail
    ))
  }
}

validate_cached_gs <- function(gs, flow_files, dataset_name) {
  expected_samples <- sort(basename(flow_files))
  cached_samples <- sort(basename(sampleNames(gs)))

  if (!identical(expected_samples, cached_samples)) {
    stop(sprintf(
      "Cached GatingSet for dataset '%s' does not match the current files on disk. Remove '%s' and rerender.",
      dataset_name,
      file.path(processed_root, sanitize_dataset_id(dataset_name), "gating_gs")
    ))
  }
}

dna_singlet_vec <- function(ff, a = "450-Violet C-A", h = "450-Violet C-H", nmads = 4) {
  xy <- exprs(ff)[, c(a, h), drop = FALSE]
  ok <- is.finite(xy[, 1]) & is.finite(xy[, 2]) & xy[, 1] > 0 & xy[, 2] > 0

  if (sum(ok) < 100) {
    return(rep(FALSE, nrow(xy)))
  }

  fit <- MASS::rlm(xy[ok, 2] ~ xy[ok, 1])
  pred <- coef(fit)[1] + coef(fit)[2] * xy[ok, 1]
  res <- xy[ok, 2] - pred

  spread <- mad(res, na.rm = TRUE)
  if (!is.finite(spread) || spread == 0) {
    spread <- sd(res, na.rm = TRUE)
  }
  if (!is.finite(spread) || spread == 0) {
    spread <- 1e-8
  }

  keep_ok <- abs(res - median(res, na.rm = TRUE)) <= nmads * spread
  keep <- rep(FALSE, nrow(xy))
  keep[which(ok)[keep_ok]] <- TRUE
  keep
}

process_masks <- function(ff_raw, qc_channels, dna_channels, IT_limit = 0.65, dna_cofactor = 500, nmads = 4) {
  idx <- match(qc_channels, colnames(ff_raw))
  idx <- idx[!is.na(idx)]

  mr <- RemoveMargins(ff_raw, channels = idx, output = "full")
  ff_margin_raw <- first_flowFrame(mr)

  orig_id_margin <- as.integer(exprs(ff_margin_raw)[, "Original_ID"])
  margin_keep_global <- rep(FALSE, nrow(ff_raw))
  margin_keep_global[orig_id_margin] <- TRUE

  use_dna <- intersect(dna_channels, colnames(ff_margin_raw))
  if (length(use_dna) > 0) {
    tf <- transformList(use_dna, arcsinhTransform(a = 0, b = 1 / dna_cofactor, c = 0))
    ff_margin_proc <- transform(ff_margin_raw, tf)
  } else {
    ff_margin_proc <- ff_margin_raw
  }

  pq <- PeacoQC(
    ff = ff_margin_proc,
    channels = idx,
    determine_good_cells = "all",
    save_fcs = FALSE,
    plot = FALSE,
    report = FALSE,
    IT_limit = IT_limit
  )
  peacoqc_keep_local <- pq$GoodCells
  ff_peacoqc <- pq$FinalFF

  dna_keep_local <- dna_singlet_vec(
    ff_peacoqc,
    a = dna_channels[1],
    h = dna_channels[2],
    nmads = nmads
  )

  list(
    margin_keep = margin_keep_global,
    peacoqc_keep = peacoqc_keep_local,
    dna_keep = dna_keep_local
  )
}

build_gs <- function(flow_files, qc_channels, dna_channels, dataset_name, dna_cofactor = 500) {
  validate_dna_channels(dna_channels, dataset_name)

  fs <- read.flowSet(
    files = flow_files,
    transformation = FALSE,
    truncate_max_range = FALSE
  )
  validate_flowset_channels(fs, unique(c(qc_channels, dna_channels)), dataset_name)

  mask_list <- lapply(sampleNames(fs), function(sn) {
    process_masks(fs[[sn]], qc_channels = qc_channels, dna_channels = dna_channels, dna_cofactor = dna_cofactor)
  })
  names(mask_list) <- sampleNames(fs)

  margin_list <- setNames(lapply(mask_list, `[[`, "margin_keep"), sampleNames(fs))
  peacoqc_list <- setNames(lapply(mask_list, `[[`, "peacoqc_keep"), sampleNames(fs))
  dna_list <- setNames(lapply(mask_list, `[[`, "dna_keep"), sampleNames(fs))

  gs <- GatingSet(fs)

  dna_ch_exist <- intersect(dna_channels, colnames(gs))
  if (length(dna_ch_exist) > 0) {
    custom_arcsinh <- trans_new(
      name = "custom_arcsinh",
      transform = function(x) asinh(x / dna_cofactor),
      inverse = function(x) sinh(x) * dna_cofactor,
      breaks = extended_breaks(),
      format = scientific_format()
    )

    trans_list <- transformerList(
      dna_ch_exist,
      lapply(dna_ch_exist, function(x) custom_arcsinh)
    )
    gs <- transform(gs, trans_list)
  }

  gs_pop_add(gs, margin_list, name = "Margin_Clean", parent = "root")
  recompute(gs)

  gs_pop_add(gs, peacoqc_list, name = "PeacoQC_Clean", parent = "Margin_Clean")
  recompute(gs)

  gs_pop_add(gs, dna_list, name = "Singlets", parent = "PeacoQC_Clean")
  recompute(gs)

  gs
}

build_or_load_dataset <- function(cfg) {
  dir.create(cfg$cache_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(cfg$figure_dir, recursive = TRUE, showWarnings = FALSE)

  gs <- if (dir.exists(cfg$gs_cache_path)) {
    cached_gs <- load_gs(cfg$gs_cache_path)
    validate_cached_gs(cached_gs, cfg$flow_files, cfg$dataset_name)
    cached_gs
  } else {
    built_gs <- build_gs(
      flow_files = cfg$flow_files,
      qc_channels = cfg$qc_channels,
      dna_channels = cfg$dna_channels,
      dataset_name = cfg$dataset_name,
      dna_cofactor = cfg$dna_cofactor
    )
    save_gs(built_gs, path = cfg$gs_cache_path, backend_opt = "copy")
    built_gs
  }

  sample_metadata <- parse_sample_metadata(cfg$flow_files, dataset_id = cfg$dataset_id)
  if (identical(cfg$dataset_id, "hypoxia-sum159") && nrow(hypoxia_passaging_day_tbl) > 0) {
    sample_metadata <- sample_metadata %>%
      left_join(hypoxia_passaging_day_tbl, by = "sample_name")
  }
  write.csv(sample_metadata, cfg$sample_metadata_export_path, row.names = FALSE)

  input_tbl <- tibble(
    dataset_id = cfg$dataset_id,
    dataset_name = cfg$dataset_name,
    sample_name = basename(cfg$flow_files),
    source_path = cfg$flow_files
  ) %>%
    left_join(sample_metadata, by = "sample_name")

  gate_counts <- as_tibble(gs_pop_get_count_fast(gs)) %>%
    mutate(
      dataset_id = cfg$dataset_id,
      dataset_name = cfg$dataset_name,
      .before = 1
    ) %>%
    add_sample_metadata(sample_metadata)

  flow_file_map <- setNames(cfg$flow_files, tools::file_path_sans_ext(basename(cfg$flow_files)))
  sample_keys <- tools::file_path_sans_ext(basename(sampleNames(gs)))
  names(sample_keys) <- sampleNames(gs)

  filtered_dna_area_raw <- setNames(
    lapply(sampleNames(gs), function(sn) {
      keep_idx <- gh_pop_get_indices(gs[[sn]], "Singlets")
      raw_ff <- read.FCS(
        flow_file_map[[sample_keys[[sn]]]],
        transformation = FALSE,
        truncate_max_range = FALSE
      )
      vals <- exprs(raw_ff)[keep_idx, cfg$dna_channels[1]]
      as.numeric(vals[is.finite(vals)])
    }),
    sample_keys
  )

  filtered_dna_area_transformed <- setNames(
    lapply(sampleNames(gs), function(sn) {
      root_ff <- gh_pop_get_data(gs[[sn]], "root")
      keep_idx <- gh_pop_get_indices(gs[[sn]], "Singlets")
      vals <- exprs(root_ff)[keep_idx, cfg$dna_channels[1]]
      as.numeric(vals[is.finite(vals)])
    }),
    sample_keys
  )

  filtered_dna_area <- list(
    raw = filtered_dna_area_raw,
    transformed = filtered_dna_area_transformed
  )
  saveRDS(filtered_dna_area, cfg$dna_vector_export_path)

  list(
    config = cfg,
    gs = gs,
    input_tbl = input_tbl,
    sample_metadata = sample_metadata,
    gate_counts = gate_counts,
    filtered_dna_area = filtered_dna_area
  )
}

parse_ethanol_sample_metadata <- function(sample_id) {
  n <- length(sample_id)

  metadata_tbl <- tibble(
    sample_id = sample_id,
    cell_line = rep(NA_character_, n),
    condition = rep(NA_character_, n),
    exposure_condition = rep(NA_character_, n),
    passage_label = rep(NA_character_, n),
    passage_number = rep(NA_integer_, n),
    acquisition_id = rep(NA_integer_, n),
    sample_role = rep(NA_character_, n)
  )

  metadata_tbl$acquisition_id <- suppressWarnings(as.integer(sub("^.*_([0-9]+)$", "\\1", sample_id)))

  jurkat_idx <- grepl("^Jurkat_[0-9]+$", sample_id)
  if (any(jurkat_idx)) {
    metadata_tbl$cell_line[jurkat_idx] <- "Jurkat"
    metadata_tbl$sample_role[jurkat_idx] <- "reference"
  }

  parental_idx <- grepl("^(.*?)(?:_|-)parental_[0-9]+$", sample_id, ignore.case = TRUE)
  if (any(parental_idx)) {
    metadata_tbl$cell_line[parental_idx] <- sub(
      "^(.*?)(?:_|-)parental_[0-9]+$",
      "\\1",
      sample_id[parental_idx],
      ignore.case = TRUE
    )
    metadata_tbl$condition[parental_idx] <- "parental"
    metadata_tbl$exposure_condition[parental_idx] <- "parental"
    metadata_tbl$sample_role[parental_idx] <- "control"
  }

  exposed_idx <- grepl("^[^_]+_[^_]+_A[0-9]+_[0-9]+$", sample_id)
  if (any(exposed_idx)) {
    exposed_parts <- strsplit(sample_id[exposed_idx], "_", fixed = TRUE)
    metadata_tbl$cell_line[exposed_idx] <- vapply(exposed_parts, `[[`, character(1), 1)
    metadata_tbl$exposure_condition[exposed_idx] <- vapply(exposed_parts, `[[`, character(1), 2)
    metadata_tbl$passage_label[exposed_idx] <- vapply(exposed_parts, `[[`, character(1), 3)
    metadata_tbl$passage_number[exposed_idx] <- suppressWarnings(as.integer(sub("^A", "", metadata_tbl$passage_label[exposed_idx])))
    metadata_tbl$condition[exposed_idx] <- paste(
      metadata_tbl$exposure_condition[exposed_idx],
      metadata_tbl$passage_label[exposed_idx],
      sep = "_"
    )
    metadata_tbl$sample_role[exposed_idx] <- "experiment"
  }

  metadata_tbl
}

parse_sample_metadata <- function(sample_name, dataset_id = NULL) {
  sample_id <- tools::file_path_sans_ext(basename(sample_name))
  cell_line <- rep(NA_character_, length(sample_id))
  condition <- rep(NA_character_, length(sample_id))
  exposure_condition <- rep(NA_character_, length(sample_id))
  passage_label <- rep(NA_character_, length(sample_id))
  passage_number <- rep(NA_integer_, length(sample_id))
  acquisition_id <- rep(NA_integer_, length(sample_id))
  sample_role <- rep(NA_character_, length(sample_id))

  cell_line[sample_id == "Sample_CEN"] <- "CEN"
  sample_role[sample_id == "Sample_CEN"] <- "reference"

  sum159_idx <- grepl("^Sample_SUM159_", sample_id)
  cell_line[sum159_idx] <- "SUM159"
  condition[sum159_idx] <- sub("^Sample_SUM159_([0-9]+_)?", "", sample_id[sum159_idx])
  sample_role[sum159_idx] <- "experiment"

  if (identical(dataset_id, "polyploidization-ethanolfixation")) {
    ethanol_metadata <- parse_ethanol_sample_metadata(sample_id)
    fill_idx <- !is.na(ethanol_metadata$cell_line) |
      !is.na(ethanol_metadata$condition) |
      !is.na(ethanol_metadata$sample_role)

    cell_line[fill_idx] <- ethanol_metadata$cell_line[fill_idx]
    condition[fill_idx] <- ethanol_metadata$condition[fill_idx]
    exposure_condition[fill_idx] <- ethanol_metadata$exposure_condition[fill_idx]
    passage_label[fill_idx] <- ethanol_metadata$passage_label[fill_idx]
    passage_number[fill_idx] <- ethanol_metadata$passage_number[fill_idx]
    acquisition_id[fill_idx] <- ethanol_metadata$acquisition_id[fill_idx]
    sample_role[fill_idx] <- ethanol_metadata$sample_role[fill_idx]
  }

  tibble(
    sample_name = basename(sample_name),
    sample_id = sample_id,
    cell_line = cell_line,
    condition = condition,
    exposure_condition = exposure_condition,
    passage_label = passage_label,
    passage_number = passage_number,
    acquisition_id = acquisition_id,
    sample_role = sample_role
  )
}

add_sample_metadata <- function(tbl, metadata_tbl) {
  sample_col <- intersect(c("sample_name", "name", "Name", "sample", "Sample"), names(tbl))[1]
  if (is.na(sample_col)) {
    return(tbl)
  }

  tbl %>%
    mutate(.sample_id = tools::file_path_sans_ext(basename(.data[[sample_col]]))) %>%
    left_join(
      select(
        metadata_tbl,
        sample_id,
        cell_line,
        condition,
        exposure_condition,
        passage_label,
        passage_number,
        acquisition_id,
        sample_role
      ),
      by = c(".sample_id" = "sample_id")
    ) %>%
    select(-.sample_id)
}

should_print_table <- function(tbl, max_rows = 12, max_cols = 6) {
  nrow(tbl) <= max_rows && ncol(tbl) <= max_cols
}

dataset_input_summary <- function(input_tbl) {
  tibble(
    dataset_id = dplyr::first(input_tbl$dataset_id),
    dataset_name = dplyr::first(input_tbl$dataset_name),
    n_files = nrow(input_tbl),
    cell_lines = paste(sort(unique(na.omit(input_tbl$cell_line))), collapse = ", "),
    conditions = paste(sort(unique(na.omit(input_tbl$condition))), collapse = ", "),
    first_sample = dplyr::first(input_tbl$sample_name),
    last_sample = dplyr::last(input_tbl$sample_name)
  )
}

dataset_metadata_summary <- function(input_tbl) {
  input_tbl %>%
    count(cell_line, condition, name = "n_files") %>%
    arrange(cell_line, condition)
}

dataset_gate_summary <- function(gate_counts) {
  value_col <- intersect(c("count", "Count"), names(gate_counts))[1]
  pop_col <- intersect(c("pop", "Population", "population", "name"), names(gate_counts))[1]

  if (is.na(value_col) || is.na(pop_col)) {
    return(tibble(note = "Gate count schema not recognized for compact summary."))
  }

  gate_counts %>%
    group_by(.data[[pop_col]]) %>%
    summarise(
      n_samples = dplyr::n(),
      min_count = min(.data[[value_col]], na.rm = TRUE),
      median_count = median(.data[[value_col]], na.rm = TRUE),
      max_count = max(.data[[value_col]], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    rename(population = 1)
}

dataset_export_summary <- function(filtered_dna_area, dna_vector_export_path, metadata_export_path) {
  vector_lengths <- lengths(filtered_dna_area$raw)

  tibble(
    dna_vector_export_path = dna_vector_export_path,
    metadata_export_path = metadata_export_path,
    n_samples = length(filtered_dna_area$raw),
    min_vector_length = min(vector_lengths),
    median_vector_length = median(vector_lengths),
    max_vector_length = max(vector_lengths),
    includes_transformed = "transformed" %in% names(filtered_dna_area)
  )
}

ensure_plot_file <- function(plot_path, plot_builder, width, height, dpi = 300) {
  should_rebuild <- force_rebuild_figures || !reuse_existing_figures || !file.exists(plot_path)

  if (should_rebuild) {
    plot_obj <- plot_builder()
    ggplot2::ggsave(
      filename = plot_path,
      plot = plot_obj,
      width = width,
      height = height,
      dpi = dpi,
      units = "in",
      limitsize = FALSE
    )
  }
}

render_or_include_plot <- function(plot_path, plot_builder, width, height, element_id, dpi = 300) {
  ensure_plot_file(plot_path, plot_builder, width, height, dpi)

  plot_src <- file.path("..", plot_path)
  plot_src <- gsub("\\\\", "/", plot_src)
  sprintf(
    '<img src="%s" id="%s" class="zoomable-plot" style="width:100%%;" />',
    plot_src,
    element_id
  )
}

render_plot_gallery <- function(plot_paths, plot_builders, width, height, element_id, captions = NULL, dpi = 300) {
  stopifnot(length(plot_paths) == length(plot_builders))
  if (is.null(captions)) {
    captions <- rep("", length(plot_paths))
  }
  stopifnot(length(captions) == length(plot_paths))

  for (i in seq_along(plot_paths)) {
    ensure_plot_file(plot_paths[[i]], plot_builders[[i]], width, height, dpi)
  }

  slides_html <- vapply(seq_along(plot_paths), function(i) {
    plot_src <- file.path("..", plot_paths[[i]])
    plot_src <- gsub("\\\\", "/", plot_src)
    slide_display <- if (i == 1) "block" else "none"
    sprintf(
      paste0(
        '<div class="plot-slide" data-index="%d" style="display:%s;">',
        '<img src="%s" class="zoomable-plot gallery-img" style="width:100%%;" />',
        '<div class="plot-caption">%s</div>',
        '</div>'
      ),
      i - 1,
      slide_display,
      plot_src,
      captions[[i]]
    )
  }, character(1))

  sprintf(
    paste0(
      '<div class="plot-gallery" id="%s">',
      '<div class="plot-gallery-controls">',
      "<button type=\"button\" onclick=\"plotGalleryPrev('%s')\">Prev</button>",
      "<button type=\"button\" onclick=\"plotGalleryNext('%s')\">Next</button>",
      '<span class="plot-gallery-status"></span>',
      '</div>',
      '%s',
      '</div>'
    ),
    element_id,
    element_id,
    element_id,
    paste(slides_html, collapse = "")
  )
}

build_fsc_ssc_plot <- function(gs) {
  p_fsc_raw <- ggcyto(gs, subset = "root", aes(x = `FSC-A`, y = `SSC-A`)) +
    geom_hex(bins = 60) +
    labs(title = "FSC vs SSC (Raw)") +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  p_fsc_filtered <- ggcyto(gs, subset = "Singlets", aes(x = `FSC-A`, y = `SSC-A`)) +
    geom_hex(bins = 60) +
    labs(title = "FSC vs SSC (Filtered)") +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  as.ggplot(p_fsc_raw) + as.ggplot(p_fsc_filtered)
}

build_fsc_time_plot <- function(gs, time_step_seconds, dna_channel, smooth_window = 5) {
  p_fsc_time_raw <- ggcyto(gs, subset = "root", aes(x = Time, y = `FSC-A`)) +
    geom_hex(bins = 60) +
    facet_wrap(~ name, scales = "free_x") +
    labs(title = "FSC-A vs Time", x = "Time (s)", y = "FSC-A") +
    scale_x_continuous(labels = function(x) scales::label_number()(x * time_step_seconds)) +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  root_data <- gs_pop_get_data(gs, "root")
  time_summary_df <- bind_rows(lapply(sampleNames(root_data), function(sn) {
    ff <- root_data[[sn]]
    expr <- exprs(ff)
    time_values <- expr[, "Time"]
    dna_values <- expr[, dna_channel]
    time_seconds <- time_values * time_step_seconds
    keep <- is.finite(time_seconds) & is.finite(dna_values)
    time_seconds <- time_seconds[keep]
    dna_values <- dna_values[keep]

    if (length(time_seconds) == 0) {
      return(tibble(
        name = sn,
        time_bin_start = numeric(0),
        event_rate_hz = numeric(0),
        dna_mean = numeric(0),
        dna_sd = numeric(0)
      ))
    }

    tibble(
      name = sn,
      time_bin_start = floor(time_seconds),
      dna_value = dna_values
    ) %>%
      group_by(name, time_bin_start) %>%
      summarise(
        event_rate_hz = dplyr::n(),
        dna_mean = mean(dna_value, na.rm = TRUE),
        dna_sd = sd(dna_value, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      arrange(time_bin_start) %>%
      mutate(
        dna_mean_smooth = as.numeric(stats::filter(
          dna_mean,
          rep(1 / smooth_window, smooth_window),
          sides = 2
        )),
        dna_sd_smooth = as.numeric(stats::filter(
          dna_sd,
          rep(1 / smooth_window, smooth_window),
          sides = 2
        ))
      ) %>%
      mutate(
        dna_mean_smooth = ifelse(is.na(dna_mean_smooth), dna_mean, dna_mean_smooth),
        dna_sd_smooth = ifelse(is.na(dna_sd_smooth), dna_sd, dna_sd_smooth)
      )
  }))

  rate_max <- max(time_summary_df$event_rate_hz, na.rm = TRUE)
  dna_upper_max <- max(time_summary_df$dna_mean_smooth + time_summary_df$dna_sd_smooth, na.rm = TRUE)
  dna_scale_factor <- if (is.finite(rate_max) && is.finite(dna_upper_max) && dna_upper_max > 0) {
    rate_max / dna_upper_max
  } else {
    1
  }

  p_event_rate <- ggplot(time_summary_df, aes(x = time_bin_start, y = event_rate_hz)) +
    geom_line(linewidth = 0.3) +
    geom_ribbon(
      aes(
        x = time_bin_start,
        ymin = pmax((dna_mean_smooth - dna_sd_smooth) * dna_scale_factor, 0),
        ymax = pmax((dna_mean_smooth + dna_sd_smooth) * dna_scale_factor, 0)
      ),
      inherit.aes = FALSE,
      fill = "#4C78A8",
      alpha = 0.15
    ) +
    geom_line(
      aes(y = dna_mean_smooth * dna_scale_factor),
      color = "#4C78A8",
      linewidth = 0.4
    ) +
    facet_wrap(~ name, scales = "free") +
    labs(
      title = sprintf("Event Rate and %s vs Time", dna_channel),
      x = "Time (s)",
      y = "Events/s"
    ) +
    scale_y_continuous(
      sec.axis = sec_axis(~ . / dna_scale_factor, name = sprintf("%s Mean +/- SD", dna_channel))
    ) +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  as.ggplot(p_fsc_time_raw) + p_event_rate
}

build_dna_scatter_plot <- function(gs, dna_channels) {
  p_dna_raw <- ggcyto(gs, subset = "root", aes(x = .data[[dna_channels[1]]], y = .data[[dna_channels[2]]])) +
    geom_hex(bins = 60) +
    labs(title = "DNA Scatter (Raw)") +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  p_dna_filtered <- ggcyto(gs, subset = "Singlets", aes(x = .data[[dna_channels[1]]], y = .data[[dna_channels[2]]])) +
    geom_hex(bins = 60) +
    labs(title = "DNA Scatter (Filtered Singlets)") +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  as.ggplot(p_dna_raw) + as.ggplot(p_dna_filtered)
}

build_dna_density_plot <- function(gs, dna_channels, dna_threshold, dna_cofactor) {
  threshold_transformed <- asinh(dna_threshold / dna_cofactor)

  p_density_raw <- ggcyto(gs, subset = "root", aes(x = .data[[dna_channels[1]]])) +
    geom_density(fill = "gray50", alpha = 0.5, linewidth = 0.5) +
    geom_vline(xintercept = threshold_transformed, color = "red3", linetype = "dashed", linewidth = 0.7) +
    labs(title = "DNA Density (Raw)", y = "Density") +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  p_density_filtered <- ggcyto(gs, subset = "Singlets", aes(x = .data[[dna_channels[1]]])) +
    geom_density(fill = "gray50", alpha = 0.5, linewidth = 0.5) +
    geom_vline(xintercept = threshold_transformed, color = "red3", linetype = "dashed", linewidth = 0.7) +
    labs(title = "DNA Density (Filtered Singlets)", y = "Density") +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  as.ggplot(p_density_raw) + as.ggplot(p_density_filtered)
}

discover_other_fluorescence_area_channels <- function(gs, dna_channels) {
  root_ff <- gh_pop_get_data(gs[[sampleNames(gs)[1]]], "root")
  area_channels <- grep("-A$", colnames(root_ff), value = TRUE)
  setdiff(area_channels, c("FSC-A", "SSC-A", dna_channels[1]))
}

empty_plot <- function(title, subtitle) {
  ggplot() +
    annotate("text", x = 0, y = 0, label = subtitle, size = 4.5) +
    labs(title = title) +
    theme_void()
}

extract_population_channel_df <- function(
  gs,
  channels,
  population = c("root", "Singlets"),
  max_points_per_sample_channel = 50000,
  seed = 1
) {
  population <- match.arg(population)
  if (length(channels) == 0) {
    return(tibble(
      sample_label = character(),
      channel = character(),
      value = numeric()
    ))
  }

  set.seed(seed)
  bind_rows(lapply(sampleNames(gs), function(sn) {
    ff <- gh_pop_get_data(gs[[sn]], population)
    expr <- exprs(ff)
    sample_label <- tools::file_path_sans_ext(basename(sn))

    bind_rows(lapply(channels, function(channel) {
      if (!channel %in% colnames(expr)) {
        return(NULL)
      }

      values <- as.numeric(expr[, channel])
      values <- values[is.finite(values)]
      if (length(values) == 0) {
        return(NULL)
      }

      if (length(values) > max_points_per_sample_channel) {
        values <- sample(values, max_points_per_sample_channel)
      }

      tibble(
        sample_label = sample_label,
        channel = channel,
        value = values
      )
    }))
  }))
}

extract_population_channel_pairs_df <- function(
  gs,
  x_channel,
  y_channels,
  population = c("root", "Singlets"),
  max_points_per_sample_channel = 30000,
  seed = 1
) {
  population <- match.arg(population)
  if (length(y_channels) == 0) {
    return(tibble(
      sample_label = character(),
      channel = character(),
      x_value = numeric(),
      y_value = numeric()
    ))
  }

  set.seed(seed)
  bind_rows(lapply(sampleNames(gs), function(sn) {
    ff <- gh_pop_get_data(gs[[sn]], population)
    expr <- exprs(ff)
    sample_label <- tools::file_path_sans_ext(basename(sn))

    if (!x_channel %in% colnames(expr)) {
      return(NULL)
    }

    bind_rows(lapply(y_channels, function(channel) {
      if (!channel %in% colnames(expr)) {
        return(NULL)
      }

      x_values <- as.numeric(expr[, x_channel])
      y_values <- as.numeric(expr[, channel])
      keep <- is.finite(x_values) & is.finite(y_values)
      if (!any(keep)) {
        return(NULL)
      }

      x_values <- x_values[keep]
      y_values <- y_values[keep]

      if (length(x_values) > max_points_per_sample_channel) {
        idx <- sample.int(length(x_values), max_points_per_sample_channel)
        x_values <- x_values[idx]
        y_values <- y_values[idx]
      }

      tibble(
        sample_label = sample_label,
        channel = channel,
        x_value = x_values,
        y_value = y_values
      )
    }))
  }))
}

build_fluorescence_area_histogram_plot <- function(gs, fluorescence_channels, bins = 80) {
  if (length(fluorescence_channels) == 0) {
    return(empty_plot(
      title = "Other Fluorescence Area Histograms",
      subtitle = "No non-DNA fluorescence area channels were found."
    ))
  }

  raw_df <- extract_population_channel_df(
    gs,
    channels = fluorescence_channels,
    population = "root",
    max_points_per_sample_channel = 50000
  )
  filtered_df <- extract_population_channel_df(
    gs,
    channels = fluorescence_channels,
    population = "Singlets",
    max_points_per_sample_channel = 50000
  )

  if (nrow(raw_df) == 0 && nrow(filtered_df) == 0) {
    return(empty_plot(
      title = "Other Fluorescence Area Histograms",
      subtitle = "No finite fluorescence area values were available."
    ))
  }

  plot_hist <- function(plot_df, title) {
    if (nrow(plot_df) == 0) {
      return(empty_plot(title, "No values were available for this population."))
    }

    ggplot(plot_df, aes(x = value)) +
      geom_histogram(bins = bins, fill = "gray55", color = "white", linewidth = 0.15) +
      facet_grid(channel ~ sample_label, scales = "free_x") +
      labs(title = title, x = "Area Signal", y = "Event Count") +
      theme_bw() +
      theme(
        strip.text.x = element_text(size = 8),
        strip.text.y = element_text(size = 8),
        panel.grid.minor = element_blank()
      )
  }

  plot_hist(raw_df, "Other Fluorescence Area Histograms (Raw)") /
    plot_hist(filtered_df, "Other Fluorescence Area Histograms (Filtered Singlets)")
}

build_fluorescence_vs_dna_plot <- function(gs, dna_channel, fluorescence_channels, bins = 50) {
  if (length(fluorescence_channels) == 0) {
    return(empty_plot(
      title = "Other Fluorescence vs DNA-A",
      subtitle = "No non-DNA fluorescence area channels were found."
    ))
  }

  raw_df <- extract_population_channel_pairs_df(
    gs,
    x_channel = dna_channel,
    y_channels = fluorescence_channels,
    population = "root",
    max_points_per_sample_channel = 30000
  )
  filtered_df <- extract_population_channel_pairs_df(
    gs,
    x_channel = dna_channel,
    y_channels = fluorescence_channels,
    population = "Singlets",
    max_points_per_sample_channel = 30000
  )

  if (nrow(raw_df) == 0 && nrow(filtered_df) == 0) {
    return(empty_plot(
      title = "Other Fluorescence vs DNA-A",
      subtitle = "No finite paired fluorescence and DNA-A values were available."
    ))
  }

  plot_hex <- function(plot_df, title) {
    if (nrow(plot_df) == 0) {
      return(empty_plot(title, "No paired values were available for this population."))
    }

    ggplot(plot_df, aes(x = x_value, y = y_value)) +
      geom_hex(bins = bins) +
      facet_grid(channel ~ sample_label, scales = "free") +
      scale_fill_viridis_c(trans = "log10", option = "C", na.value = "white") +
      labs(
        title = title,
        x = dna_channel,
        y = "Fluorescence Area Signal",
        fill = "Bin Count"
      ) +
      theme_bw() +
      theme(
        strip.text.x = element_text(size = 8),
        strip.text.y = element_text(size = 8),
        panel.grid.minor = element_blank()
      )
  }

  plot_hex(raw_df, sprintf("Other Fluorescence vs %s (Raw)", dna_channel)) /
    plot_hex(filtered_df, sprintf("Other Fluorescence vs %s (Filtered Singlets)", dna_channel))
}

build_dna_fraction_summary_plot <- function(gs, filtered_dna_area, sample_metadata, dna_threshold) {
  total_event_df <- tibble(
    sample_name = basename(sampleNames(gs)),
    total_events = vapply(sampleNames(gs), function(sn) {
      nrow(exprs(gh_pop_get_data(gs[[sn]], "root")))
    }, numeric(1))
  ) %>%
    mutate(sample_id = tools::file_path_sans_ext(sample_name))

  fraction_df <- tibble(
    sample_id = names(filtered_dna_area$raw),
    singlet_events = lengths(filtered_dna_area$raw),
    fraction_above_threshold = vapply(filtered_dna_area$raw, function(x) {
      x <- x[is.finite(x)]
      if (length(x) == 0) return(NA_real_)
      mean(x > dna_threshold)
    }, numeric(1))
  )

  plot_df <- total_event_df %>%
    left_join(fraction_df, by = "sample_id") %>%
    left_join(select(sample_metadata, sample_id, cell_line, condition), by = "sample_id")

  ggplot(
    plot_df,
    aes(x = total_events, y = fraction_above_threshold, color = condition, shape = cell_line)
  ) +
    geom_point(size = 2.2, alpha = 0.9, na.rm = TRUE) +
    scale_x_continuous(labels = scales::label_comma()) +
    scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
    labs(
      title = sprintf("DNA-A > %s Fraction vs Total Events", scales::comma(dna_threshold)),
      x = "Total Events",
      y = "Fraction of DNA Singlets Above Threshold",
      color = "Condition",
      shape = "Cell Line"
    ) +
    theme_bw()
}

raw_dna_plot_palette <- function(group_values) {
  base_palette <- c(
    "Controls" = "#1b9e77",
    "A1" = "#7570b3",
    "AOA" = "#e7298a",
    "AN" = "#66a61e",
    "H1" = "#e6ab02",
    "HN" = "#a6761d",
    "HOH" = "#1f78b4",
    "N1" = "#b2df8a",
    "Unannotated" = "#666666"
  )

  missing_groups <- setdiff(group_values, names(base_palette))
  if (length(missing_groups) > 0) {
    extra_cols <- grDevices::hcl.colors(length(missing_groups), palette = "Dark 3")
    names(extra_cols) <- missing_groups
    base_palette <- c(base_palette, extra_cols)
  }

  base_palette[group_values]
}

build_raw_dna_plot_df <- function(filtered_dna_area, sample_metadata, max_signal = Inf) {
  sample_tbl <- tibble(sample_id = names(filtered_dna_area$raw)) %>%
    left_join(
      select(
        sample_metadata,
        sample_id,
        cell_line,
        condition,
        exposure_condition,
        passage_label,
        passage_number,
        acquisition_id,
        sample_role
      ),
      by = "sample_id"
    ) %>%
    mutate(
      condition_plot = dplyr::coalesce(condition, exposure_condition, "Unannotated"),
      is_control = sample_role %in% c("control", "reference") |
        condition_plot %in% c("2N_Control", "4N_Control") |
        sample_id == "Sample_CEN",
      overlay_group = ifelse(is_control, "Controls", condition_plot),
      sample_label = sample_id
    )

  plot_df <- bind_rows(lapply(names(filtered_dna_area$raw), function(sample_id) {
    values <- filtered_dna_area$raw[[sample_id]]
    values <- values[is.finite(values) & values < max_signal]
    if (length(values) == 0) {
      return(NULL)
    }

    tibble(
      sample_id = sample_id,
      raw_dna_a = values
    )
  }))

  if (nrow(plot_df) == 0) {
    stop(sprintf("No raw DNA-A values remained after filtering below %s.", scales::comma(max_signal)))
  }

  plot_df %>%
    left_join(sample_tbl, by = "sample_id")
}

build_raw_dna_plot_context <- function(filtered_dna_area, sample_metadata) {
  raw_all_df <- build_raw_dna_plot_df(filtered_dna_area, sample_metadata, max_signal = Inf)
  finite_values <- raw_all_df$raw_dna_a[is.finite(raw_all_df$raw_dna_a)]
  if (length(finite_values) == 0) {
    stop("No finite raw DNA-A values were available.")
  }

  unique_values <- sort(unique(finite_values))
  if (length(unique_values) < 2) {
    stop("Could not derive a raw DNA-A bin width from the filtered singlet vectors.")
  }

  diffs <- diff(unique_values)
  diffs <- diffs[is.finite(diffs) & diffs > 0]
  if (length(diffs) == 0) {
    stop("Could not derive a positive raw DNA-A bin width from the filtered singlet vectors.")
  }

  group_values <- unique(c("Controls", sort(unique(raw_all_df$overlay_group[!raw_all_df$is_control])), "Unannotated"))

  list(
    raw_all_df = raw_all_df,
    bin_width = min(diffs),
    color_values = raw_dna_plot_palette(group_values)
  )
}

interpolate_half_height_crossing <- function(x_left, y_left, x_right, y_right, target_y) {
  if (!is.finite(x_left) || !is.finite(y_left) || !is.finite(x_right) || !is.finite(y_right)) {
    return(NA_real_)
  }
  if (y_left == y_right) {
    return(mean(c(x_left, x_right)))
  }
  x_left + (target_y - y_left) * (x_right - x_left) / (y_right - y_left)
}

summarise_threshold_peak <- function(values, side = c("below", "above"), threshold = 12000, density_n = 1024) {
  side <- match.arg(side)
  values <- values[is.finite(values)]
  values <- if (side == "below") values[values < threshold] else values[values > threshold]

  if (length(values) < 10 || length(unique(values)) < 2) {
    return(tibble(
      peak_location = NA_real_,
      peak_height = NA_real_,
      half_height_left = NA_real_,
      half_height_right = NA_real_,
      peak_width_fwhm = NA_real_,
      peak_n = length(values)
    ))
  }

  density_obj <- density(values, n = density_n, from = min(values), to = max(values), na.rm = TRUE)
  x <- density_obj$x
  y <- density_obj$y
  if (length(x) < 3 || all(!is.finite(y))) {
    return(tibble(
      peak_location = NA_real_,
      peak_height = NA_real_,
      half_height_left = NA_real_,
      half_height_right = NA_real_,
      peak_width_fwhm = NA_real_,
      peak_n = length(values)
    ))
  }

  local_max_idx <- which(y >= dplyr::lag(y, default = -Inf) & y >= dplyr::lead(y, default = -Inf))
  if (length(local_max_idx) == 0) {
    local_max_idx <- which.max(y)
  }
  peak_idx <- local_max_idx[which.max(y[local_max_idx])]
  peak_x <- x[peak_idx]
  peak_y <- y[peak_idx]
  half_height <- peak_y / 2

  left_idx <- which(y[seq_len(peak_idx)] < half_height)
  if (length(left_idx) == 0) {
    left_cross <- x[1]
  } else {
    left_low_idx <- max(left_idx)
    left_high_idx <- min(left_low_idx + 1, peak_idx)
    left_cross <- interpolate_half_height_crossing(
      x[left_low_idx], y[left_low_idx],
      x[left_high_idx], y[left_high_idx],
      half_height
    )
  }

  right_range <- peak_idx:length(y)
  right_idx_rel <- which(y[right_range] < half_height)
  if (length(right_idx_rel) == 0) {
    right_cross <- x[length(x)]
  } else {
    right_low_idx <- right_range[min(right_idx_rel)] - 1
    right_low_idx <- max(right_low_idx, peak_idx)
    right_high_idx <- min(right_low_idx + 1, length(x))
    right_cross <- interpolate_half_height_crossing(
      x[right_low_idx], y[right_low_idx],
      x[right_high_idx], y[right_high_idx],
      half_height
    )
  }

  tibble(
    peak_location = peak_x,
    peak_height = peak_y,
    half_height_left = left_cross,
    half_height_right = right_cross,
    peak_width_fwhm = right_cross - left_cross,
    peak_n = length(values)
  )
}

build_threshold_peak_summary_df <- function(raw_all_df, threshold = 12000) {
  non_control_df <- raw_all_df %>% filter(!is_control)
  if (nrow(non_control_df) == 0) {
    stop("No non-control samples available for threshold peak summaries.")
  }

  split_values <- split(non_control_df$raw_dna_a, non_control_df$sample_label)
  sample_meta <- non_control_df %>%
    distinct(sample_label, overlay_group) %>%
    arrange(overlay_group, sample_label)

  summary_df <- bind_rows(lapply(names(split_values), function(sample_label) {
    values <- split_values[[sample_label]]
    below_tbl <- summarise_threshold_peak(values, side = "below", threshold = threshold)
    above_tbl <- summarise_threshold_peak(values, side = "above", threshold = threshold)
    tibble(
      sample_label = sample_label,
      below_peak_location = below_tbl$peak_location,
      below_peak_height = below_tbl$peak_height,
      below_half_height_left = below_tbl$half_height_left,
      below_half_height_right = below_tbl$half_height_right,
      below_peak_width_fwhm = below_tbl$peak_width_fwhm,
      below_peak_n = below_tbl$peak_n,
      above_peak_location = above_tbl$peak_location,
      above_peak_height = above_tbl$peak_height,
      above_half_height_left = above_tbl$half_height_left,
      above_half_height_right = above_tbl$half_height_right,
      above_peak_width_fwhm = above_tbl$peak_width_fwhm,
      above_peak_n = above_tbl$peak_n,
      fraction_above_threshold = mean(values > threshold)
    )
  })) %>%
    left_join(sample_meta, by = "sample_label")

  summary_df %>%
    mutate(sample_label = factor(sample_label, levels = sample_meta$sample_label))
}

plot_threshold_peak_locations <- function(peak_summary_df, color_values, threshold = 12000) {
  plot_df <- bind_rows(
    peak_summary_df %>%
      transmute(
        sample_label,
        overlay_group,
        side = sprintf("Below %s", scales::comma(threshold)),
        peak_location = below_peak_location,
        whisker_left = below_half_height_left,
        whisker_right = below_half_height_right
      ),
    peak_summary_df %>%
      transmute(
        sample_label,
        overlay_group,
        side = sprintf("Above %s", scales::comma(threshold)),
        peak_location = above_peak_location,
        whisker_left = above_half_height_left,
        whisker_right = above_half_height_right
      )
  )

  ggplot(plot_df, aes(x = peak_location, y = sample_label, color = overlay_group, shape = side)) +
    geom_segment(
      aes(x = whisker_left, xend = whisker_right, y = sample_label, yend = sample_label),
      linewidth = 0.8,
      alpha = 0.85,
      na.rm = TRUE
    ) +
    geom_point(size = 2.6, na.rm = TRUE) +
    scale_color_manual(values = color_values, drop = FALSE) +
    labs(
      title = "Dominant peak locations above and below threshold",
      subtitle = sprintf("Controls excluded. Points mark the highest smoothed-density peak on each side of %s; whiskers show the corresponding FWHM.", scales::comma(threshold)),
      x = "Peak location",
      y = NULL,
      color = "Overlay group",
      shape = "Peak side"
    ) +
    theme_bw(base_size = 13.5) +
    theme(
      axis.text.y = element_text(size = 9.5),
      panel.grid.minor = element_blank()
    )
}

plot_threshold_peak_location_covariation <- function(peak_summary_df, color_values, threshold = 12000) {
  line_fit <- lm(above_peak_location ~ below_peak_location, data = peak_summary_df)
  line_summary <- summary(line_fit)
  fit_grid <- tibble(
    below_peak_location = seq(
      min(peak_summary_df$below_peak_location, na.rm = TRUE),
      max(peak_summary_df$below_peak_location, na.rm = TRUE),
      length.out = 200
    )
  )
  fit_grid$above_peak_location <- predict(line_fit, newdata = fit_grid)

  annotation_df <- tibble(
    below_peak_location = min(peak_summary_df$below_peak_location, na.rm = TRUE) + 0.05 * diff(range(peak_summary_df$below_peak_location, na.rm = TRUE)),
    above_peak_location = max(peak_summary_df$above_peak_location, na.rm = TRUE) - 0.05 * diff(range(peak_summary_df$above_peak_location, na.rm = TRUE)),
    label = sprintf(
      "line slope = %.4f\nline R^2 = %.4f",
      unname(coef(line_fit)[["below_peak_location"]]),
      line_summary$r.squared
    )
  )

  ggplot(peak_summary_df, aes(x = below_peak_location, y = above_peak_location, color = overlay_group)) +
    geom_point(size = 2.8, na.rm = TRUE) +
    geom_line(
      data = fit_grid,
      inherit.aes = FALSE,
      aes(x = below_peak_location, y = above_peak_location),
      color = "black",
      linewidth = 0.8
    ) +
    geom_label(
      data = annotation_df,
      inherit.aes = FALSE,
      aes(x = below_peak_location, y = above_peak_location, label = label),
      hjust = 0,
      vjust = 1,
      size = 3.5,
      label.size = 0.25
    ) +
    scale_color_manual(values = color_values, drop = FALSE) +
    labs(
      title = "Above-threshold peak location vs below-threshold peak location",
      subtitle = sprintf("Controls excluded. Both axes come from the highest smoothed-density peak on each side of %s.", scales::comma(threshold)),
      x = "Below-threshold peak location",
      y = "Above-threshold peak location",
      color = "Overlay group"
    ) +
    theme_bw(base_size = 13.5) +
    theme(panel.grid.minor = element_blank())
}

plot_threshold_peak_width_covariation <- function(peak_summary_df, color_values, threshold = 12000) {
  ggplot(peak_summary_df, aes(x = below_peak_width_fwhm, y = above_peak_width_fwhm, color = overlay_group)) +
    geom_point(size = 2.8, na.rm = TRUE) +
    scale_color_manual(values = color_values, drop = FALSE) +
    labs(
      title = "Above-threshold peak width vs below-threshold peak width",
      subtitle = sprintf("Controls excluded. Width is FWHM of the highest smoothed-density peak on each side of %s.", scales::comma(threshold)),
      x = "Below-threshold peak width (FWHM)",
      y = "Above-threshold peak width (FWHM)",
      color = "Overlay group"
    ) +
    theme_bw(base_size = 13.5) +
    theme(panel.grid.minor = element_blank())
}

plot_threshold_peak_vs_fraction_above_threshold <- function(peak_summary_df, color_values, threshold = 12000) {
  ggplot(peak_summary_df, aes(x = above_peak_location, y = fraction_above_threshold, color = overlay_group)) +
    geom_point(size = 2.8, na.rm = TRUE) +
    scale_color_manual(values = color_values, drop = FALSE) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(
      title = "Fraction of cells above threshold vs above-threshold peak location",
      subtitle = sprintf("Controls excluded. Peak location is the highest smoothed-density maximum above %s.", scales::comma(threshold)),
      x = "Above-threshold peak location",
      y = "Fraction of cells above threshold",
      color = "Overlay group"
    ) +
    theme_bw(base_size = 13.5) +
    theme(panel.grid.minor = element_blank())
}
```

# Build or Load Dataset Results

```{r dataset-results}
dataset_results <- lapply(dataset_configs, build_or_load_dataset)
```

# Cross-Dataset Gate Counts

```{r gate-counts}
dataset_gate_overview <- bind_rows(lapply(dataset_results, function(res) {
  gate_summary <- dataset_gate_summary(res$gate_counts)
  gate_summary$dataset_id <- res$config$dataset_id
  gate_summary
})) %>%
  select(dataset_id, everything())

if (should_print_table(dataset_gate_overview)) {
  kable(dataset_gate_overview)
}
```

# Cross-Dataset Metadata

```{r metadata-summary}
dataset_metadata_overview <- bind_rows(lapply(dataset_results, function(res) {
  metadata_summary <- dataset_metadata_summary(res$input_tbl)
  metadata_summary$dataset_id <- res$config$dataset_id
  metadata_summary
})) %>%
  select(dataset_id, everything())

if (should_print_table(dataset_metadata_overview)) {
  kable(dataset_metadata_overview)
}
```

# Figure Galleries

```{r dataset-sections, results='asis'}
dataset_detail_tbl <- bind_rows(lapply(dataset_results, function(res) {
  cfg <- res$config
  tibble(
    dataset_name = cfg$dataset_name,
    dataset_id = cfg$dataset_id,
    source_directory = cfg$dataset_dir,
    cache_path = cfg$gs_cache_path,
    figure_directory = cfg$figure_dir
  )
}))

if (should_print_table(dataset_detail_tbl, max_rows = 12, max_cols = 5)) {
  kable(dataset_detail_tbl)
  cat("\n\n")
}

dataset_plot_specs <- lapply(dataset_results, function(res) {
  local({
    cfg <- res$config
    gs <- res$gs
    filtered_dna_area <- res$filtered_dna_area
    sample_metadata <- res$sample_metadata
    raw_dna_plot_context <- build_raw_dna_plot_context(filtered_dna_area, sample_metadata)
    fluorescence_channels <- discover_other_fluorescence_area_channels(gs, cfg$dna_channels)
    raw_dna_peak_summary_df <- build_threshold_peak_summary_df(
      raw_all_df = raw_dna_plot_context$raw_all_df,
      threshold = cfg$raw_dna_reference_threshold
    )

    list(
      dataset_name = cfg$dataset_name,
      dataset_id = cfg$dataset_id,
      plot_paths = cfg$plot_paths,
      builders = list(
        fsc_ssc = function() build_fsc_ssc_plot(gs),
        fsc_time = function() build_fsc_time_plot(gs, cfg$time_step_seconds, cfg$dna_channels[1]),
        dna_scatter = function() build_dna_scatter_plot(gs, cfg$dna_channels),
        dna_density = function() build_dna_density_plot(
          gs,
          cfg$dna_channels,
          cfg$raw_dna_reference_threshold,
          cfg$dna_cofactor
        ),
        fluorescence_area_histograms = function() build_fluorescence_area_histogram_plot(
          gs,
          fluorescence_channels
        ),
        fluorescence_vs_dna = function() build_fluorescence_vs_dna_plot(
          gs,
          dna_channel = cfg$dna_channels[1],
          fluorescence_channels = fluorescence_channels
        ),
        dna_fraction_summary = function() build_dna_fraction_summary_plot(
          gs,
          filtered_dna_area,
          sample_metadata,
          cfg$dna_fraction_threshold
        ),
        raw_dna_peak_locations = function() plot_threshold_peak_locations(
          peak_summary_df = raw_dna_peak_summary_df,
          color_values = raw_dna_plot_context$color_values,
          threshold = cfg$raw_dna_reference_threshold
        ),
        raw_dna_peak_location_covariation = function() plot_threshold_peak_location_covariation(
          peak_summary_df = raw_dna_peak_summary_df,
          color_values = raw_dna_plot_context$color_values,
          threshold = cfg$raw_dna_reference_threshold
        ),
        raw_dna_peak_width_covariation = function() plot_threshold_peak_width_covariation(
          peak_summary_df = raw_dna_peak_summary_df,
          color_values = raw_dna_plot_context$color_values,
          threshold = cfg$raw_dna_reference_threshold
        ),
        raw_dna_peak_vs_fraction_above_threshold = function() plot_threshold_peak_vs_fraction_above_threshold(
          peak_summary_df = raw_dna_peak_summary_df,
          color_values = raw_dna_plot_context$color_values,
          threshold = cfg$raw_dna_reference_threshold
        )
      )
    )
  })
})

render_dataset_gallery <- function(plot_key, width, height, element_id) {
  plot_paths <- vapply(dataset_plot_specs, function(spec) spec$plot_paths[[plot_key]], character(1))
  plot_builders <- lapply(dataset_plot_specs, function(spec) spec$builders[[plot_key]])
  captions <- vapply(
    dataset_plot_specs,
    function(spec) sprintf("%s (%s)", spec$dataset_name, spec$dataset_id),
    character(1)
  )

  cat(render_plot_gallery(
    plot_paths = plot_paths,
    plot_builders = plot_builders,
    width = width,
    height = height,
    element_id = element_id,
    captions = captions,
    dpi = 300
  ))
  cat("\n\n")
}

cat("## FSC vs SSC\n\n")
render_dataset_gallery("fsc_ssc", width = 21, height = 14, element_id = "gallery-fsc-ssc")

cat("## FSC-A vs Time\n\n")
render_dataset_gallery("fsc_time", width = 21, height = 7, element_id = "gallery-fsc-time")

cat("## DNA Scatter\n\n")
render_dataset_gallery("dna_scatter", width = 21, height = 14, element_id = "gallery-dna-scatter")

cat("## DNA Density\n\n")
render_dataset_gallery("dna_density", width = 21, height = 14, element_id = "gallery-dna-density")

cat("## Other Fluorescence Area Histograms\n\n")
render_dataset_gallery("fluorescence_area_histograms", width = 24, height = 16, element_id = "gallery-fluorescence-area-histograms")

cat("## Other Fluorescence vs DNA-A\n\n")
render_dataset_gallery("fluorescence_vs_dna", width = 24, height = 16, element_id = "gallery-fluorescence-vs-dna")

cat("## DNA Threshold Fraction Summary\n\n")
render_dataset_gallery("dna_fraction_summary", width = 10, height = 7, element_id = "gallery-dna-fraction-summary")

cat("## Raw DNA-A Peak Locations\n\n")
render_dataset_gallery("raw_dna_peak_locations", width = 12, height = 8, element_id = "gallery-raw-dna-peak-locations")

cat("## Raw DNA-A Peak Location Covariation\n\n")
render_dataset_gallery("raw_dna_peak_location_covariation", width = 12, height = 8, element_id = "gallery-raw-dna-peak-location-covariation")

cat("## Raw DNA-A Peak Width Covariation\n\n")
render_dataset_gallery("raw_dna_peak_width_covariation", width = 12, height = 8, element_id = "gallery-raw-dna-peak-width-covariation")

cat("## Raw DNA-A Tail Fraction vs Above-Threshold Peak\n\n")
render_dataset_gallery("raw_dna_peak_vs_fraction_above_threshold", width = 12, height = 8, element_id = "gallery-raw-dna-peak-vs-fraction")
```

```{r activate-zoom, results='asis'}
cat('
<script>
window.plotGalleryState = window.plotGalleryState || {};

function initPlotGallery(id) {
  var gallery = document.getElementById(id);
  if (!gallery) return;

  var slides = gallery.querySelectorAll(".plot-slide");
  var statusEl = gallery.querySelector(".plot-gallery-status");
  if (!slides.length) return;

  window.plotGalleryState[id] = {
    i: 0,
    n: slides.length,
    panzoomInit: {}
  };

  updatePlotGallery(id);
}

function updatePlotGallery(id) {
  var gallery = document.getElementById(id);
  var state = window.plotGalleryState[id];
  if (!gallery || !state) return;

  var slides = gallery.querySelectorAll(".plot-slide");
  var statusEl = gallery.querySelector(".plot-gallery-status");

  Array.prototype.forEach.call(slides, function(slide, idx) {
    slide.style.display = idx === state.i ? "block" : "none";
  });

  if (statusEl) {
    statusEl.textContent = (state.i + 1) + " / " + state.n;
  }

  var activeSlide = slides[state.i];
  if (!activeSlide) return;

  var img = activeSlide.querySelector("img.zoomable-plot");
  if (!img || typeof panzoom === "undefined") return;

  if (!state.panzoomInit[state.i]) {
    panzoom(img, { maxZoom: 10, minZoom: 1 });
    state.panzoomInit[state.i] = true;
  }
}

function plotGalleryNext(id) {
  var state = window.plotGalleryState[id];
  if (!state) return;
  state.i = (state.i + 1) % state.n;
  updatePlotGallery(id);
}

function plotGalleryPrev(id) {
  var state = window.plotGalleryState[id];
  if (!state) return;
  state.i = (state.i - 1 + state.n) % state.n;
  updatePlotGallery(id);
}

document.addEventListener("DOMContentLoaded", function() {
  document.querySelectorAll(".plot-gallery").forEach(function(el) {
    initPlotGallery(el.id);
  });
});
</script>
')
```
