---
title: "Fixed-Init Anoxia Checkpoint"
knit: !expr function(input_file, encoding) rmarkdown::render(input_file, encoding = encoding, output_dir = dirname(input_file), intermediates_dir = dirname(input_file), envir = new.env(parent = globalenv()))
output:
  html_document:
    toc: true
    toc_float: true
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
---

```{r setup, include=FALSE}
input_path <- knitr::current_input(dir = TRUE)
workflow_dir <- if (length(input_path) == 0) {
  normalizePath(getwd(), winslash = "/", mustWork = TRUE)
} else {
  normalizePath(dirname(input_path), winslash = "/", mustWork = TRUE)
}
project_root <- normalizePath(file.path(workflow_dir, ".."), winslash = "/", mustWork = TRUE)
if (!dir.exists(file.path(project_root, "data")) && dir.exists(file.path(getwd(), "data"))) {
  project_root <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
}
knitr::opts_knit$set(root.dir = workflow_dir)
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  autodep = TRUE,
  fig.align = "center",
  out.width = "100%",
  fig.path = "fixed_init_anoxia_checkpoint_files/figure-html/",
  cache.path = "cache/fixed_init_anoxia_checkpoint/"
)

```

# Goal

Explore the archived fixed-initialization Stan optimization checkpoint for the
anoxia-only subset.

This was chosen because - unlike the LTEE experiments ('hypoxia' samples), the CEN peaks all seemed to be in exactly the same location here, and the 2N/4N peaks appeared pretty consistent too.

This document reads the optimizer CSV from
`results/2026-04-10-fixed-init-anoxia/`, reconstructs the fitted mixture
components from the saved parameters, and overlays them on the observed
DNA-area histograms.

```{r libraries, cache=FALSE}
library(ggplot2)
library(knitr)
source(file.path(project_root, "R", "fit_specs.R"))
```

```{r paths, cache=FALSE}
first_pass_spec <- get_fit_spec("first_pass", project_root = project_root)
dye_ratio_spec <- get_fit_spec("dye_ratio", project_root = project_root)
observed_cen_scaled_spec <- get_fit_spec("observed_cen_scaled", project_root = project_root)

experiment_filter <- first_pass_spec$experiment_filter
checkpoint_dir <- first_pass_spec$checkpoint_dir
optimizer_csv <- first_pass_spec$report_csv_path
dye_ratio_optimizer_csv <- dye_ratio_spec$report_csv_path
observed_cen_scaled_dir <- file.path(checkpoint_dir, "fits", "observed_cen_scaled")
observed_cen_scaled_optimizer_csv <- observed_cen_scaled_spec$report_csv_path
observed_cen_scaled_manifest_path <- observed_cen_scaled_spec$report_manifest_path
stan_data_path <- first_pass_spec$report_stan_data_path
sample_metadata_path <- first_pass_spec$report_sample_metadata_path
observed_cen_scaled_excluded_sample_patterns <- observed_cen_scaled_spec$exclude_sample_patterns

assert_scalar_path_exists <- function(path_value, label) {
  if (length(path_value) != 1 || !is.character(path_value) || is.na(path_value) || !nzchar(path_value)) {
    stop(sprintf("Expected %s to be a single non-empty character path.", label))
  }
  if (!file.exists(path_value)) {
    stop(sprintf("%s does not exist: %s", label, path_value))
  }
}

assert_scalar_path_exists(optimizer_csv, "optimizer_csv")
assert_scalar_path_exists(dye_ratio_optimizer_csv, "dye_ratio_optimizer_csv")
assert_scalar_path_exists(observed_cen_scaled_optimizer_csv, "observed_cen_scaled_optimizer_csv")
assert_scalar_path_exists(observed_cen_scaled_manifest_path, "observed_cen_scaled_manifest_path")
assert_scalar_path_exists(stan_data_path, "stan_data_path")
assert_scalar_path_exists(sample_metadata_path, "sample_metadata_path")
```

```{r helpers}
subset_stan_bundle_by_experiment <- function(stan_data, lut, experiment_pattern) {
  keep_experiments <- grepl(experiment_pattern, lut$experiment$experiment_name, ignore.case = TRUE)
  if (!any(keep_experiments)) {
    stop(sprintf("No experiments matched pattern '%s'.", experiment_pattern))
  }

  old_exp_ids <- lut$experiment$experiment_id[keep_experiments]
  keep_samples <- lut$sample$experiment_id %in% old_exp_ids
  old_sample_ids <- lut$sample$sample_id[keep_samples]
  keep_bins <- stan_data$sampleID %in% old_sample_ids

  exp_map <- setNames(seq_along(old_exp_ids), old_exp_ids)
  sample_map <- setNames(seq_along(old_sample_ids), old_sample_ids)

  experiment_lut <- lut$experiment[keep_experiments, , drop = FALSE]
  experiment_lut$experiment_id <- seq_len(nrow(experiment_lut))

  sample_lut <- lut$sample[keep_samples, , drop = FALSE]
  sample_lut$sample_id <- seq_len(nrow(sample_lut))
  sample_lut$experiment_id <- unname(exp_map[as.character(sample_lut$experiment_id)])

  stan_data$areaDNA_bin_center <- stan_data$areaDNA_bin_center[keep_bins]
  stan_data$count <- stan_data$count[keep_bins]
  stan_data$expID <- unname(exp_map[as.character(stan_data$expID[keep_bins])])
  stan_data$sampleID <- unname(sample_map[as.character(stan_data$sampleID[keep_bins])])
  stan_data$sampleExpID <- unname(exp_map[as.character(lut$sample$experiment_id[keep_samples])])
  stan_data$N <- length(stan_data$count)
  stan_data$N_exp <- nrow(experiment_lut)
  stan_data$N_sample <- nrow(sample_lut)

  list(
    stan_data = stan_data,
    lut = list(
      experiment = experiment_lut,
      sample = sample_lut
    )
  )
}

extract_indexed_vector <- function(row_df, prefix, n) {
  vapply(seq_len(n), function(i) row_df[[sprintf("%s.%d", prefix, i)]], numeric(1))
}

extract_theta_matrix <- function(row_df, n_sample) {
  theta <- matrix(NA_real_, nrow = n_sample, ncol = 7)
  for (s in seq_len(n_sample)) {
    for (k in seq_len(7)) {
      theta[s, k] <- row_df[[sprintf("theta.%d.%d", s, k)]]
    }
  }
  theta
}

s_phase_density <- function(x, lower_peak, upper_peak, n_subpeaks) {
  width <- upper_peak - lower_peak
  if (!is.finite(width) || width <= 0) {
    return(0)
  }
  dnorm(x, mean = 0.5 * (lower_peak + upper_peak), sd = 0.5 * width)
}

build_overlay_df <- function(stan_data, lut, mu_cen, ratio_tumor_cen_sample, R_4n_sample, R_8n_sample, cv, theta) {
  x <- stan_data$areaDNA_bin_center
  bin_width <- diff(sort(unique(x)))[1]

  plot_df <- data.frame(
    areaDNA_bin_center = stan_data$areaDNA_bin_center,
    count = stan_data$count,
    expID = stan_data$expID,
    sampleID = stan_data$sampleID,
    stringsAsFactors = FALSE
  )

  sample_totals <- aggregate(count ~ sampleID, data = plot_df, FUN = sum)
  names(sample_totals)[2] <- "sample_total"
  plot_df <- merge(plot_df, sample_totals, by = "sampleID", sort = FALSE)

  plot_df$mixture_count <- mapply(
    FUN = function(x_i, sample_id, exp_id, sample_total) {
      sig_cen <- mu_cen[exp_id] * cv
      mu_2n <- mu_cen[exp_id] * ratio_tumor_cen_sample[sample_id]
      mu_4n <- mu_2n * R_4n_sample[sample_id]
      mu_8n <- mu_4n * R_8n_sample[sample_id]
      sig_2n <- mu_2n * cv
      sig_4n <- mu_4n * cv
      sig_8n <- mu_8n * cv

      density_val <- theta[sample_id, 1] * dnorm(x_i, mean = mu_cen[exp_id], sd = sig_cen) +
        theta[sample_id, 2] * dnorm(x_i, mean = mu_2n, sd = sig_2n) +
        theta[sample_id, 3] * dnorm(x_i, mean = mu_4n, sd = sig_4n) +
        theta[sample_id, 4] * dnorm(x_i, mean = mu_8n, sd = sig_8n) +
        theta[sample_id, 5] * s_phase_density(x_i, mu_2n, mu_4n, stan_data$N_s_phase) +
        theta[sample_id, 6] * s_phase_density(x_i, mu_4n, mu_8n, stan_data$N_s_phase) +
        theta[sample_id, 7] * dnorm(x_i, mean = 2 * mu_cen[exp_id], sd = sqrt(2) * sig_cen)

      sample_total * bin_width * density_val
    },
    x_i = plot_df$areaDNA_bin_center,
    sample_id = plot_df$sampleID,
    exp_id = plot_df$expID,
    sample_total = plot_df$sample_total
  )

  peak_df <- data.frame(
    sampleID = seq_len(nrow(lut$sample)),
    expID = lut$sample$experiment_id,
    stringsAsFactors = FALSE
  )
  peak_df$mu_cen <- mu_cen[peak_df$expID]
  peak_df$mu_2n <- peak_df$mu_cen * ratio_tumor_cen_sample[peak_df$sampleID]
  peak_df$mu_4n <- peak_df$mu_2n * R_4n_sample[peak_df$sampleID]
  peak_df$mu_8n <- peak_df$mu_4n * R_8n_sample[peak_df$sampleID]
  peak_df$mu_4n_ratio_2 <- 2 * peak_df$mu_2n

  plot_df <- merge(
    plot_df,
    lut$sample[, c("sample_id", "sample_name", "experiment_name")],
    by.x = "sampleID",
    by.y = "sample_id",
    all.x = TRUE,
    sort = FALSE
  )

  plot_df <- merge(
    plot_df,
    peak_df[, c("sampleID", "mu_cen", "mu_2n", "mu_4n", "mu_8n", "mu_4n_ratio_2")],
    by = "sampleID",
    all.x = TRUE,
    sort = FALSE
  )

  plot_df$sample_label <- sprintf("%s | %s", plot_df$experiment_name, plot_df$sample_name)
  plot_df
}

add_sample_metadata <- function(plot_df, sample_metadata) {
  sample_metadata$sample_key <- tools::file_path_sans_ext(basename(sample_metadata$sample_name))

  metadata_cols <- c("sample_key", "sample_id", "cell_line", "condition")
  plot_df <- merge(
    plot_df,
    sample_metadata[, metadata_cols],
    by.x = "sample_name",
    by.y = "sample_key",
    all.x = TRUE,
    sort = FALSE
  )

  names(plot_df)[names(plot_df) == "sample_id"] <- "metadata_sample_id"
  plot_df$condition_plot <- ifelse(is.na(plot_df$condition), "Unannotated", plot_df$condition)
  plot_df$is_control <- plot_df$condition_plot %in% c("2N_Control", "4N_Control") | plot_df$sample_name == "Sample_CEN"
  plot_df
}

ensure_plot_file <- function(plot_path, plot_builder, width, height, dpi = 300) {
  full_plot_path <- file.path(workflow_dir, plot_path)
  dir.create(dirname(full_plot_path), recursive = TRUE, showWarnings = FALSE)

  if (!file.exists(full_plot_path)) {
    plot_obj <- plot_builder()
    ggplot2::ggsave(
      filename = full_plot_path,
      plot = plot_obj,
      width = width,
      height = height,
      dpi = dpi,
      units = "in",
      limitsize = FALSE
    )
  }
}

render_or_include_plot <- function(plot_path, plot_builder, width, height, element_id, dpi = 300) {
  ensure_plot_file(plot_path, plot_builder, width, height, dpi)

  plot_src <- gsub("\\\\", "/", plot_path)
  sprintf(
    '<img src="%s" id="%s" style="width:100%%;" />',
    plot_src,
    element_id
  )
}

safe_plot_slug <- function(x) {
  x <- tolower(x)
  x <- gsub("[^a-z0-9]+", "-", x)
  gsub("(^-+|-+$)", "", x)
}

build_raw_dna_density_df <- function(raw_dna_area, sample_metadata, max_signal = 12000) {
  sample_metadata$sample_key <- tools::file_path_sans_ext(basename(sample_metadata$sample_name))

  density_df <- do.call(
    rbind,
    lapply(names(raw_dna_area), function(sample_name) {
      values <- raw_dna_area[[sample_name]]
      values <- values[is.finite(values) & values < max_signal]
      if (length(values) == 0) {
        return(NULL)
      }

      data.frame(
        sample_name = sample_name,
        raw_dna_a = values,
        stringsAsFactors = FALSE
      )
    })
  )

  if (is.null(density_df) || nrow(density_df) == 0) {
    stop(sprintf("No raw DNA-A values remained after filtering below %s.", max_signal))
  }

  density_df <- merge(
    density_df,
    sample_metadata[, c("sample_key", "condition")],
    by.x = "sample_name",
    by.y = "sample_key",
    all.x = TRUE,
    sort = FALSE
  )

  density_df$condition_plot <- ifelse(is.na(density_df$condition), "Unannotated", density_df$condition)
  density_df$is_control <- density_df$condition_plot %in% c("2N_Control", "4N_Control") | density_df$sample_name == "Sample_CEN"
  density_df$overlay_group <- ifelse(density_df$is_control, "Controls", density_df$condition_plot)
  density_df$sample_label <- density_df$sample_name
  density_df
}

build_filtered_raw_dna_df <- function(raw_dna_area, sample_metadata, max_signal) {
  sample_metadata$sample_key <- tools::file_path_sans_ext(basename(sample_metadata$sample_name))

  filtered_df <- do.call(
    rbind,
    lapply(names(raw_dna_area), function(sample_name) {
      values <- raw_dna_area[[sample_name]]
      values <- values[is.finite(values) & values < max_signal]
      if (length(values) == 0) {
        return(NULL)
      }

      data.frame(
        sample_name = sample_name,
        raw_dna_a = values,
        stringsAsFactors = FALSE
      )
    })
  )

  if (is.null(filtered_df) || nrow(filtered_df) == 0) {
    stop(sprintf("No raw DNA-A values remained after filtering below %s.", max_signal))
  }

  filtered_df <- merge(
    filtered_df,
    sample_metadata[, c("sample_key", "condition")],
    by.x = "sample_name",
    by.y = "sample_key",
    all.x = TRUE,
    sort = FALSE
  )

  filtered_df$condition_plot <- ifelse(is.na(filtered_df$condition), "Unannotated", filtered_df$condition)
  filtered_df$is_control <- filtered_df$condition_plot %in% c("2N_Control", "4N_Control") | filtered_df$sample_name == "Sample_CEN"
  filtered_df$overlay_group <- ifelse(filtered_df$is_control, "Controls", filtered_df$condition_plot)
  filtered_df$sample_label <- filtered_df$sample_name
  filtered_df
}

build_raw_dna_all_df <- function(raw_dna_area, sample_metadata) {
  sample_metadata$sample_key <- tools::file_path_sans_ext(basename(sample_metadata$sample_name))

  raw_df <- do.call(
    rbind,
    lapply(names(raw_dna_area), function(sample_name) {
      values <- raw_dna_area[[sample_name]]
      values <- values[is.finite(values)]
      if (length(values) == 0) {
        return(NULL)
      }

      data.frame(
        sample_name = sample_name,
        raw_dna_a = values,
        stringsAsFactors = FALSE
      )
    })
  )

  if (is.null(raw_df) || nrow(raw_df) == 0) {
    stop("No raw DNA-A values were available.")
  }

  raw_df <- merge(
    raw_df,
    sample_metadata[, c("sample_key", "condition")],
    by.x = "sample_name",
    by.y = "sample_key",
    all.x = TRUE,
    sort = FALSE
  )

  raw_df$condition_plot <- ifelse(is.na(raw_df$condition), "Unannotated", raw_df$condition)
  raw_df$is_control <- raw_df$condition_plot %in% c("2N_Control", "4N_Control") | raw_df$sample_name == "Sample_CEN"
  raw_df$overlay_group <- ifelse(raw_df$is_control, "Controls", raw_df$condition_plot)
  raw_df$sample_label <- raw_df$sample_name
  raw_df
}

build_condition_palette <- function(condition_values) {
  base_palette <- c(
    "Controls" = "#1b9e77",
    "A1" = "#7570b3",
    "AOA" = "#e7298a",
    "AN" = "#66a61e",
    "H1" = "#e6ab02",
    "HN" = "#a6761d",
    "HOH" = "#1f78b4",
    "N1" = "#b2df8a",
    "Unannotated" = "#666666"
  )

  missing_conditions <- setdiff(condition_values, names(base_palette))
  if (length(missing_conditions) > 0) {
    extra_cols <- grDevices::hcl.colors(length(missing_conditions), palette = "Dark 3")
    names(extra_cols) <- missing_conditions
    base_palette <- c(base_palette, extra_cols)
  }

  base_palette[condition_values]
}

plot_raw_dna_quantiles <- function(density_df, color_values, max_signal = 12000) {
  sample_order_df <- unique(density_df[, c("sample_label", "overlay_group", "is_control")])
  sample_order_df <- sample_order_df[order(sample_order_df$is_control, sample_order_df$overlay_group, sample_order_df$sample_label), , drop = FALSE]
  density_df$sample_label <- factor(density_df$sample_label, levels = sample_order_df$sample_label)

  ggplot(density_df, aes(x = sample_label, y = raw_dna_a, color = overlay_group)) +
    geom_boxplot(
      width = 0.65,
      outlier.shape = NA,
      fill = NA,
      linewidth = 0.7
    ) +
    coord_flip() +
    scale_color_manual(values = color_values, drop = FALSE) +
    labs(
      title = sprintf("Raw DNA-A quantiles per sample after filtering below %s", scales::comma(max_signal)),
      subtitle = "Samples are ordered by the same overlay grouping used in the histogram panels.",
      x = NULL,
      y = "Raw DNA-A signal",
      color = "Overlay group"
    ) +
    theme_bw(base_size = 13.5) +
    theme(
      axis.text.y = element_text(size = 9.5),
      panel.grid.minor = element_blank()
    )
}

build_mode_bin_df <- function(density_df, bin_width) {
  non_control_df <- density_df[!density_df$is_control, , drop = FALSE]
  if (nrow(non_control_df) == 0) {
    stop("No non-control samples available for mode-bin plotting.")
  }

  non_control_df$bin_center <- floor(non_control_df$raw_dna_a / bin_width) * bin_width + (bin_width / 2)

  count_df <- aggregate(
    raw_dna_a ~ sample_label + overlay_group + bin_center,
    data = non_control_df,
    FUN = length
  )
  names(count_df)[names(count_df) == "raw_dna_a"] <- "bin_count"

  mode_idx <- unlist(tapply(seq_len(nrow(count_df)), count_df$sample_label, function(idx) {
    sample_counts <- count_df[idx, , drop = FALSE]
    idx[order(-sample_counts$bin_count, sample_counts$bin_center)[1]]
  }))

  mode_df <- count_df[mode_idx, , drop = FALSE]
  sample_order_df <- unique(mode_df[, c("sample_label", "overlay_group")])
  sample_order_df <- sample_order_df[order(sample_order_df$overlay_group, sample_order_df$sample_label), , drop = FALSE]
  mode_df$sample_label <- factor(mode_df$sample_label, levels = sample_order_df$sample_label)
  mode_df
}

plot_mode_bin_locations <- function(mode_df, color_values, bin_width) {
  ggplot(mode_df, aes(x = bin_center, y = sample_label, color = overlay_group)) +
    geom_point(size = 2.6) +
    scale_color_manual(values = color_values, drop = FALSE) +
    labs(
      title = "Most-populated raw DNA-A bin per non-control sample",
      subtitle = sprintf("Computed from filtered raw DNA-A values below 12,000 using bin width %s.", scales::comma(bin_width)),
      x = "Location of most-populated bin",
      y = NULL,
      color = "Overlay group"
    ) +
    theme_bw(base_size = 13.5) +
    theme(
      axis.text.y = element_text(size = 9.5),
      panel.grid.minor = element_blank()
    )
}

build_mode_vs_median_df <- function(raw_all_df, raw_filtered_df, bin_width) {
  mode_df <- build_mode_bin_df(
    density_df = raw_all_df,
    bin_width = bin_width
  )

  median_df <- aggregate(
    raw_dna_a ~ sample_label + overlay_group,
    data = raw_filtered_df[!raw_filtered_df$is_control, , drop = FALSE],
    FUN = median
  )
  names(median_df)[names(median_df) == "raw_dna_a"] <- "median_filtered_raw_dna_a"

  scatter_df <- merge(
    mode_df[, c("sample_label", "overlay_group", "bin_center")],
    median_df,
    by = c("sample_label", "overlay_group"),
    all = FALSE,
    sort = FALSE
  )

  sample_order_df <- unique(scatter_df[, c("sample_label", "overlay_group")])
  sample_order_df <- sample_order_df[order(sample_order_df$overlay_group, sample_order_df$sample_label), , drop = FALSE]
  scatter_df$sample_label <- factor(scatter_df$sample_label, levels = sample_order_df$sample_label)
  scatter_df
}

plot_mode_vs_median <- function(scatter_df, color_values, median_filter_threshold) {
  line_fit <- lm(bin_center ~ median_filtered_raw_dna_a, data = scatter_df)
  line_summary <- summary(line_fit)
  line_slope <- unname(coef(line_fit)[["median_filtered_raw_dna_a"]])
  line_r_squared <- line_summary$r.squared

  x_vals <- scatter_df$median_filtered_raw_dna_a
  y_vals <- scatter_df$bin_center
  line_annotation <- sprintf(
    "line slope = %.4f\nline R^2 = %.4f",
    line_slope,
    line_r_squared
  )

  fit_grid <- data.frame(
    median_filtered_raw_dna_a = seq(min(x_vals, na.rm = TRUE), max(x_vals, na.rm = TRUE), length.out = 200)
  )
  fit_grid$line_pred <- predict(line_fit, newdata = fit_grid)

  x_range <- range(x_vals, na.rm = TRUE)
  y_range <- range(y_vals, na.rm = TRUE)
  annotation_df <- data.frame(
    median_filtered_raw_dna_a = x_range[1] + 0.05 * diff(x_range),
    bin_center = y_range[2] - 0.05 * diff(y_range),
    label = line_annotation
  )

  ggplot(scatter_df, aes(x = median_filtered_raw_dna_a, y = bin_center, color = overlay_group)) +
    geom_point(size = 2.8) +
    geom_line(
      data = fit_grid,
      inherit.aes = FALSE,
      aes(x = median_filtered_raw_dna_a, y = line_pred),
      color = "black",
      linewidth = 0.8
    ) +
    geom_label(
      data = annotation_df,
      inherit.aes = FALSE,
      aes(x = median_filtered_raw_dna_a, y = bin_center, label = label),
      hjust = 0,
      vjust = 1,
      size = 3.5,
      label.size = 0.25
    ) +
    scale_color_manual(values = color_values, drop = FALSE) +
    labs(
      title = "Filtered raw DNA-A median vs most-populated raw DNA-A bin",
      subtitle = sprintf("Controls excluded. Median computed after filtering raw DNA-A below %s.", scales::comma(median_filter_threshold)),
      x = "Median raw DNA-A signal after filtering",
      y = "Location of most-populated raw DNA-A bin",
      color = "Overlay group"
    ) +
    theme_bw(base_size = 13.5) +
    theme(
      panel.grid.minor = element_blank()
    )
}

build_mode_vs_fraction_above_threshold_df <- function(raw_all_df, bin_width, threshold = 12000) {
  mode_df <- build_mode_bin_df(
    density_df = raw_all_df,
    bin_width = bin_width
  )

  non_control_df <- raw_all_df[!raw_all_df$is_control, , drop = FALSE]
  if (nrow(non_control_df) == 0) {
    stop("No non-control samples available for fraction-above-threshold plotting.")
  }

  fraction_df <- aggregate(
    raw_dna_a ~ sample_label + overlay_group,
    data = non_control_df,
    FUN = function(values) mean(values > threshold)
  )
  names(fraction_df)[names(fraction_df) == "raw_dna_a"] <- "fraction_above_threshold"

  scatter_df <- merge(
    mode_df[, c("sample_label", "overlay_group", "bin_center")],
    fraction_df,
    by = c("sample_label", "overlay_group"),
    all = FALSE,
    sort = FALSE
  )

  sample_order_df <- unique(scatter_df[, c("sample_label", "overlay_group")])
  sample_order_df <- sample_order_df[order(sample_order_df$overlay_group, sample_order_df$sample_label), , drop = FALSE]
  scatter_df$sample_label <- factor(scatter_df$sample_label, levels = sample_order_df$sample_label)
  scatter_df
}

plot_mode_vs_fraction_above_threshold <- function(scatter_df, color_values, threshold = 12000) {
  ggplot(scatter_df, aes(x = bin_center, y = fraction_above_threshold, color = overlay_group)) +
    geom_point(size = 2.8) +
    scale_color_manual(values = color_values, drop = FALSE) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(
      title = "Fraction of cells above DNA-A 12,000 vs most-populated raw DNA-A bin",
      subtitle = sprintf("Controls excluded. Fractions are computed from all finite raw DNA-A values using threshold %s.", scales::comma(threshold)),
      x = "Location of most-populated raw DNA-A bin",
      y = "Fraction of cells above threshold",
      color = "Overlay group"
    ) +
    theme_bw(base_size = 13.5) +
    theme(
      panel.grid.minor = element_blank()
    )
}

build_peak_line_df <- function(plot_df) {
  peak_line_df <- unique(plot_df[, c(
    "sample_label",
    "mu_cen",
    "mu_2n",
    "mu_4n",
    "mu_8n",
    "mu_4n_ratio_2"
  )])
  rbind(
    data.frame(sample_label = peak_line_df$sample_label, peak = "CEN", peak_location = peak_line_df$mu_cen),
    data.frame(sample_label = peak_line_df$sample_label, peak = "Diploid G0/G1", peak_location = peak_line_df$mu_2n),
    data.frame(sample_label = peak_line_df$sample_label, peak = "WGD+ G0/G1", peak_location = peak_line_df$mu_4n),
    data.frame(sample_label = peak_line_df$sample_label, peak = "WGD+ G2/M", peak_location = peak_line_df$mu_8n),
    data.frame(sample_label = peak_line_df$sample_label, peak = "WGD+ G0/G1 if ratio = 2", peak_location = peak_line_df$mu_4n_ratio_2)
  )
}

plot_overlay_subset <- function(
  subset_df,
  bin_width,
  plot_title,
  subtitle_path
) {
  peak_line_df <- build_peak_line_df(subset_df)
  peak_line_df$peak <- factor(
    peak_line_df$peak,
    levels = c("CEN", "Diploid G0/G1", "WGD+ G0/G1", "WGD+ G2/M", "WGD+ G0/G1 if ratio = 2")
  )

  ggplot(subset_df, aes(x = areaDNA_bin_center)) +
    geom_col(aes(y = count), width = bin_width, fill = "grey82", color = "grey55", linewidth = 0.225) +
    geom_line(aes(y = mixture_count), color = "#d94801", linewidth = 0.75) +
    geom_vline(
      data = peak_line_df,
      aes(xintercept = peak_location, color = peak),
      linetype = "dashed",
      linewidth = 0.525
    ) +
    scale_color_manual(
      name = "Peak guide",
      values = c(
        "CEN" = "#1b9e77",
        "Diploid G0/G1" = "#7570b3",
        "WGD+ G0/G1" = "#e7298a",
        "WGD+ G2/M" = "#66a61e",
        "WGD+ G0/G1 if ratio = 2" = "#e6ab02"
      ),
      guide = guide_legend(override.aes = list(linetype = "dashed", linewidth = 0.75))
    ) +
    facet_wrap(~ sample_label, scales = "free_y") +
    labs(
      title = plot_title,
      subtitle = subtitle_path,
      x = "Raw DNA area bin center",
      y = "Bin count",
      caption = "Hypothetical WGD+ G0/G1 line marks 2x the fitted diploid G0/G1 peak."
    ) +
    theme_bw(base_size = 13.5) +
    theme(
      strip.text = element_text(size = 10.5),
      panel.grid.minor = element_blank()
    )
}


plot_condition_overlay <- function(
  plot_df,
  condition_value,
  bin_width,
  title_prefix = "Fixed-init optimization checkpoint overlay",
  subtitle_path = optimizer_csv
) {
  condition_df <- plot_df[plot_df$condition_plot == condition_value & !plot_df$is_control, , drop = FALSE]
  if (nrow(condition_df) == 0) {
    return(NULL)
  }
  plot_overlay_subset(
    subset_df = condition_df,
    bin_width = bin_width,
    plot_title = sprintf("%s: %s", title_prefix, condition_value),
    subtitle_path = subtitle_path
  )
}

plot_controls_overlay <- function(
  plot_df,
  bin_width,
  title_prefix = "Fixed-init optimization checkpoint overlay",
  subtitle_path = optimizer_csv
) {
  controls_df <- plot_df[plot_df$is_control, , drop = FALSE]
  if (nrow(controls_df) == 0) {
    return(NULL)
  }
  plot_overlay_subset(
    subset_df = controls_df,
    bin_width = bin_width,
    plot_title = sprintf("%s: controls", title_prefix),
    subtitle_path = subtitle_path
  )
}

mixture_cdf <- function(x, mu_cen, mu_2n, mu_4n, mu_8n, cv, theta_row) {
  sig_cen <- mu_cen * cv
  sig_2n <- mu_2n * cv
  sig_4n <- mu_4n * cv
  sig_8n <- mu_8n * cv

  s12_mid <- 0.5 * (mu_2n + mu_4n)
  s12_sd <- 0.5 * (mu_4n - mu_2n)
  s48_mid <- 0.5 * (mu_4n + mu_8n)
  s48_sd <- 0.5 * (mu_8n - mu_4n)

  theta_row[1] * pnorm(x, mean = mu_cen, sd = sig_cen) +
    theta_row[2] * pnorm(x, mean = mu_2n, sd = sig_2n) +
    theta_row[3] * pnorm(x, mean = mu_4n, sd = sig_4n) +
    theta_row[4] * pnorm(x, mean = mu_8n, sd = sig_8n) +
    theta_row[5] * pnorm(x, mean = s12_mid, sd = s12_sd) +
    theta_row[6] * pnorm(x, mean = s48_mid, sd = s48_sd) +
    theta_row[7] * pnorm(x, mean = 2 * mu_cen, sd = sqrt(2) * sig_cen)
}

build_ecdf_df <- function(plot_df, theta, cv) {
  sample_ids <- sort(unique(plot_df$sampleID))
  ecdf_list <- vector("list", length(sample_ids))

  for (i in seq_along(sample_ids)) {
    sample_id <- sample_ids[i]
    sample_df <- plot_df[plot_df$sampleID == sample_id, , drop = FALSE]
    sample_df <- sample_df[order(sample_df$areaDNA_bin_center), ]
    cumulative_observed <- cumsum(sample_df$count) / sample_df$sample_total[1]

    predicted_cdf <- mixture_cdf(
      x = sample_df$areaDNA_bin_center,
      mu_cen = sample_df$mu_cen[1],
      mu_2n = sample_df$mu_2n[1],
      mu_4n = sample_df$mu_4n[1],
      mu_8n = sample_df$mu_8n[1],
      cv = cv,
      theta_row = theta[sample_id, ]
    )

    sample_df$observed_ecdf <- cumulative_observed
    sample_df$predicted_ecdf <- predicted_cdf
    ecdf_list[[i]] <- sample_df
  }

  do.call(rbind, ecdf_list)
}

plot_condition_ecdf <- function(
  ecdf_df,
  condition_value,
  title_prefix,
  subtitle_path
) {
  condition_df <- ecdf_df[ecdf_df$condition_plot == condition_value & !ecdf_df$is_control, , drop = FALSE]
  if (nrow(condition_df) == 0) {
    return(NULL)
  }
  peak_line_df <- build_peak_line_df(condition_df)

  ggplot(condition_df, aes(x = areaDNA_bin_center)) +
    geom_step(aes(y = observed_ecdf), color = "grey35", linewidth = 0.55) +
    geom_line(aes(y = predicted_ecdf), color = "#d94801", linewidth = 0.8) +
    geom_vline(
      data = peak_line_df,
      aes(xintercept = peak_location),
      linetype = "dashed",
      color = "grey55",
      linewidth = 0.45
    ) +
    facet_wrap(~ sample_label, scales = "free_x") +
    labs(
      title = sprintf("%s: %s", title_prefix, condition_value),
      subtitle = subtitle_path,
      x = "Raw DNA area bin center",
      y = "ECDF"
    ) +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw(base_size = 13.5) +
    theme(
      strip.text = element_text(size = 10.5),
      panel.grid.minor = element_blank()
    )
}

plot_controls_ecdf <- function(
  ecdf_df,
  title_prefix,
  subtitle_path
) {
  controls_df <- ecdf_df[ecdf_df$is_control, , drop = FALSE]
  if (nrow(controls_df) == 0) {
    return(NULL)
  }
  peak_line_df <- build_peak_line_df(controls_df)

  ggplot(controls_df, aes(x = areaDNA_bin_center)) +
    geom_step(aes(y = observed_ecdf), color = "grey35", linewidth = 0.55) +
    geom_line(aes(y = predicted_ecdf), color = "#d94801", linewidth = 0.8) +
    geom_vline(
      data = peak_line_df,
      aes(xintercept = peak_location),
      linetype = "dashed",
      color = "grey55",
      linewidth = 0.45
    ) +
    facet_wrap(~ sample_label, scales = "free_x") +
    labs(
      title = sprintf("%s: controls", title_prefix),
      subtitle = subtitle_path,
      x = "Raw DNA area bin center",
      y = "ECDF"
    ) +
    coord_cartesian(ylim = c(0, 1)) +
    theme_bw(base_size = 13.5) +
    theme(
      strip.text = element_text(size = 10.5),
      panel.grid.minor = element_blank()
    )
}


build_population_fraction_df <- function(theta, plot_df) {
  sample_fraction_df <- data.frame(
    sampleID = seq_len(nrow(theta)),
    CEN = rowSums(theta[, c(1, 7), drop = FALSE]),
    Diploid = rowSums(theta[, c(2, 5), drop = FALSE]),
    WGD = rowSums(theta[, c(3, 4, 6), drop = FALSE])
  )

  sample_annotations <- unique(plot_df[, c("sampleID", "sample_name", "sample_label", "condition_plot", "is_control")])
  sample_fraction_df <- merge(sample_fraction_df, sample_annotations, by = "sampleID", all.x = TRUE, sort = FALSE)

  long_values <- stack(sample_fraction_df[, c("CEN", "Diploid", "WGD")])
  fraction_df <- sample_fraction_df[rep(seq_len(nrow(sample_fraction_df)), 3), c(
    "sampleID",
    "sample_name",
    "sample_label",
    "condition_plot",
    "is_control"
  )]
  fraction_df$population <- factor(long_values$ind, levels = c("CEN", "Diploid", "WGD"))
  fraction_df$fraction <- long_values$values
  fraction_df
}

plot_population_fractions <- function(fraction_df) {
  sample_order_df <- unique(fraction_df[, c("sample_label", "condition_plot", "is_control")])
  sample_order_df <- sample_order_df[order(sample_order_df$is_control, sample_order_df$condition_plot, sample_order_df$sample_label), ]
  fraction_df$sample_label <- factor(fraction_df$sample_label, levels = sample_order_df$sample_label)

  ggplot(fraction_df, aes(x = sample_label, y = fraction, fill = population)) +
    geom_col(width = 0.82) +
    facet_wrap(~ condition_plot, scales = "free_x") +
    scale_fill_manual(values = c("CEN" = "#1b9e77", "Diploid" = "#7570b3", "WGD" = "#e7298a")) +
    scale_y_continuous(labels = function(x) sprintf("%d%%", round(100 * x))) +
    labs(
      title = "Dye-ratio fit: population fractions by sample",
      x = NULL,
      y = "Population fraction",
      fill = "Population"
    ) +
    theme_bw(base_size = 13.5) +
    theme(
      axis.text.x = element_blank(),
      panel.grid.minor = element_blank()
    )
}

build_dye_ratio_global_param_tbl <- function(opt_row, n_exp = 1L) {
  scalar_tbl <- data.frame(
    parameter = c(
      "lp__",
      "cv",
      "beta",
      "ratio_tumor_cen_geometric_mean",
      "R_4n_geometric_mean",
      "R_8n_geometric_mean"
    ),
    value = signif(c(
      opt_row$lp__[1],
      opt_row$cv[1],
      opt_row$beta[1],
      exp(opt_row$log_ratio_tumor_cen_mu[1]),
      exp(opt_row$log_R_4n_mu[1]),
      exp(opt_row$log_R_8n_mu[1])
    ), 6)
  )

  vector_specs <- list(
    mu_cen = n_exp,
    cen_phase = 2L,
    diploid_phase = 3L,
    wgd_phase = 3L
  )

  vector_tbl <- do.call(
    rbind,
    lapply(names(vector_specs), function(prefix) {
      n_vals <- vector_specs[[prefix]]
      values <- extract_indexed_vector(opt_row, prefix, n_vals)
      data.frame(
        parameter = sprintf("%s[%d]", prefix, seq_len(n_vals)),
        value = signif(values, 6)
      )
    })
  )

  rbind(scalar_tbl, vector_tbl)
}

derive_sample_group <- function(sample_name) {
  sample_id <- tools::file_path_sans_ext(basename(sample_name))

  if (identical(sample_id, "Sample_CEN")) {
    return("CEN")
  }

  if (grepl("^Sample_SUM159_", sample_id)) {
    return(sub("^Sample_SUM159_([0-9]+_)?", "", sample_id))
  }

  sample_id
}

augment_stan_bundle_with_groups <- function(stan_data, lut) {
  sample_groups <- vapply(lut$sample$sample_name, derive_sample_group, character(1))
  group_levels <- unique(sample_groups)
  group_lut <- data.frame(
    group_id = seq_along(group_levels),
    group_name = group_levels,
    stringsAsFactors = FALSE
  )

  lut$sample$sample_group <- sample_groups
  lut$sample$group_id <- match(sample_groups, group_lut$group_name)
  lut$group <- group_lut
  stan_data$N_group <- nrow(group_lut)
  stan_data$sampleGroupID <- as.integer(lut$sample$group_id)

  list(stan_data = stan_data, lut = lut)
}

exclude_samples_from_stan_bundle <- function(stan_data, lut, sample_patterns) {
  if (length(sample_patterns) == 0) {
    return(list(
      stan_data = stan_data,
      lut = lut,
      excluded_samples = lut$sample[FALSE, , drop = FALSE]
    ))
  }

  sample_name <- tools::file_path_sans_ext(basename(lut$sample$sample_name))
  exclude_samples <- Reduce(`|`, lapply(
    sample_patterns,
    function(pattern) grepl(pattern, sample_name, ignore.case = TRUE)
  ))

  if (!any(exclude_samples)) {
    return(list(
      stan_data = stan_data,
      lut = lut,
      excluded_samples = lut$sample[FALSE, , drop = FALSE]
    ))
  }

  keep_samples <- !exclude_samples
  old_sample_ids <- lut$sample$sample_id[keep_samples]
  keep_bins <- stan_data$sampleID %in% old_sample_ids
  sample_map <- setNames(seq_along(old_sample_ids), old_sample_ids)

  excluded_samples <- lut$sample[exclude_samples, , drop = FALSE]
  sample_lut <- lut$sample[keep_samples, , drop = FALSE]
  sample_lut$sample_id <- seq_len(nrow(sample_lut))

  stan_data$areaDNA_bin_center <- stan_data$areaDNA_bin_center[keep_bins]
  stan_data$count <- stan_data$count[keep_bins]
  stan_data$expID <- stan_data$expID[keep_bins]
  stan_data$sampleID <- unname(sample_map[as.character(stan_data$sampleID[keep_bins])])
  stan_data$sampleExpID <- sample_lut$experiment_id
  stan_data$N <- length(stan_data$count)
  stan_data$N_sample <- nrow(sample_lut)

  list(
    stan_data = stan_data,
    lut = list(
      experiment = lut$experiment,
      sample = sample_lut
    ),
    excluded_samples = excluded_samples
  )
}

build_overlay_df_observed_cen_scaled <- function(
  stan_data,
  lut,
  mu_cen,
  ratio_tumor_cen_sample,
  R_4n_sample,
  R_8n_sample,
  cv,
  theta,
  dye_scale_sample
) {
  x <- stan_data$areaDNA_bin_center
  bin_width <- diff(sort(unique(x)))[1]

  plot_df <- data.frame(
    areaDNA_bin_center = stan_data$areaDNA_bin_center,
    count = stan_data$count,
    expID = stan_data$expID,
    sampleID = stan_data$sampleID,
    stringsAsFactors = FALSE
  )

  sample_totals <- aggregate(count ~ sampleID, data = plot_df, FUN = sum)
  names(sample_totals)[2] <- "sample_total"
  plot_df <- merge(plot_df, sample_totals, by = "sampleID", sort = FALSE)

  plot_df$mixture_count <- mapply(
    FUN = function(x_i, sample_id, exp_id, sample_total) {
      mu_cen_sample <- mu_cen[exp_id] * dye_scale_sample[sample_id]
      sig_cen <- mu_cen_sample * cv
      mu_2n <- mu_cen_sample * ratio_tumor_cen_sample[sample_id]
      mu_4n <- mu_2n * R_4n_sample[sample_id]
      mu_8n <- mu_4n * R_8n_sample[sample_id]
      sig_2n <- mu_2n * cv
      sig_4n <- mu_4n * cv
      sig_8n <- mu_8n * cv

      density_val <- theta[sample_id, 1] * dnorm(x_i, mean = mu_cen_sample, sd = sig_cen) +
        theta[sample_id, 2] * dnorm(x_i, mean = mu_2n, sd = sig_2n) +
        theta[sample_id, 3] * dnorm(x_i, mean = mu_4n, sd = sig_4n) +
        theta[sample_id, 4] * dnorm(x_i, mean = mu_8n, sd = sig_8n) +
        theta[sample_id, 5] * s_phase_density(x_i, mu_2n, mu_4n, stan_data$N_s_phase) +
        theta[sample_id, 6] * s_phase_density(x_i, mu_4n, mu_8n, stan_data$N_s_phase) +
        theta[sample_id, 7] * dnorm(x_i, mean = 2 * mu_cen_sample, sd = sqrt(2) * sig_cen)

      sample_total * bin_width * density_val
    },
    x_i = plot_df$areaDNA_bin_center,
    sample_id = plot_df$sampleID,
    exp_id = plot_df$expID,
    sample_total = plot_df$sample_total
  )

  peak_df <- data.frame(
    sampleID = seq_len(nrow(lut$sample)),
    expID = lut$sample$experiment_id,
    stringsAsFactors = FALSE
  )
  peak_df$mu_cen <- mu_cen[peak_df$expID] * dye_scale_sample[peak_df$sampleID]
  peak_df$mu_2n <- peak_df$mu_cen * ratio_tumor_cen_sample[peak_df$sampleID]
  peak_df$mu_4n <- peak_df$mu_2n * R_4n_sample[peak_df$sampleID]
  peak_df$mu_8n <- peak_df$mu_4n * R_8n_sample[peak_df$sampleID]
  peak_df$mu_4n_ratio_2 <- 2 * peak_df$mu_2n

  plot_df <- merge(
    plot_df,
    lut$sample[, c("sample_id", "sample_name", "experiment_name", "sample_group")],
    by.x = "sampleID",
    by.y = "sample_id",
    all.x = TRUE,
    sort = FALSE
  )

  plot_df <- merge(
    plot_df,
    peak_df[, c("sampleID", "mu_cen", "mu_2n", "mu_4n", "mu_8n", "mu_4n_ratio_2")],
    by = "sampleID",
    all.x = TRUE,
    sort = FALSE
  )

  plot_df$sample_label <- sprintf("%s | %s", plot_df$experiment_name, plot_df$sample_name)
  plot_df
}

build_observed_cen_scaled_param_tbl <- function(opt_row, n_exp = 1L) {
  scalar_tbl <- data.frame(
    parameter = c(
      "lp__",
      "cv",
      "ratio_tumor_cen_geometric_mean",
      "R_4n_geometric_mean",
      "R_8n_geometric_mean",
      "D_tilde",
      "A_C_tilde",
      "A_T_tilde",
      "rho",
      "sigma_group_extra_load",
      "sigma_group_diploid_phase[1]",
      "sigma_group_diploid_phase[2]",
      "sigma_group_wgd_phase[1]",
      "sigma_group_wgd_phase[2]"
    ),
    value = signif(c(
      opt_row$lp__[1],
      opt_row$cv[1],
      exp(opt_row$log_ratio_tumor_cen_mu[1]),
      exp(opt_row$log_R_4n_mu[1]),
      exp(opt_row$log_R_8n_mu[1]),
      exp(opt_row$log_D_tilde[1]),
      exp(opt_row$log_A_C_tilde[1]),
      exp(opt_row$log_A_T_tilde[1]),
      exp(opt_row$log_rho[1]),
      opt_row$sigma_group_extra_load[1],
      opt_row$`sigma_group_diploid_phase.1`[1],
      opt_row$`sigma_group_diploid_phase.2`[1],
      opt_row$`sigma_group_wgd_phase.1`[1],
      opt_row$`sigma_group_wgd_phase.2`[1]
    ), 6)
  )

  vector_tbl <- do.call(
    rbind,
    lapply(list(mu_cen = n_exp), function(n_vals) NULL)
  )
  vector_tbl <- data.frame(
    parameter = sprintf("mu_cen[%d]", seq_len(n_exp)),
    value = signif(extract_indexed_vector(opt_row, "mu_cen", n_exp), 6)
  )

  rbind(scalar_tbl, vector_tbl)
}

softmax3 <- function(eta) {
  eta <- eta - max(eta)
  exp(eta) / sum(exp(eta))
}

extract_group_phase_tbl_observed_cen_scaled <- function(opt_row, lut) {
  out <- vector("list", nrow(lut$group))

  for (g in seq_len(nrow(lut$group))) {
    dip_eta <- c(
      0,
      opt_row$`diploid_phase_global_logit.1`[1] +
        opt_row$`sigma_group_diploid_phase.1`[1] * opt_row[[sprintf("z_group_diploid_phase.1.%d", g)]],
      opt_row$`diploid_phase_global_logit.2`[1] +
        opt_row$`sigma_group_diploid_phase.2`[1] * opt_row[[sprintf("z_group_diploid_phase.2.%d", g)]]
    )
    wgd_eta <- c(
      0,
      opt_row$`wgd_phase_global_logit.1`[1] +
        opt_row$`sigma_group_wgd_phase.1`[1] * opt_row[[sprintf("z_group_wgd_phase.1.%d", g)]],
      opt_row$`wgd_phase_global_logit.2`[1] +
        opt_row$`sigma_group_wgd_phase.2`[1] * opt_row[[sprintf("z_group_wgd_phase.2.%d", g)]]
    )

    dip <- softmax3(dip_eta)
    wgd <- softmax3(wgd_eta)

    out[[g]] <- data.frame(
      group_name = lut$group$group_name[g],
      state = c("Diploid", "Diploid", "Diploid", "WGD", "WGD", "WGD"),
      phase = rep(c("G0/G1", "S", "G2/M"), 2),
      fraction = c(dip, wgd),
      stringsAsFactors = FALSE
    )
  }

  do.call(rbind, out)
}

plot_group_phase_fractions_observed_cen_scaled <- function(group_phase_df) {
  ggplot(group_phase_df, aes(x = group_name, y = fraction, fill = phase)) +
    geom_col(width = 0.82) +
    facet_wrap(~ state, ncol = 1) +
    scale_y_continuous(labels = function(x) sprintf("%d%%", round(100 * x))) +
    labs(
      title = "Observed-CEN-scaled fit: estimated phase fractions by group",
      x = "Sample group",
      y = "Phase fraction",
      fill = "Phase"
    ) +
    theme_bw(base_size = 13.5) +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
}
```

```{r load-data, cache=FALSE}
prepared_first_pass <- prepare_fit_inputs(
  spec = first_pass_spec,
  project_root = project_root,
  stan_data_path = stan_data_path
)
stan_data <- prepared_first_pass$stan_data
lut <- prepared_first_pass$lut

opt_row <- read.csv(optimizer_csv, comment.char = "#", check.names = FALSE)[1, , drop = FALSE]
sample_metadata <- read.csv(sample_metadata_path, stringsAsFactors = FALSE)
```

# Checkpoint

```{r checkpoint-summary}
summary_tbl <- data.frame(
  field = c(
    "checkpoint_dir",
    "optimizer_csv",
    "sample_metadata",
    "experiment_filter",
    "N_exp",
    "N_sample",
    "N_bins",
    "lp__",
    "cv"
  ),
  value = c(
    checkpoint_dir,
    optimizer_csv,
    sample_metadata_path,
    experiment_filter,
    stan_data$N_exp,
    stan_data$N_sample,
    stan_data$N,
    signif(opt_row$lp__[1], 6),
    signif(opt_row$cv[1], 6)
  )
)

kable(summary_tbl)
```

# Histogram Overlay

We fit a simple model to the data where the locations of all the peaks (CEN/CEN doublet/2N/4N/8N) were globally fixed among all samples. The only sample level variable was the fraction of the population considered to be CEN/diploid/WGD. The CEN peak was fit very well, and the 2N peak was also fit decently (with the exception of the 2N control and 4N control samples, which we still don't have a good theory of why these appear so different from everything else). But we were intrigued by the observation that all flow samples for a given condition had their 2N peak shifted (relative to the estimate) in the same way. It appears there was a strong condition-specific batch effect.

```{r optimized-overlay-data, cache=TRUE}
overlay_df <- build_overlay_df(
  stan_data = stan_data,
  lut = lut,
  mu_cen = extract_indexed_vector(opt_row, "mu_cen", stan_data$N_exp),
  ratio_tumor_cen_sample = extract_indexed_vector(opt_row, "ratio_tumor_cen_sample", stan_data$N_sample),
  R_4n_sample = extract_indexed_vector(opt_row, "R_4n_sample", stan_data$N_sample),
  R_8n_sample = extract_indexed_vector(opt_row, "R_8n_sample", stan_data$N_sample),
  cv = opt_row$cv[1],
  theta = extract_theta_matrix(opt_row, stan_data$N_sample)
)
overlay_df <- add_sample_metadata(overlay_df, sample_metadata)

bin_width <- diff(sort(unique(overlay_df$areaDNA_bin_center)))[1]
condition_levels <- sort(unique(overlay_df$condition_plot))
```

```{r optimized-overlay, results='asis', cache=FALSE}
cat(render_or_include_plot(
  plot_path = file.path("fixed_init_anoxia_checkpoint_files", "manual", "optimized-overlay-controls.png"),
  plot_builder = function() plot_controls_overlay(overlay_df, bin_width),
  width = 16,
  height = 8,
  element_id = "optimized-overlay-controls"
))

for (condition_value in condition_levels) {
  cat(render_or_include_plot(
    plot_path = file.path(
      "fixed_init_anoxia_checkpoint_files",
      "manual",
      sprintf("optimized-overlay-%s.png", safe_plot_slug(condition_value))
    ),
    plot_builder = function() plot_condition_overlay(overlay_df, condition_value, bin_width),
    width = 16,
    height = 8,
    element_id = sprintf("optimized-overlay-%s", safe_plot_slug(condition_value))
  ))
}
```

# Dye-Ratio Fit

Our working theory for these experiments is as follows:

1) The CEN cells were prepared (fixed/stained) ONCE and then added to the (already fixed/stained) SUM-159 cells. This assumption is needed to explain why the CEN cells appear so consistent across all experiments.
2) A fixed quantity of CEN cells were added to each SUM-159 sample.
3) The ratio of DNA:nuclear dye is an important determinant of the location of the peaks. If there is insufficient dye to saturate the DNA (i.e. fully load all DNA binding sites with dye), then the location of the 2N/4N/8N peaks will move. 
4) We can use the CEN/Sum-159 ratio to determine the relative quantity of tumor cells per-sample, and therefore estimate the relative amount of DNA per sample (it actually occurs to me now that this only works if the SUM-159 cells all have the same amount of DNA per cell, which they clearly don't - this suggests an improvement to the model to-be-presented).
5) We introduce a model where the location of the non-CEN peaks is shifted proportional to the CEN/Sum-159 ratio.
6) The model results in a really good fit! With some caveats...

```{r dye-ratio-summary}
dye_ratio_opt_row <- read.csv(dye_ratio_optimizer_csv, comment.char = "#", check.names = FALSE)[1, , drop = FALSE]

dye_ratio_global_tbl <- build_dye_ratio_global_param_tbl(
  opt_row = dye_ratio_opt_row,
  n_exp = stan_data$N_exp
)

kable(dye_ratio_global_tbl, digits = 4)
```

```{r dye-ratio-overlay-data, cache=TRUE}
dye_ratio_overlay_df <- build_overlay_df(
  stan_data = stan_data,
  lut = lut,
  mu_cen = extract_indexed_vector(dye_ratio_opt_row, "mu_cen", stan_data$N_exp),
  ratio_tumor_cen_sample = extract_indexed_vector(dye_ratio_opt_row, "ratio_tumor_cen_sample", stan_data$N_sample),
  R_4n_sample = extract_indexed_vector(dye_ratio_opt_row, "R_4n_sample", stan_data$N_sample),
  R_8n_sample = extract_indexed_vector(dye_ratio_opt_row, "R_8n_sample", stan_data$N_sample),
  cv = dye_ratio_opt_row$cv[1],
  theta = extract_theta_matrix(dye_ratio_opt_row, stan_data$N_sample)
)
dye_ratio_overlay_df <- add_sample_metadata(dye_ratio_overlay_df, sample_metadata)
dye_ratio_theta <- extract_theta_matrix(dye_ratio_opt_row, stan_data$N_sample)
```

```{r dye-ratio-overlay, results='asis', cache=FALSE}
cat(render_or_include_plot(
  plot_path = file.path("fixed_init_anoxia_checkpoint_files", "manual", "dye-ratio-overlay-controls.png"),
  plot_builder = function() plot_controls_overlay(
    dye_ratio_overlay_df,
    bin_width,
    title_prefix = "Dye-ratio fixed-init optimization overlay",
    subtitle_path = dye_ratio_optimizer_csv
  ),
  width = 16,
  height = 8,
  element_id = "dye-ratio-overlay-controls"
))

for (condition_value in condition_levels) {
  cat(render_or_include_plot(
    plot_path = file.path(
      "fixed_init_anoxia_checkpoint_files",
      "manual",
      sprintf("dye-ratio-overlay-%s.png", safe_plot_slug(condition_value))
    ),
    plot_builder = function() plot_condition_overlay(
      dye_ratio_overlay_df,
      condition_value,
      bin_width,
      title_prefix = "Dye-ratio fixed-init optimization overlay",
      subtitle_path = dye_ratio_optimizer_csv
    ),
    width = 16,
    height = 8,
    element_id = sprintf("dye-ratio-overlay-%s", safe_plot_slug(condition_value))
  ))
}
```

# Population Fractions
These broadly make sense. CEN is all CEN. Everything else is broadly diploid, except the anoxia only sample (which we already know is WGDish). THe other exceptions are the 2N and 4N controls, which... we're just ignoring for now.

```{r dye-ratio-population-fractions-data, cache=TRUE}
population_fraction_df <- build_population_fraction_df(
  theta = dye_ratio_theta,
  plot_df = dye_ratio_overlay_df
)
```

```{r dye-ratio-population-fractions, fig.width=18, fig.height=9, cache=FALSE}
print(plot_population_fractions(population_fraction_df))
```

# ECDF Diagnostics

These diagnostics are very revealing. They show that the model predicted CEN/tumor ratio is way off in the 4N control condition, and a bit off in the anoxia condition (where WGD occurred). Basically what's happening is that in order to nail the peak locations, the model is deliberately assigning an incorrect CEN/tumor ratio. So either our theory isn't correct (or at least isn't quite the whole story), or the model for how CEN/tumor ratio impacts peak location is misspecified. 

One other thing to note is that we just don't get the 8N peak at all, so that's certainly something to look at.

```{r dye-ratio-ecdf-data, cache=TRUE}
dye_ratio_ecdf_df <- build_ecdf_df(
  plot_df = dye_ratio_overlay_df,
  theta = dye_ratio_theta,
  cv = dye_ratio_opt_row$cv[1]
)
```

```{r dye-ratio-ecdf, fig.width=16, fig.height=8, cache=FALSE}
print(plot_controls_ecdf(
  dye_ratio_ecdf_df,
  title_prefix = "Dye-ratio fit ECDF comparison",
  subtitle_path = dye_ratio_optimizer_csv
))

for (condition_value in condition_levels) {
  print(plot_condition_ecdf(
    dye_ratio_ecdf_df,
    condition_value,
    title_prefix = "Dye-ratio fit ECDF comparison",
    subtitle_path = dye_ratio_optimizer_csv
  ))
}
```

# Observed-CEN-Scaled Fluorescence Model

This keeps the original dye-ratio section intact as a reference point, but adds
the newer fluorescence derivation. In this model, the observed CEN fraction is
used to scale the tumor DNA load proxy, the fluorescence compression is derived
from the reduced mechanistic free-dye model, and the diploid/WGD G0/G1/S/G2/M
ratios are allowed small group-level deviations around a global baseline.

```{r observed-cen-scaled-load}
observed_cen_scaled_manifest <- readRDS(observed_cen_scaled_manifest_path)

observed_cen_scaled_bundle <- prepare_fit_inputs(
  spec = observed_cen_scaled_spec,
  project_root = project_root,
  stan_data_path = stan_data_path
)
observed_cen_scaled_stan_data <- observed_cen_scaled_bundle$stan_data
observed_cen_scaled_lut <- observed_cen_scaled_bundle$lut
observed_cen_scaled_opt_row <- read.csv(
  observed_cen_scaled_optimizer_csv,
  comment.char = "#",
  check.names = FALSE
)[1, , drop = FALSE]
```

```{r observed-cen-scaled-summary}
observed_cen_scaled_param_tbl <- build_observed_cen_scaled_param_tbl(
  opt_row = observed_cen_scaled_opt_row,
  n_exp = observed_cen_scaled_stan_data$N_exp
)

kable(observed_cen_scaled_param_tbl, digits = 4)
```

```{r observed-cen-scaled-overlay-data, cache=TRUE}
observed_cen_scaled_overlay_df <- build_overlay_df_observed_cen_scaled(
  stan_data = observed_cen_scaled_stan_data,
  lut = observed_cen_scaled_lut,
  mu_cen = extract_indexed_vector(observed_cen_scaled_opt_row, "mu_cen", observed_cen_scaled_stan_data$N_exp),
  ratio_tumor_cen_sample = extract_indexed_vector(observed_cen_scaled_opt_row, "ratio_tumor_cen_sample", observed_cen_scaled_stan_data$N_sample),
  R_4n_sample = extract_indexed_vector(observed_cen_scaled_opt_row, "R_4n_sample", observed_cen_scaled_stan_data$N_sample),
  R_8n_sample = extract_indexed_vector(observed_cen_scaled_opt_row, "R_8n_sample", observed_cen_scaled_stan_data$N_sample),
  cv = observed_cen_scaled_opt_row$cv[1],
  theta = extract_theta_matrix(observed_cen_scaled_opt_row, observed_cen_scaled_stan_data$N_sample),
  dye_scale_sample = extract_indexed_vector(observed_cen_scaled_opt_row, "dye_scale_sample", observed_cen_scaled_stan_data$N_sample)
)
observed_cen_scaled_overlay_df <- add_sample_metadata(observed_cen_scaled_overlay_df, sample_metadata)
observed_cen_scaled_theta <- extract_theta_matrix(observed_cen_scaled_opt_row, observed_cen_scaled_stan_data$N_sample)
observed_cen_scaled_bin_width <- diff(sort(unique(observed_cen_scaled_overlay_df$areaDNA_bin_center)))[1]
```

```{r observed-cen-scaled-overlay, results='asis', cache=FALSE}
cat(render_or_include_plot(
  plot_path = file.path("fixed_init_anoxia_checkpoint_files", "manual", "observed-cen-scaled-overlay-controls.png"),
  plot_builder = function() plot_controls_overlay(
    observed_cen_scaled_overlay_df,
    observed_cen_scaled_bin_width,
    title_prefix = "Observed-CEN-scaled optimization overlay",
    subtitle_path = observed_cen_scaled_manifest$output_file
  ),
  width = 16,
  height = 8,
  element_id = "observed-cen-scaled-overlay-controls"
))

for (condition_value in condition_levels) {
  cat(render_or_include_plot(
    plot_path = file.path(
      "fixed_init_anoxia_checkpoint_files",
      "manual",
      sprintf("observed-cen-scaled-overlay-%s.png", safe_plot_slug(condition_value))
    ),
    plot_builder = function() plot_condition_overlay(
      observed_cen_scaled_overlay_df,
      condition_value,
      observed_cen_scaled_bin_width,
      title_prefix = "Observed-CEN-scaled optimization overlay",
      subtitle_path = observed_cen_scaled_manifest$output_file
    ),
    width = 16,
    height = 8,
    element_id = sprintf("observed-cen-scaled-overlay-%s", safe_plot_slug(condition_value))
  ))
}
```

```{r observed-cen-scaled-population-fractions-data, cache=TRUE}
observed_cen_scaled_population_fraction_df <- build_population_fraction_df(
  theta = observed_cen_scaled_theta,
  plot_df = observed_cen_scaled_overlay_df
)
```

```{r observed-cen-scaled-population-fractions, fig.width=18, fig.height=9, cache=FALSE}
print(
  plot_population_fractions(observed_cen_scaled_population_fraction_df) +
    labs(title = "Observed-CEN-scaled fit: population fractions by sample")
)
```

```{r observed-cen-scaled-ecdf-data, cache=TRUE}
observed_cen_scaled_ecdf_df <- build_ecdf_df(
  plot_df = observed_cen_scaled_overlay_df,
  theta = observed_cen_scaled_theta,
  cv = observed_cen_scaled_opt_row$cv[1]
)
```

```{r observed-cen-scaled-ecdf, fig.width=16, fig.height=8, cache=FALSE}
print(plot_controls_ecdf(
  observed_cen_scaled_ecdf_df,
  title_prefix = "Observed-CEN-scaled fit ECDF comparison",
  subtitle_path = observed_cen_scaled_manifest$output_file
))

for (condition_value in condition_levels) {
  print(plot_condition_ecdf(
    observed_cen_scaled_ecdf_df,
    condition_value,
    title_prefix = "Observed-CEN-scaled fit ECDF comparison",
    subtitle_path = observed_cen_scaled_manifest$output_file
  ))
}
```

```{r observed-cen-scaled-group-phase-data, cache=TRUE}
observed_cen_scaled_group_phase_df <- extract_group_phase_tbl_observed_cen_scaled(
  opt_row = observed_cen_scaled_opt_row,
  lut = observed_cen_scaled_lut
)
```

```{r observed-cen-scaled-group-phase, fig.width=14, fig.height=10, cache=FALSE}
print(plot_group_phase_fractions_observed_cen_scaled(observed_cen_scaled_group_phase_df))
```

# Model Comparison

The original dye-ratio fit above is still useful as a baseline because it shows
what the simpler peak-shift story was doing. The newer observed-CEN-scaled
model changes two things at once: it uses the revised fluorescence derivation,
and it allows slight group-level deviations in the diploid/WGD phase ratios.

The practical comparison to focus on is:

1. Whether the overlay and ECDF diagnostics improve in the conditions where the
   older dye-ratio model had to distort the inferred CEN/tumor ratio.
2. Whether the fitted group-level phase deviations remain small, which would
   support the intended “mostly global, slightly group-specific” interpretation.
