---
title: "Peak Detection Across Project Datasets"
knit: !expr function(input_file, encoding) rmarkdown::render(input_file, encoding = encoding, output_dir = dirname(input_file), intermediates_dir = dirname(input_file), envir = new.env(parent = globalenv()))
output:
  html_document:
    toc: true
    toc_float: true
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
---

```{r setup, include=FALSE}
input_path <- knitr::current_input(dir = TRUE)
workflow_dir <- if (length(input_path) == 0) {
  normalizePath(getwd(), winslash = "/", mustWork = TRUE)
} else {
  normalizePath(dirname(input_path), winslash = "/", mustWork = TRUE)
}
project_root <- normalizePath(file.path(workflow_dir, ".."), winslash = "/", mustWork = TRUE)
if (!dir.exists(file.path(project_root, "data")) && dir.exists(file.path(getwd(), "data"))) {
  project_root <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
}
knitr::opts_knit$set(root.dir = workflow_dir)
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

# Goal

This report applies the promoted density-based peak detection workflow to each
project dataset under `data/`, using the gated DNA vectors exported under
`processed_data/`. For every dataset it detects one below-threshold reference
peak, up to two above-threshold peaks, and flags the primary outlier sample
using the same two-peak ratio rule that was developed in `dev/`.

```{r libraries}
library(ggplot2)
library(knitr)
library(dplyr)
library(tidyr)
library(tibble)
library(scales)

source(file.path(project_root, "R", "peak_detection.R"))
```

```{r config}
data_root <- file.path(project_root, "data")
processed_root <- file.path(project_root, "processed_data")

threshold <- 12000
tumor_peak_min_separation <- 35000
tumor_peak_min_prominence <- 0
max_hist_signal <- 200000

sanitize_dataset_id <- function(x) {
  x <- tolower(gsub("[^A-Za-z0-9]+", "-", x))
  x <- gsub("(^-+|-+$)", "", x)
  if (!nzchar(x)) {
    stop("Could not derive a dataset id from the supplied name.")
  }
  x
}

discover_dataset_dirs <- function(root_path) {
  dirs <- list.dirs(root_path, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[!startsWith(basename(dirs), ".")]
  dirs <- dirs[basename(dirs) != "cache"]
  dirs <- dirs[basename(dirs) != "__MACOSX"]
  sort(normalizePath(dirs, winslash = "/", mustWork = TRUE))
}

load_dataset_peak_bundle <- function(dataset_dir) {
  dataset_name <- basename(dataset_dir)
  dataset_id <- sanitize_dataset_id(dataset_name)
  cache_dir <- file.path(processed_root, dataset_id)
  dna_path <- file.path(cache_dir, "filtered_dna_area_vectors.rds")
  metadata_path <- file.path(cache_dir, "sample_metadata.csv")

  if (!file.exists(dna_path) || !file.exists(metadata_path)) {
    cache_parts <- build_peak_detection_cache(
      dataset_dir = dataset_dir,
      cache_dir = cache_dir
    )
    dna_vectors <- cache_parts$dna_vectors
    metadata_tbl <- cache_parts$metadata_tbl
    cache_source <- "bootstrap_from_raw_fcs"
  } else {
    dna_vectors <- readRDS(dna_path)
    metadata_tbl <- read.csv(metadata_path, stringsAsFactors = FALSE, check.names = FALSE)
    cache_source <- "existing_processed_data"
  }

  peak_tbl <- detect_dataset_peaks(
    dna_vectors = dna_vectors$raw,
    sample_metadata = metadata_tbl,
    dataset_id = dataset_id,
    dataset_name = dataset_name,
    threshold = threshold,
    tumor_peak_min_separation = tumor_peak_min_separation,
    tumor_peak_min_prominence = tumor_peak_min_prominence
  )

  summary_path <- file.path(cache_dir, "peak_detection_summary.csv")
  write.csv(peak_tbl, summary_path, row.names = FALSE)

  plot_parts <- build_peak_plot_data(
    peak_tbl = peak_tbl,
    dna_vectors = dna_vectors$raw,
    max_signal = max_hist_signal
  )

  list(
    dataset_name = dataset_name,
    dataset_id = dataset_id,
    cache_dir = cache_dir,
    summary_path = summary_path,
    cache_source = cache_source,
    peak_tbl = peak_tbl,
    overview_tbl = build_peak_detection_overview(peak_tbl),
    hist_tbl = plot_parts$hist_tbl,
    peak_line_tbl = plot_parts$peak_line_tbl
  )
}

dataset_dirs <- discover_dataset_dirs(data_root)
dataset_results <- lapply(dataset_dirs, load_dataset_peak_bundle)
names(dataset_results) <- vapply(dataset_results, `[[`, character(1), "dataset_id")

overview_tbl <- bind_rows(lapply(dataset_results, `[[`, "overview_tbl"))
summary_manifest_tbl <- tibble(
  dataset_name = vapply(dataset_results, `[[`, character(1), "dataset_name"),
  dataset_id = vapply(dataset_results, `[[`, character(1), "dataset_id"),
  cache_source = vapply(dataset_results, `[[`, character(1), "cache_source"),
  summary_csv = vapply(dataset_results, `[[`, character(1), "summary_path")
)
```

# Dataset Overview

```{r dataset-overview}
kable(overview_tbl %>%
  mutate(
    median_ratio_above_below = round(median_ratio_above_below, 3),
    median_count_ratio_above_below = round(median_count_ratio_above_below, 3)
  ))
```

# Saved Outputs

```{r saved-outputs}
kable(summary_manifest_tbl)
```

# Per-Dataset Results

```{r dataset-sections, results='asis', fig.width=12, fig.height=18}
plot_dataset_histograms <- function(res) {
  if (!nrow(res$hist_tbl)) {
    return(
      ggplot() +
        annotate("text", x = 1, y = 1, label = "No histogram data available.") +
        theme_void()
    )
  }

  ggplot(res$hist_tbl, aes(x = bin_center, y = count)) +
    geom_col(width = diff(range(res$hist_tbl$bin_center, na.rm = TRUE)) / 256, fill = "#9ecae1", color = NA) +
    geom_vline(
      data = res$peak_line_tbl,
      aes(xintercept = peak_location, color = outlier_status, linetype = peak_type),
      linewidth = 0.35,
      alpha = 0.9,
      show.legend = TRUE
    ) +
    facet_wrap(~ sample_id, scales = "free_y", ncol = 3) +
    scale_color_manual(values = c("Primary outlier" = "#cb181d", "Not primary outlier" = "#2171b5")) +
    scale_linetype_manual(values = c("CEN peak" = "dotted", "Tumor peak 1" = "solid", "Tumor peak 2" = "longdash")) +
    labs(
      title = paste(res$dataset_name, "raw DNA histograms with peak annotations"),
      subtitle = "Colored lines show the dev-methodology peak calls and the primary outlier flag.",
      x = "Raw DNA area",
      y = "Event count",
      color = NULL,
      linetype = NULL
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(size = 7),
      legend.position = "bottom"
    )
}

plot_peak_locations <- function(res) {
  plot_tbl <- bind_rows(
    res$peak_tbl %>%
      transmute(
        sample_id,
        peak_type = "CEN peak",
        peak_location = cen_peak,
        is_primary_outlier
      ),
    res$peak_tbl %>%
      transmute(
        sample_id,
        peak_type = "Tumor peak 1",
        peak_location = tumor_peak_1,
        is_primary_outlier
      ),
    res$peak_tbl %>%
      transmute(
        sample_id,
        peak_type = "Tumor peak 2",
        peak_location = tumor_peak_2,
        is_primary_outlier
      )
  ) %>%
    filter(is.finite(peak_location))

  if (!nrow(plot_tbl)) {
    return(
      ggplot() +
        annotate("text", x = 1, y = 1, label = "No peak locations are available.") +
        theme_void()
    )
  }

  sample_levels <- sort(unique(res$peak_tbl$sample_id))

  ggplot(
    plot_tbl,
    aes(
      x = factor(sample_id, levels = sample_levels),
      y = peak_location,
      color = peak_type,
      shape = is_primary_outlier
    )
  ) +
    geom_point(size = 2.3, alpha = 0.9) +
    coord_flip() +
    scale_color_manual(
      values = c(
        "CEN peak" = "#1b9e77",
        "Tumor peak 1" = "#d95f02",
        "Tumor peak 2" = "#7570b3"
      )
    ) +
    scale_shape_manual(
      values = c("TRUE" = 17, "FALSE" = 16),
      labels = c("FALSE" = "Not primary outlier", "TRUE" = "Primary outlier")
    ) +
    labs(
      title = paste(res$dataset_name, "peak locations by sample"),
      x = NULL,
      y = "Peak location",
      color = NULL,
      shape = NULL
    ) +
    theme_bw() +
    theme(legend.position = "bottom")
}

for (res in dataset_results) {
  cat(sprintf("## %s\n\n", res$dataset_name))
  cat(sprintf("Peak input source: `%s`  \n", res$cache_source))
  cat(sprintf("Cached summary: `%s`\n\n", gsub("\\\\", "/", res$summary_path)))
  cat(sprintf(
    "Samples: %d. Two-peak calls: %d. Primary outliers: %d.\n\n",
    nrow(res$peak_tbl),
    sum(is.finite(res$peak_tbl$tumor_peak_1) & is.finite(res$peak_tbl$tumor_peak_2)),
    sum(res$peak_tbl$is_primary_outlier, na.rm = TRUE)
  ))
  cat("\n")
  print(plot_dataset_histograms(res))
  cat("\n")
  print(plot_peak_locations(res))
  cat("\n\n")
}
```
