---
title: "Ploidy-Oxygen Workflow: Data Processing and Initial Modelling"
output:
  html_document:
    toc: true
    toc_float: true
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "./")
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

# Workflow Goal

This workflow is an assay-characterization and structure-recovery exercise, not
yet a downstream biological hypothesis test. The goal is to learn as much as
possible from the flow data in isolation while keeping the interpretation
conservative.

The central questions at this stage are:

1. How does this assay behave across samples and runs?
2. How much variation appears technical rather than biological?
3. Which automated QC, gating, and peak-detection strategies are defensible?
4. What population structure is robustly recoverable from the flow DNA signal
   alone?
5. Where does the flow data stop supporting an unambiguous interpretation?

The workflow is anchored by three assumptions:

1. The internal chicken reference is expected to be the same biological
   population in every sample, so its between-sample variation is treated as
   primarily technical unless the data argue otherwise.
2. The 2N and 4N control lineages provide an additional layer of expected
   structure and serve as an important sanity check on recoverable peak
   architecture.
3. O1 and O2 samples may contain real ploidy shifts, so they are analyzed more
   cautiously and are not imposed as rigid reference structure.

# Input Configuration

```{r libraries}
library(flowCore)
library(ggplot2)
library(knitr)
library(scales)
```

```{r input-config}
path <- "data/Hypoxia_SUM159"
stopifnot(dir.exists(path))

dna_channel <- "X450.Violet.C.A"
scatter_channels <- c("FSC.A", "SSC.A", "FSC.H")
time_channel <- "Time"
saturation_cap <- 262143

ff <- list.files(path, pattern = "\\.fcs$", full.names = FALSE)
stopifnot(length(ff) > 0)
```

# Sample Metadata Parsing

```{r sample-metadata}
parse_sample_name <- function(filename) {
  core <- sub("^Sample_", "", filename)
  core <- sub("\\.fcs$", "", core)
  parts <- strsplit(core, "_", fixed = TRUE)[[1]]

  replicate_raw <- parts[4]
  replicate_id <- sub("_+$", "", replicate_raw)

  data.frame(
    file_name = filename,
    sample_id = core,
    cell_line = parts[1],
    starting_ploidy = parts[2],
    oxygen_condition = parts[3],
    replicate_id = replicate_id,
    stringsAsFactors = FALSE
  )
}

sample_meta <- do.call(rbind, lapply(ff, parse_sample_name))
sample_meta$starting_ploidy <- factor(sample_meta$starting_ploidy, levels = c("2N", "4N"))
sample_meta$oxygen_condition <- factor(sample_meta$oxygen_condition, levels = c("C", "O1", "O2"))

kable(sample_meta, caption = "Parsed sample metadata from FCS filenames.")
```

# Analysis Strategy

This workflow uses a deliberately conservative sequence:

1. Import all events and apply minimal hard exclusions only for non-finite,
   non-positive, and near-saturated events.
2. Apply a singlet-oriented gate using the robust relationship between `FSC-A`
   and `FSC-H` instead of trimming the dominant scatter cloud.
3. Remove severe time-instability segments when event counts or median DNA
   signal drift sharply across acquisition.
4. Detect DNA peaks on the log-intensity scale using smoothed density estimates,
   local maxima, prominence filtering, minimum mass filtering, and minimum
   separation.
5. Detect low-DNA candidate peaks with a second, more permissive pass so the
   chicken reference is not discarded simply because it is smaller and less
   dominant than the tumor compartment.
6. Build a consensus chicken-reference expectation from control samples and use
   that expectation to classify or withhold per-sample chicken assignments.
7. Summarize what can be recovered robustly, and explicitly label unresolved or
   ambiguous samples rather than forcing a chicken + tumor G0/G1 interpretation.

# Helper Functions

```{r helper-functions}
read_flow_table <- function(file_name) {
  fcs <- read.FCS(file.path(path, file_name))
  dfi <- as.data.frame(exprs(fcs))
  colnames(dfi) <- make.names(colnames(dfi))
  dfi
}

robust_scale <- function(x) {
  x <- x[is.finite(x)]
  center <- median(x, na.rm = TRUE)
  scale <- mad(x, constant = 1.4826, na.rm = TRUE)

  if (!is.finite(scale) || scale <= 0) {
    scale <- IQR(x, na.rm = TRUE) / 1.349
  }
  if (!is.finite(scale) || scale <= 0) {
    scale <- sd(x, na.rm = TRUE)
  }
  if (!is.finite(scale) || scale <= 0) {
    scale <- 1
  }

  c(center = center, scale = scale)
}

estimate_debris_gate <- function(qc_df) {
  if (nrow(qc_df) < 100) {
    return(list(
      data = qc_df,
      metrics = data.frame(
        events_after_debris = nrow(qc_df),
        retention_debris = 1,
        debris_cutoff = NA_real_,
        debris_gate_status = "not_attempted",
        debris_reference_peak = NA_real_,
        debris_reference_width = NA_real_,
        low_scatter_fraction = NA_real_,
        stringsAsFactors = FALSE
      )
    ))
  }

  log_fsc <- log10(qc_df$FSC.A)
  log_ssc <- log10(qc_df$SSC.A)
  fsc_stats <- robust_scale(log_fsc)
  ssc_stats <- robust_scale(log_ssc)

  scatter_index <- 0.5 * (
    (log_fsc - fsc_stats["center"]) / fsc_stats["scale"] +
      (log_ssc - ssc_stats["center"]) / ssc_stats["scale"]
  )
  low_scatter <- scatter_index <= median(scatter_index, na.rm = TRUE)
  low_scatter_fraction <- mean(low_scatter, na.rm = TRUE)

  low_scatter_peaks <- find_density_peaks(
    qc_df[[dna_channel]][low_scatter],
    adjust = 1.8,
    min_prom_frac = 0.03,
    min_area = 0.03,
    min_sep_log10 = 0.06
  )

  if (nrow(low_scatter_peaks) == 0) {
    return(list(
      data = qc_df,
      metrics = data.frame(
        events_after_debris = nrow(qc_df),
        retention_debris = 1,
        debris_cutoff = NA_real_,
        debris_gate_status = "unresolved",
        debris_reference_peak = NA_real_,
        debris_reference_width = NA_real_,
        low_scatter_fraction = low_scatter_fraction,
        stringsAsFactors = FALSE
      )
    ))
  }

  first_peak <- low_scatter_peaks[1, , drop = FALSE]
  debris_cutoff <- first_peak$left_value[1]
  keep_debris <- qc_df[[dna_channel]] >= debris_cutoff
  debris_df <- qc_df[keep_debris, , drop = FALSE]

  list(
    data = debris_df,
    metrics = data.frame(
      events_after_debris = nrow(debris_df),
      retention_debris = nrow(debris_df) / max(nrow(qc_df), 1),
      debris_cutoff = debris_cutoff,
      debris_gate_status = "resolved",
      debris_reference_peak = first_peak$peak_value[1],
      debris_reference_width = log10(first_peak$right_value[1]) - log10(first_peak$left_value[1]),
      low_scatter_fraction = low_scatter_fraction,
      stringsAsFactors = FALSE
    )
  )
}

apply_qc_filters <- function(dfi) {
  required_channels <- c(dna_channel, scatter_channels, time_channel)

  keep_basic <- complete.cases(dfi[, required_channels]) &
    dfi[[dna_channel]] > 0 &
    dfi$FSC.A > 0 &
    dfi$FSC.H > 0 &
    dfi$SSC.A > 0

  sat_threshold <- saturation_cap * 0.995
  keep_basic <- keep_basic &
    dfi[[dna_channel]] < sat_threshold &
    dfi$FSC.A < sat_threshold &
    dfi$FSC.H < sat_threshold &
    dfi$SSC.A < sat_threshold

  qc_df <- dfi[keep_basic, , drop = FALSE]
  if (nrow(qc_df) < 100) {
    return(list(
      data = qc_df,
      metrics = data.frame(
        events_raw = nrow(dfi),
        events_after_basic = nrow(qc_df),
        events_after_singlets = nrow(qc_df),
        events_after_time = nrow(qc_df),
        events_after_debris = nrow(qc_df),
        retention_basic = nrow(qc_df) / nrow(dfi),
        retention_singlets = 1,
        retention_time = 1,
        retention_debris = 1,
        debris_cutoff = NA_real_,
        debris_gate_status = "not_attempted",
        debris_reference_peak = NA_real_,
        debris_reference_width = NA_real_,
        low_scatter_fraction = NA_real_,
        stringsAsFactors = FALSE
      )
    ))
  }

  singlet_fit <- lm(log10(FSC.H) ~ log10(FSC.A), data = qc_df)
  singlet_resid <- resid(singlet_fit)
  singlet_stats <- robust_scale(singlet_resid)
  keep_singlets <- abs((singlet_resid - singlet_stats["center"]) / singlet_stats["scale"]) <= 3.5

  qc_df <- qc_df[keep_singlets, , drop = FALSE]
  if (nrow(qc_df) < 100) {
    return(list(
      data = qc_df,
      metrics = data.frame(
        events_raw = nrow(dfi),
        events_after_basic = sum(keep_basic),
        events_after_singlets = nrow(qc_df),
        events_after_time = nrow(qc_df),
        events_after_debris = nrow(qc_df),
        retention_basic = sum(keep_basic) / nrow(dfi),
        retention_singlets = nrow(qc_df) / max(sum(keep_basic), 1),
        retention_time = 1,
        retention_debris = 1,
        debris_cutoff = NA_real_,
        debris_gate_status = "not_attempted",
        debris_reference_peak = NA_real_,
        debris_reference_width = NA_real_,
        low_scatter_fraction = NA_real_,
        stringsAsFactors = FALSE
      )
    ))
  }

  time_breaks <- unique(quantile(qc_df[[time_channel]], probs = seq(0, 1, length.out = 21), na.rm = TRUE))
  keep_time <- rep(TRUE, nrow(qc_df))

  if (length(time_breaks) > 2) {
    time_bin <- cut(qc_df[[time_channel]], breaks = time_breaks, include.lowest = TRUE)
    bin_counts <- tapply(qc_df[[dna_channel]], time_bin, length)
    bin_medians <- tapply(log10(qc_df[[dna_channel]]), time_bin, median, na.rm = TRUE)

    count_stats <- robust_scale(log10(as.numeric(bin_counts)))
    median_stats <- robust_scale(as.numeric(bin_medians))

    good_bins <- abs((log10(as.numeric(bin_counts)) - count_stats["center"]) / count_stats["scale"]) <= 3 &
      abs((as.numeric(bin_medians) - median_stats["center"]) / median_stats["scale"]) <= 3

    keep_time <- good_bins[as.integer(time_bin)]
    keep_time[is.na(keep_time)] <- FALSE
  }

  qc_time_df <- qc_df[keep_time, , drop = FALSE]
  debris_result <- estimate_debris_gate(qc_time_df)

  list(
    data = debris_result$data,
    metrics = cbind(
      data.frame(
        events_raw = nrow(dfi),
        events_after_basic = sum(keep_basic),
        events_after_singlets = nrow(qc_df),
        events_after_time = nrow(qc_time_df),
        retention_basic = sum(keep_basic) / nrow(dfi),
        retention_singlets = nrow(qc_df) / max(sum(keep_basic), 1),
        retention_time = nrow(qc_time_df) / max(nrow(qc_df), 1),
        stringsAsFactors = FALSE
      ),
      debris_result$metrics
    )
  )
}

find_density_peaks <- function(x, adjust = 1.4, min_prom_frac = 0.06, min_area = 0.01, min_sep_log10 = 0.08) {
  x <- x[is.finite(x) & x > 0]
  lx <- log10(x)

  if (length(unique(round(lx, 4))) < 20) {
    return(data.frame())
  }

  dens <- density(lx, n = 4096, adjust = adjust)
  peak_idx <- which(diff(sign(diff(dens$y))) == -2) + 1
  valley_idx <- which(diff(sign(diff(dens$y))) == 2) + 1

  if (length(peak_idx) == 0) {
    return(data.frame())
  }

  peaks <- lapply(peak_idx, function(idx) {
    left_valley <- if (any(valley_idx < idx)) tail(valley_idx[valley_idx < idx], 1) else 1
    right_valley <- if (any(valley_idx > idx)) head(valley_idx[valley_idx > idx], 1) else length(dens$x)
    baseline <- max(dens$y[left_valley], dens$y[right_valley])

    data.frame(
      peak_log10 = dens$x[idx],
      peak_value = 10^dens$x[idx],
      density_height = dens$y[idx],
      prominence = dens$y[idx] - baseline,
      left_value = 10^dens$x[left_valley],
      right_value = 10^dens$x[right_valley],
      area_fraction = mean(x >= 10^dens$x[left_valley] & x <= 10^dens$x[right_valley]),
      stringsAsFactors = FALSE
    )
  })

  peaks <- do.call(rbind, peaks)
  peaks <- peaks[
    peaks$prominence >= max(dens$y) * min_prom_frac &
      peaks$area_fraction >= min_area,
    ,
    drop = FALSE
  ]

  if (nrow(peaks) == 0) {
    return(peaks)
  }

  peaks <- peaks[order(peaks$density_height, decreasing = TRUE), , drop = FALSE]
  keep <- logical(nrow(peaks))
  chosen <- numeric(0)

  for (i in seq_len(nrow(peaks))) {
    if (length(chosen) == 0 || all(abs(peaks$peak_log10[i] - chosen) >= min_sep_log10)) {
      keep[i] <- TRUE
      chosen <- c(chosen, peaks$peak_log10[i])
    }
  }

  peaks <- peaks[keep, , drop = FALSE]
  peaks[order(peaks$peak_value), , drop = FALSE]
}

annotate_peaks <- function(qc_df, peaks, pass_label) {
  if (nrow(peaks) == 0) {
    peaks$pass <- character(0)
    return(peaks)
  }

  overall_logfsc <- log10(qc_df$FSC.A)
  overall_center <- median(overall_logfsc, na.rm = TRUE)
  overall_iqr <- IQR(overall_logfsc, na.rm = TRUE)
  if (!is.finite(overall_iqr) || overall_iqr <= 0) {
    overall_iqr <- 1
  }

  peaks$median_logFSC <- NA_real_
  peaks$median_logSSC <- NA_real_
  peaks$scatter_shift <- NA_real_
  peaks$log_width <- log10(peaks$right_value) - log10(peaks$left_value)
  peaks$local_cv <- NA_real_
  peaks$peak_quality <- NA_real_
  peaks$pass <- pass_label

  for (i in seq_len(nrow(peaks))) {
    in_window <- qc_df[[dna_channel]] >= peaks$left_value[i] & qc_df[[dna_channel]] <= peaks$right_value[i]
    sub_df <- qc_df[in_window, , drop = FALSE]
    peaks$median_logFSC[i] <- median(log10(sub_df$FSC.A), na.rm = TRUE)
    peaks$median_logSSC[i] <- median(log10(sub_df$SSC.A), na.rm = TRUE)
    peaks$scatter_shift[i] <- (peaks$median_logFSC[i] - overall_center) / overall_iqr
    peaks$local_cv[i] <- sd(sub_df[[dna_channel]], na.rm = TRUE) / mean(sub_df[[dna_channel]], na.rm = TRUE)
    peaks$peak_quality[i] <- log1p(peaks$prominence[i]) - peaks$log_width[i] - peaks$local_cv[i]
  }

  peaks
}

merge_peak_sets <- function(strict_peaks, permissive_peaks) {
  if (nrow(strict_peaks) == 0 && nrow(permissive_peaks) == 0) {
    return(data.frame())
  }

  all_peaks <- rbind(strict_peaks, permissive_peaks)
  all_peaks <- all_peaks[order(all_peaks$peak_value, -all_peaks$prominence), , drop = FALSE]

  keep <- rep(TRUE, nrow(all_peaks))
  for (i in seq_len(nrow(all_peaks))) {
    if (!keep[i]) {
      next
    }
    close_idx <- which(
      seq_len(nrow(all_peaks)) > i &
        abs(all_peaks$peak_log10 - all_peaks$peak_log10[i]) < 0.03
    )
    keep[close_idx] <- FALSE
  }

  all_peaks[keep, , drop = FALSE]
}

analyze_sample <- function(file_name) {
  raw_df <- read_flow_table(file_name)
  qc_result <- apply_qc_filters(raw_df)
  qc_df <- qc_result$data
  qc_metrics <- qc_result$metrics

  strict_peaks <- annotate_peaks(
    qc_df,
    find_density_peaks(qc_df[[dna_channel]], adjust = 1.4, min_prom_frac = 0.06, min_area = 0.01, min_sep_log10 = 0.08),
    pass_label = "strict"
  )

  permissive_peaks <- annotate_peaks(
    qc_df,
    find_density_peaks(qc_df[[dna_channel]], adjust = 1.8, min_prom_frac = 0.03, min_area = 0.03, min_sep_log10 = 0.06),
    pass_label = "permissive"
  )

  all_peaks <- merge_peak_sets(strict_peaks, permissive_peaks)
  dominant_peak <- if (nrow(strict_peaks) > 0) strict_peaks$peak_value[which.max(strict_peaks$area_fraction * strict_peaks$prominence)] else NA_real_

  all_peaks$peak_rank <- rank(all_peaks$peak_value, ties.method = "first")

  sample_summary <- cbind(
    data.frame(file_name = file_name, stringsAsFactors = FALSE),
    qc_metrics,
    data.frame(
      peak_count_strict = nrow(strict_peaks),
      peak_count_total = nrow(all_peaks),
      dominant_peak = dominant_peak,
      stringsAsFactors = FALSE
    )
  )

  list(
    raw = raw_df,
    qc = qc_df,
    peaks = all_peaks,
    strict_peaks = strict_peaks,
    summary = sample_summary
  )
}

fit_reference_template <- function(reference_tbl, feature_cols) {
  center <- setNames(numeric(length(feature_cols)), feature_cols)
  scale <- center

  for (feature in feature_cols) {
    stats <- robust_scale(reference_tbl[[feature]])
    center[feature] <- stats["center"]
    scale[feature] <- stats["scale"]
    if (!is.finite(scale[feature]) || scale[feature] <= 0) {
      scale[feature] <- 1
    }
  }

  list(center = center, scale = scale, feature_cols = feature_cols)
}

score_reference_candidates <- function(peak_tbl, template) {
  if (nrow(peak_tbl) == 0) {
    peak_tbl$reference_distance <- numeric(0)
    peak_tbl$reference_support <- numeric(0)
    return(peak_tbl)
  }

  feature_cols <- template$feature_cols
  peak_tbl$reference_distance <- apply(
    peak_tbl[, feature_cols, drop = FALSE],
    1,
    function(x) sqrt(sum(((x - template$center[feature_cols]) / template$scale[feature_cols])^2))
  )

  scaled_distance <- peak_tbl$reference_distance - min(peak_tbl$reference_distance, na.rm = TRUE)
  raw_support <- exp(-0.5 * scaled_distance^2)
  support_total <- sum(raw_support)
  if (!is.finite(support_total) || support_total <= 0) {
    peak_tbl$reference_support <- rep(1 / nrow(peak_tbl), nrow(peak_tbl))
  } else {
    peak_tbl$reference_support <- raw_support / support_total
  }
  peak_tbl
}

classify_chicken_reference <- function(sample_peak_tbl, sample_meta) {
  feature_cols <- c("peak_log10", "log_width", "local_cv", "scatter_shift", "peak_quality")

  control_peak_tbl <- merge(
    sample_peak_tbl,
    subset(sample_meta, oxygen_condition == "C")[, c("file_name", "starting_ploidy")],
    by = "file_name"
  )

  control_candidates <- lapply(
    split(control_peak_tbl, control_peak_tbl$file_name),
    function(d) {
      d <- d[order(d$peak_value), , drop = FALSE]
      if (nrow(d) > 1) {
        d <- d[d$peak_rank < max(d$peak_rank), , drop = FALSE]
      }
      head(d, min(2, nrow(d)))
    }
  )

  control_grid <- expand.grid(
    lapply(control_candidates, function(d) seq_len(nrow(d))),
    KEEP.OUT.ATTRS = FALSE,
    stringsAsFactors = FALSE
  )

  evaluate_control_combo <- function(selection_row) {
    chosen <- do.call(
      rbind,
      Map(
        function(candidate_tbl, idx) candidate_tbl[idx, , drop = FALSE],
        control_candidates,
        as.integer(selection_row)
      )
    )

    control_major_tbl <- do.call(
      rbind,
      lapply(split(control_peak_tbl, control_peak_tbl$file_name), function(d) {
        selected_peak <- chosen$peak_value[chosen$file_name == d$file_name[1]]
        non_reference <- d[abs(d$peak_log10 - log10(selected_peak)) >= 0.08, , drop = FALSE]
        if (nrow(non_reference) == 0) {
          return(data.frame(file_name = d$file_name[1], major_peak = NA_real_, stringsAsFactors = FALSE))
        }
        best_non_reference <- non_reference[which.max(non_reference$area_fraction * non_reference$prominence), , drop = FALSE]
        data.frame(file_name = d$file_name[1], major_peak = best_non_reference$peak_value, stringsAsFactors = FALSE)
      })
    )

    chosen <- merge(chosen, control_major_tbl, by = "file_name", all.x = TRUE)
    chosen$major_to_reference_ratio <- chosen$major_peak / chosen$peak_value

    position_stability <- robust_scale(chosen$peak_log10)["scale"]
    shape_stability <- sum(vapply(c("log_width", "local_cv", "scatter_shift"), function(feature) {
      robust_scale(chosen[[feature]])["scale"]
    }, numeric(1)))
    coherence_2n <- robust_scale(log(subset(chosen, starting_ploidy == "2N")$major_to_reference_ratio))["scale"]
    coherence_4n <- robust_scale(log(subset(chosen, starting_ploidy == "4N")$major_to_reference_ratio))["scale"]
    lineage_separation <- median(log(subset(chosen, starting_ploidy == "4N")$major_to_reference_ratio), na.rm = TRUE) -
      median(log(subset(chosen, starting_ploidy == "2N")$major_to_reference_ratio), na.rm = TRUE)
    median_quality <- median(chosen$peak_quality, na.rm = TRUE)

    data.frame(
      chosen,
      position_stability = position_stability,
      shape_stability = shape_stability,
      coherence_2n = coherence_2n,
      coherence_4n = coherence_4n,
      lineage_separation = lineage_separation,
      median_quality = median_quality,
      stringsAsFactors = FALSE
    )
  }

  combo_tbl <- do.call(
    rbind,
    lapply(seq_len(nrow(control_grid)), function(i) {
      chosen <- evaluate_control_combo(control_grid[i, , drop = FALSE])
      data.frame(
        combo_id = i,
        position_stability = chosen$position_stability[1],
        shape_stability = chosen$shape_stability[1],
        coherence_2n = chosen$coherence_2n[1],
        coherence_4n = chosen$coherence_4n[1],
        lineage_separation = chosen$lineage_separation[1],
        median_quality = chosen$median_quality[1],
        stringsAsFactors = FALSE
      )
    })
  )

  combo_tbl$rank_position <- rank(combo_tbl$position_stability, ties.method = "min")
  combo_tbl$rank_shape <- rank(combo_tbl$shape_stability, ties.method = "min")
  combo_tbl$rank_coherence_2n <- rank(combo_tbl$coherence_2n, ties.method = "min")
  combo_tbl$rank_coherence_4n <- rank(combo_tbl$coherence_4n, ties.method = "min")
  combo_tbl$rank_separation <- rank(-combo_tbl$lineage_separation, ties.method = "min")
  combo_tbl$rank_quality <- rank(-combo_tbl$median_quality, ties.method = "min")
  combo_tbl$combo_rank_sum <- rowSums(combo_tbl[, grepl("^rank_", names(combo_tbl)), drop = FALSE])

  best_combo_id <- combo_tbl$combo_id[which.min(combo_tbl$combo_rank_sum)]
  best_control_tbl <- evaluate_control_combo(control_grid[best_combo_id, , drop = FALSE])

  template <- fit_reference_template(best_control_tbl, feature_cols)
  scored_peaks <- score_reference_candidates(sample_peak_tbl, template)
  control_scored_tbl <- score_reference_candidates(best_control_tbl, template)
  distance_limit <- max(control_scored_tbl$reference_distance, na.rm = TRUE)
  support_limit <- min(control_scored_tbl$reference_support, na.rm = TRUE)

  assigned_tbl <- do.call(
    rbind,
    lapply(split(scored_peaks, scored_peaks$file_name), function(d) {
      d <- d[order(d$reference_distance, -d$peak_quality, d$peak_value), , drop = FALSE]
      chosen <- d[1, , drop = FALSE]
      chosen$chicken_status <- if (chosen$reference_distance <= distance_limit && chosen$reference_support >= support_limit) {
        "assigned"
      } else {
        "ambiguous"
      }
      chosen
    })
  )

  list(
    assigned = assigned_tbl,
    scored_peaks = scored_peaks,
    selected_controls = best_control_tbl,
    control_scored_tbl = control_scored_tbl,
    control_combo_tbl = combo_tbl,
    template = template,
    distance_limit = distance_limit,
    support_limit = support_limit
  )
}

classify_major_structure <- function(sample_peak_tbl, chicken_tbl) {
  out <- merge(sample_peak_tbl, chicken_tbl[, c("file_name", "peak_value", "chicken_status")], by = "file_name", all.x = TRUE, suffixes = c("", "_chicken"))
  names(out)[names(out) == "peak_value_chicken"] <- "assigned_chicken_peak"

  out$non_chicken_peak <- with(out, is.na(assigned_chicken_peak) | abs(log10(peak_value) - log10(assigned_chicken_peak)) >= 0.08)
  out$major_population_candidate <- out$non_chicken_peak & out$pass == "strict"

  major_tbl <- do.call(
    rbind,
    lapply(split(out, out$file_name), function(d) {
      d <- subset(d, major_population_candidate)
      if (nrow(d) == 0) {
        return(data.frame())
      }
      d[which.max(d$area_fraction * d$prominence), , drop = FALSE]
    })
  )

  major_tbl
}
```

# Import, QC, and Peak Detection

```{r sample-analysis}
sample_analysis <- setNames(lapply(ff, analyze_sample), ff)

flow_raw <- lapply(sample_analysis, `[[`, "raw")
flow_qc <- lapply(sample_analysis, `[[`, "qc")
sample_summary <- do.call(rbind, lapply(sample_analysis, `[[`, "summary"))
sample_peak_tbl <- do.call(
  rbind,
  lapply(names(sample_analysis), function(file_name) {
    peak_df <- sample_analysis[[file_name]]$peaks
    if (nrow(peak_df) == 0) {
      return(data.frame(
        file_name = character(0),
        peak_log10 = numeric(0),
        peak_value = numeric(0),
        density_height = numeric(0),
        prominence = numeric(0),
        left_value = numeric(0),
        right_value = numeric(0),
        area_fraction = numeric(0),
        median_logFSC = numeric(0),
        median_logSSC = numeric(0),
        scatter_shift = numeric(0),
        log_width = numeric(0),
        local_cv = numeric(0),
        peak_quality = numeric(0),
        pass = character(0),
        peak_rank = numeric(0),
        stringsAsFactors = FALSE
      ))
    }
    peak_df$file_name <- file_name
    peak_df
  })
)

chicken_result <- classify_chicken_reference(sample_peak_tbl, sample_meta)
chicken_tbl <- chicken_result$assigned
major_tbl <- classify_major_structure(sample_peak_tbl, chicken_tbl)

analysis_tbl <- merge(sample_meta, sample_summary, by = "file_name", sort = FALSE, all.x = TRUE)
analysis_tbl <- merge(
  analysis_tbl,
  chicken_tbl[, c("file_name", "peak_value", "prominence", "area_fraction", "scatter_shift", "peak_quality", "reference_distance", "reference_support", "pass", "chicken_status")],
  by = "file_name",
  all.x = TRUE
)
names(analysis_tbl)[names(analysis_tbl) == "peak_value"] <- "chicken_peak"
names(analysis_tbl)[names(analysis_tbl) == "prominence"] <- "chicken_prominence"
names(analysis_tbl)[names(analysis_tbl) == "area_fraction"] <- "chicken_area_fraction"
names(analysis_tbl)[names(analysis_tbl) == "scatter_shift"] <- "chicken_scatter_shift"
names(analysis_tbl)[names(analysis_tbl) == "peak_quality"] <- "chicken_peak_quality"
names(analysis_tbl)[names(analysis_tbl) == "reference_distance"] <- "chicken_reference_distance"
names(analysis_tbl)[names(analysis_tbl) == "reference_support"] <- "chicken_reference_support"
names(analysis_tbl)[names(analysis_tbl) == "pass"] <- "chicken_detection_pass"

analysis_tbl <- merge(
  analysis_tbl,
  major_tbl[, c("file_name", "peak_value", "prominence", "area_fraction", "scatter_shift")],
  by = "file_name",
  all.x = TRUE
)
names(analysis_tbl)[names(analysis_tbl) == "peak_value"] <- "major_peak"
names(analysis_tbl)[names(analysis_tbl) == "prominence"] <- "major_peak_prominence"
names(analysis_tbl)[names(analysis_tbl) == "area_fraction"] <- "major_peak_area_fraction"
names(analysis_tbl)[names(analysis_tbl) == "scatter_shift"] <- "major_peak_scatter_shift"

analysis_tbl$major_to_chicken_ratio <- ifelse(
  analysis_tbl$chicken_status == "assigned" & is.finite(analysis_tbl$major_peak),
  analysis_tbl$major_peak / analysis_tbl$chicken_peak,
  NA_real_
)
analysis_tbl$qc_flag_low_events <- analysis_tbl$events_after_debris < 2000
analysis_tbl$qc_flag_low_retention <- with(analysis_tbl, retention_basic * retention_singlets * retention_time * retention_debris < 0.35)
analysis_tbl$qc_flag_debris_gate_unresolved <- analysis_tbl$debris_gate_status != "resolved"
analysis_tbl$qc_flag_no_chicken <- !is.finite(analysis_tbl$chicken_peak)
analysis_tbl$qc_flag_chicken_ambiguous <- is.na(analysis_tbl$chicken_status) | analysis_tbl$chicken_status != "assigned"
analysis_tbl$qc_flag_no_major_peak <- !is.finite(analysis_tbl$major_peak)
analysis_tbl$qc_flag_structure_unresolved <- analysis_tbl$qc_flag_no_major_peak | analysis_tbl$qc_flag_no_chicken

analysis_tbl <- analysis_tbl[
  order(analysis_tbl$starting_ploidy, analysis_tbl$oxygen_condition, analysis_tbl$replicate_id),
]

kable(
  analysis_tbl[, c(
    "sample_id", "starting_ploidy", "oxygen_condition",
    "events_raw", "events_after_debris", "peak_count_total",
    "chicken_peak", "chicken_status", "major_peak",
    "major_to_chicken_ratio", "qc_flag_low_events",
    "qc_flag_low_retention", "qc_flag_debris_gate_unresolved", "qc_flag_structure_unresolved"
  )],
  digits = 3,
  caption = "Per-sample QC and peak summary after conservative filtering, two-stage peak detection, and explicit chicken-reference classification."
)
```

# Assay and QC Summary

```{r qc-overview-table}
qc_overview <- analysis_tbl[, c(
  "sample_id", "starting_ploidy", "oxygen_condition",
  "events_raw", "events_after_basic", "events_after_singlets", "events_after_time", "events_after_debris",
  "retention_basic", "retention_singlets", "retention_time", "retention_debris",
  "debris_gate_status", "debris_cutoff",
  "qc_flag_low_events", "qc_flag_low_retention", "qc_flag_debris_gate_unresolved"
)]

kable(
  qc_overview,
  digits = 3,
  caption = "QC retention across the staged filtering workflow. The intent is to remove clear acquisition artifacts while preserving potentially real biological structure."
)
```

```{r qc-example-plot}
example_file <- analysis_tbl$file_name[1]
example_raw <- flow_raw[[example_file]]
example_qc <- flow_qc[[example_file]]

ggplot(example_raw, aes(x = FSC.A, y = SSC.A)) +
  geom_point(alpha = 0.05, size = 0.25, color = "#9c6644") +
  geom_point(
    data = example_qc,
    aes(x = FSC.A, y = SSC.A),
    alpha = 0.05,
    size = 0.25,
    color = "#386641"
  ) +
  labs(
    title = paste("QC overlay for", example_file),
    subtitle = "Raw events in brown, retained events in green; the workflow avoids a narrow tumor-centered scatter gate",
    x = "FSC-A",
    y = "SSC-A"
  ) +
  theme_minimal(base_size = 12)
```

```{r qc-time-summary}
time_qc_tbl <- do.call(
  rbind,
  lapply(names(flow_qc), function(file_name) {
    df <- flow_qc[[file_name]]
    data.frame(
      file_name = file_name,
      sample_id = sample_meta$sample_id[sample_meta$file_name == file_name],
      median_dna = median(df[[dna_channel]], na.rm = TRUE),
      iqr_dna = IQR(df[[dna_channel]], na.rm = TRUE),
      median_fsc = median(df$FSC.A, na.rm = TRUE),
      median_ssc = median(df$SSC.A, na.rm = TRUE),
      stringsAsFactors = FALSE
    )
  })
)

kable(
  time_qc_tbl,
  digits = 2,
  caption = "Assay-level summaries after QC. Between-sample shifts here are descriptive and should not automatically be interpreted biologically."
)
```

# Chicken Reference Assessment

The chicken-reference strategy now treats the chicken compartment as a
cross-sample calibration object rather than just a low-DNA peak. Following the
literature already summarized in [`docs/literature.txt`](C:\Users\4473331\Documents\projects\023_ploidyFlow\docs\literature.txt),
the workflow does three things differently:

1. It adds an explicit debris gate by locating the left boundary of the first
   resolved low-scatter nuclear peak and excluding events below that boundary.
2. It uses the control samples to define a multifeature reference template
   based on DNA position, peak width, local CV, scatter profile, and a
   peak-quality term that rewards sharp, prominent peaks.
3. It assigns the chicken peak in each sample only if the best candidate
   behaves like the control-derived calibration object under leave-one-out
   control validation.

This remains conservative: if the best candidate does not look sufficiently
reference-like, the sample is left ambiguous.

```{r chicken-consensus-text, results='asis'}
cat(
  sprintf(
      paste(
        "The selected control-reference model sets a control-calibrated acceptance envelope of distance <= %.3f.",
        "Support is reported for ranking within each sample, but assignment is driven primarily by the control-derived distance envelope.",
        "Candidates outside that envelope are left ambiguous."
      ),
      chicken_result$distance_limit,
      chicken_result$support_limit
  )
)
```

```{r control-reference-table}
control_reference_tbl <- chicken_result$selected_controls[, c(
  "file_name", "starting_ploidy", "peak_value", "major_peak",
  "major_to_reference_ratio", "log_width", "local_cv",
  "scatter_shift", "peak_quality"
)]

kable(
  control_reference_tbl,
  digits = 3,
  caption = "Control-sample peaks selected as the reference template. These control assignments are used to define the allowable reference-like feature envelope."
)
```

```{r chicken-reference-plot}
chicken_plot_tbl <- merge(
  chicken_result$scored_peaks,
  sample_meta[, c("file_name", "sample_id", "starting_ploidy", "oxygen_condition")],
  by = "file_name"
)

chicken_plot_tbl$peak_role <- "Other peak"
if (nrow(chicken_tbl) > 0) {
  chosen_keys <- paste(chicken_tbl$file_name, signif(chicken_tbl$peak_value, 8))
  plot_keys <- paste(chicken_plot_tbl$file_name, signif(chicken_plot_tbl$peak_value, 8))
  chicken_plot_tbl$peak_role[plot_keys %in% chosen_keys] <- "Assigned chicken"
}

ggplot(
  chicken_plot_tbl,
  aes(x = reference_distance, y = peak_quality, color = peak_role, shape = oxygen_condition)
) +
  geom_vline(xintercept = chicken_result$distance_limit, linetype = "dotted", color = "grey45") +
  geom_point(aes(size = area_fraction), alpha = 0.85) +
  scale_color_manual(
    values = c(
      "Other peak" = "#6c757d",
      "Assigned chicken" = "#d62828"
    )
  ) +
  labs(
    title = "Chicken-reference candidates across samples",
    subtitle = "Candidates are judged by similarity to the control-derived reference template rather than by a fixed low-DNA rule",
    x = "Reference-template distance",
    y = "Peak quality score",
    color = "Peak class",
    shape = "Oxygen condition",
    size = "Peak mass"
  ) +
  theme_minimal(base_size = 12)
```

```{r chicken-summary-table}
chicken_summary_tbl <- analysis_tbl[, c(
  "sample_id", "starting_ploidy", "oxygen_condition",
  "chicken_peak", "chicken_status", "chicken_detection_pass",
  "chicken_area_fraction", "chicken_prominence",
  "chicken_scatter_shift", "chicken_peak_quality",
  "chicken_reference_distance", "chicken_reference_support"
)]

kable(
  chicken_summary_tbl,
  digits = 3,
  caption = "Chicken-reference assignments. Samples marked ambiguous or unresolved are intentionally not forced into a reference interpretation."
)
```

# Recoverable Population Structure

The strict peak call represents structure that is reasonably well supported by
the DNA density after QC. The permissive pass is used primarily to avoid missing
the chicken reference. Interpretation of the major non-chicken peak remains
descriptive unless the sample also has a credible chicken assignment.

```{r example-density}
example_peaks <- subset(sample_peak_tbl, file_name == example_file)

ggplot(example_qc, aes(x = .data[[dna_channel]])) +
  geom_density(fill = "#8ecae6", alpha = 0.5, color = "#023047", linewidth = 0.8, adjust = 1.4) +
  geom_vline(
    data = example_peaks,
    aes(xintercept = peak_value, color = pass),
    linewidth = 0.7,
    linetype = "dashed"
  ) +
  scale_color_manual(values = c("strict" = "#d00000", "permissive" = "#6a4c93")) +
  labs(
    title = paste("DNA density and called peaks for", example_file),
    subtitle = "Strict peaks summarize robust structure; permissive peaks are retained mainly to rescue low-DNA reference structure",
    x = "450-Violet C-A",
    y = "Density",
    color = "Peak pass"
  ) +
  theme_minimal(base_size = 12)
```

```{r all-sample-density-grid, fig.height=14, fig.width=16}
sample_order <- analysis_tbl$file_name

density_tbl <- do.call(
  rbind,
  lapply(sample_order, function(file_name) {
    dens <- density(log10(flow_qc[[file_name]][[dna_channel]]), n = 2048, adjust = 1.4)
    data.frame(
      file_name = file_name,
      dna_value = 10^dens$x,
      density_value = dens$y,
      stringsAsFactors = FALSE
    )
  })
)

peak_grid_tbl <- sample_peak_tbl
peak_grid_tbl$file_name <- factor(peak_grid_tbl$file_name, levels = sample_order)
density_tbl$file_name <- factor(density_tbl$file_name, levels = sample_order)

facet_labels <- setNames(analysis_tbl$sample_id, analysis_tbl$file_name)

ggplot(density_tbl, aes(x = dna_value, y = density_value)) +
  geom_area(fill = "#8ecae6", alpha = 0.5) +
  geom_line(color = "#023047", linewidth = 0.5) +
  geom_vline(
    data = peak_grid_tbl,
    aes(xintercept = peak_value, color = pass),
    linetype = "dashed",
    linewidth = 0.4
  ) +
  scale_x_log10(labels = label_number()) +
  scale_color_manual(values = c("strict" = "#d00000", "permissive" = "#6a4c93")) +
  facet_wrap(
    ~file_name,
    ncol = 4,
    nrow = 5,
    scales = "free_y",
    labeller = as_labeller(facet_labels)
  ) +
  labs(
    title = "QC-filtered DNA density for every flow sample",
    subtitle = "The plot shows recoverable density structure rather than forcing a fixed two-peak interpretation",
    x = "450-Violet C-A",
    y = "Density",
    color = "Peak pass"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    strip.text = element_text(size = 8),
    panel.grid.minor = element_blank()
  )
```

```{r structure-summary-plot}
plot_tbl <- rbind(
  data.frame(
    sample_id = analysis_tbl$sample_id,
    starting_ploidy = analysis_tbl$starting_ploidy,
    oxygen_condition = analysis_tbl$oxygen_condition,
    peak_type = "Assigned chicken",
    peak_value = analysis_tbl$chicken_peak,
    status = analysis_tbl$chicken_status,
    stringsAsFactors = FALSE
  ),
  data.frame(
    sample_id = analysis_tbl$sample_id,
    starting_ploidy = analysis_tbl$starting_ploidy,
    oxygen_condition = analysis_tbl$oxygen_condition,
    peak_type = "Major non-chicken peak",
    peak_value = analysis_tbl$major_peak,
    status = ifelse(is.finite(analysis_tbl$major_peak), "detected", "unresolved"),
    stringsAsFactors = FALSE
  )
)

ggplot(plot_tbl, aes(x = oxygen_condition, y = peak_value, fill = peak_type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_point(
    aes(shape = status),
    position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.75),
    size = 1.8,
    alpha = 0.9
  ) +
  scale_y_log10(labels = label_number()) +
  scale_fill_manual(values = c("Assigned chicken" = "#f4a261", "Major non-chicken peak" = "#2a9d8f")) +
  facet_wrap(~starting_ploidy, nrow = 1) +
  labs(
    title = "Recovered peak positions by control lineage and oxygen condition",
    subtitle = "These are descriptive assay summaries, not yet a downstream biological model",
    x = "Oxygen condition",
    y = "Peak position on DNA channel",
    fill = "Peak type",
    shape = "Status"
  ) +
  theme_minimal(base_size = 12)
```

# Technical Variability and Control Sanity Checks

```{r technical-variability-table}
control_tbl <- subset(analysis_tbl, oxygen_condition == "C" & chicken_status == "assigned")

technical_variability_tbl <- data.frame(
  metric = c(
    "Chicken peak CV among C samples",
    "Major peak CV among C samples",
    "2N control ratio median",
    "4N control ratio median",
    "Max control-template distance",
    "Minimum support reported"
  ),
  value = c(
    sd(control_tbl$chicken_peak, na.rm = TRUE) / mean(control_tbl$chicken_peak, na.rm = TRUE),
    sd(control_tbl$major_peak, na.rm = TRUE) / mean(control_tbl$major_peak, na.rm = TRUE),
    median(subset(control_tbl, starting_ploidy == "2N")$major_to_chicken_ratio, na.rm = TRUE),
    median(subset(control_tbl, starting_ploidy == "4N")$major_to_chicken_ratio, na.rm = TRUE),
    chicken_result$distance_limit,
    chicken_result$support_limit
  ),
  stringsAsFactors = FALSE
)

kable(
  technical_variability_tbl,
  digits = 3,
  caption = "Simple technical summaries from the control condition. These values are intended as assay diagnostics, not final calibration constants."
)
```

```{r control-sanity-plot}
control_plot_tbl <- subset(analysis_tbl, oxygen_condition == "C" & is.finite(major_to_chicken_ratio))

ggplot(control_plot_tbl, aes(x = starting_ploidy, y = major_to_chicken_ratio, color = starting_ploidy)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.2) +
  geom_point(position = position_jitter(width = 0.06, height = 0), size = 2.5) +
  scale_color_manual(values = c("2N" = "#1d3557", "4N" = "#d62828")) +
  labs(
    title = "Control-lineage sanity check",
    subtitle = "The 2N and 4N lineages should show distinguishable normalized structure if the reference and major peak assignments are credible",
    x = "Starting ploidy lineage",
    y = "Major peak / chicken peak",
    color = "Starting ploidy"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

# Ambiguity and Failure Modes

This workflow does not assume every sample contains a clean chicken peak plus a
single tumor G0/G1 peak. The main ambiguity modes observed here are:

1. The low-DNA compartment can be present but weak, broad, or poorly separated
   from debris or acquisition artifacts.
2. Some samples show strong high-DNA structure without an equally convincing
   chicken assignment.
3. In O1 and O2 samples, the major non-chicken peak may shift or split in ways
   that should not be normalized away without a defensible reference call.

```{r ambiguity-table}
ambiguity_tbl <- subset(
  analysis_tbl,
  qc_flag_low_events |
    qc_flag_low_retention |
    qc_flag_debris_gate_unresolved |
    qc_flag_no_chicken |
    qc_flag_chicken_ambiguous |
    qc_flag_no_major_peak
)[, c(
  "sample_id", "starting_ploidy", "oxygen_condition",
  "qc_flag_low_events", "qc_flag_low_retention",
  "qc_flag_debris_gate_unresolved",
  "qc_flag_no_chicken", "qc_flag_chicken_ambiguous",
  "qc_flag_no_major_peak"
)]

kable(
  ambiguity_tbl,
  caption = "Samples carrying at least one unresolved or ambiguous QC/interpretation flag."
)
```

# Interpretation at This Stage

```{r interpretation-text, results='asis'}
assigned_chicken_n <- sum(analysis_tbl$chicken_status == "assigned", na.rm = TRUE)
ambiguous_chicken_n <- sum(analysis_tbl$qc_flag_chicken_ambiguous, na.rm = TRUE)
resolved_major_n <- sum(is.finite(analysis_tbl$major_peak), na.rm = TRUE)

cat(
  sprintf(
    paste(
      "Across %s samples, the workflow assigns a chicken reference in %s samples,",
      "marks %s samples as ambiguous or unresolved for the chicken reference,",
      "and recovers a major non-chicken peak in %s samples.",
      "This document should therefore be read as a characterization of assay behavior and recoverable structure,",
      "not as a final estimate of biological ploidy shifts."
    ),
    nrow(analysis_tbl),
    assigned_chicken_n,
    ambiguous_chicken_n,
    resolved_major_n
  )
)
```

# Next Steps

1. Manually inspect the samples flagged as ambiguous to determine whether the
   low-DNA compartment is true chicken, debris-adjacent material, or an
   acquisition artifact.
2. If a verified chicken gate can be defined from scatter plus fluorescence
   jointly, incorporate that gate explicitly rather than relying on DNA density
   alone.
3. Once the reference assignment is stable, move to lineage-aware comparisons of
   normalized structure and only then consider oxygen-associated biological
   shifts.
