---
title: "Flow Gating Workflow"
output:
  html_document:
    toc: true
    toc_float: true
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "./")
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

# Goal

This document runs the gating workflow independently for each dataset directory
under `data/`. Each dataset gets its own cache under `processed_data/` and its
own figure namespace under `figure/`, so additional datasets can be added
without colliding with existing outputs.

```{r load-js, results='asis'}
cat('<script src="https://unpkg.com/panzoom@9.4.0/dist/panzoom.min.js"></script>')
```

```{r libraries}
library(flowCore)
library(flowWorkspace)
library(PeacoQC)
library(MASS)
library(ggcyto)
library(scales)
library(knitr)
library(tibble)
library(dplyr)
library(patchwork)
```

```{r config}
data_root <- "data"
processed_root <- "processed_data"
figure_root <- "figure"

reuse_existing_figures <- TRUE
force_rebuild_figures <- FALSE

default_qc_channels <- c(
  "FSC-A", "FSC-H", "FSC-W",
  "SSC-A", "SSC-H", "SSC-W",
  "450-Violet C-A", "450-Violet C-H"
)
default_dna_channels <- c("450-Violet C-A", "450-Violet C-H")
default_dna_cofactor <- 500

sanitize_dataset_id <- function(x) {
  x <- tolower(gsub("[^A-Za-z0-9]+", "-", x))
  x <- gsub("(^-+|-+$)", "", x)
  if (!nzchar(x)) {
    stop("Could not derive a dataset id from the supplied name.")
  }
  x
}

discover_dataset_dirs <- function(root_path) {
  dirs <- list.dirs(root_path, recursive = FALSE, full.names = TRUE)
  dirs <- dirs[!startsWith(basename(dirs), ".")]
  dirs <- dirs[basename(dirs) != "cache"]
  sort(normalizePath(dirs, winslash = "/", mustWork = TRUE))
}

discover_flow_files <- function(dataset_dir) {
  files <- list.files(
    dataset_dir,
    pattern = "\\.fcs$",
    full.names = TRUE,
    recursive = TRUE,
    ignore.case = TRUE
  )
  files <- files[!grepl("(^|[\\\\/])__MACOSX([\\\\/]|$)", files)]
  files <- files[!startsWith(basename(files), "._")]
  sort(normalizePath(files, winslash = "/", mustWork = TRUE))
}

make_plot_paths <- function(figure_dir) {
  list(
    fsc_ssc = file.path(figure_dir, "fsc-ssc-plots-1.png"),
    dna_scatter = file.path(figure_dir, "dna-scatter-plots-1.png"),
    dna_density = file.path(figure_dir, "dna-density-plots-1.png")
  )
}

dir.create(processed_root, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_root, recursive = TRUE, showWarnings = FALSE)
stopifnot(dir.exists(data_root))

dataset_dirs <- discover_dataset_dirs(data_root)
stopifnot(length(dataset_dirs) > 0)

dataset_tbl <- tibble(
  dataset_name = basename(dataset_dirs),
  dataset_dir = dataset_dirs,
  dataset_id = vapply(basename(dataset_dirs), sanitize_dataset_id, character(1)),
  enabled = TRUE
)

dataset_tbl$qc_channels <- rep(list(default_qc_channels), nrow(dataset_tbl))
dataset_tbl$dna_channels <- rep(list(default_dna_channels), nrow(dataset_tbl))
dataset_tbl$dna_cofactor <- rep(default_dna_cofactor, nrow(dataset_tbl))

dataset_tbl <- dataset_tbl %>%
  filter(enabled) %>%
  mutate(
    cache_dir = file.path(processed_root, dataset_id),
    gs_cache_path = file.path(cache_dir, "gating_gs"),
    figure_dir = file.path(figure_root, dataset_id)
  )

dataset_tbl$flow_files <- lapply(dataset_tbl$dataset_dir, discover_flow_files)
dataset_tbl$n_files <- lengths(dataset_tbl$flow_files)
stopifnot(all(dataset_tbl$n_files > 0))

dataset_tbl$plot_paths <- lapply(dataset_tbl$figure_dir, make_plot_paths)
dataset_tbl$has_cache <- dir.exists(dataset_tbl$gs_cache_path)
dataset_tbl$has_all_reusable_plots <- vapply(
  dataset_tbl$plot_paths,
  function(x) all(file.exists(unlist(x))),
  logical(1)
)

dataset_configs <- lapply(seq_len(nrow(dataset_tbl)), function(i) {
  list(
    dataset_name = dataset_tbl$dataset_name[[i]],
    dataset_dir = dataset_tbl$dataset_dir[[i]],
    dataset_id = dataset_tbl$dataset_id[[i]],
    qc_channels = dataset_tbl$qc_channels[[i]],
    dna_channels = dataset_tbl$dna_channels[[i]],
    dna_cofactor = dataset_tbl$dna_cofactor[[i]],
    cache_dir = dataset_tbl$cache_dir[[i]],
    gs_cache_path = dataset_tbl$gs_cache_path[[i]],
    figure_dir = dataset_tbl$figure_dir[[i]],
    flow_files = dataset_tbl$flow_files[[i]],
    plot_paths = dataset_tbl$plot_paths[[i]],
    has_cache = dataset_tbl$has_cache[[i]],
    has_all_reusable_plots = dataset_tbl$has_all_reusable_plots[[i]]
  )
})
names(dataset_configs) <- vapply(dataset_configs, `[[`, character(1), "dataset_id")
```

# Dataset Inventory

```{r dataset-inventory}
dataset_inventory <- dataset_tbl %>%
  transmute(
    dataset_name,
    dataset_id,
    dataset_dir,
    n_files,
    has_cache,
    has_all_reusable_plots
  )

if (nrow(dataset_inventory) <= 12 && ncol(dataset_inventory) <= 6) {
  kable(dataset_inventory)
}
```

# Helper Functions

```{r helpers}
first_flowFrame <- function(x) {
  if (inherits(x, "flowFrame")) return(x)
  if (is.list(x)) {
    hits <- which(vapply(x, function(z) inherits(z, "flowFrame"), logical(1)))
    if (length(hits) > 0) return(x[[hits[1]]])
  }
  stop("Could not find a flowFrame in RemoveMargins output.")
}

validate_dna_channels <- function(dna_channels, dataset_name) {
  if (length(dna_channels) != 2) {
    stop(sprintf(
      "Dataset '%s' must define exactly two DNA channels; got %d.",
      dataset_name,
      length(dna_channels)
    ))
  }
}

validate_flowset_channels <- function(fs, required_channels, dataset_name) {
  missing_by_sample <- lapply(sampleNames(fs), function(sn) {
    setdiff(required_channels, colnames(fs[[sn]]))
  })
  names(missing_by_sample) <- sampleNames(fs)

  bad_samples <- names(missing_by_sample)[lengths(missing_by_sample) > 0]
  if (length(bad_samples) > 0) {
    detail <- paste(
      sprintf(
        "%s [%s]",
        bad_samples,
        vapply(
          missing_by_sample[bad_samples],
          function(x) paste(x, collapse = ", "),
          character(1)
        )
      ),
      collapse = "; "
    )
    stop(sprintf(
      "Dataset '%s' is missing required channels in one or more samples: %s",
      dataset_name,
      detail
    ))
  }
}

validate_cached_gs <- function(gs, flow_files, dataset_name) {
  expected_samples <- sort(basename(flow_files))
  cached_samples <- sort(basename(sampleNames(gs)))

  if (!identical(expected_samples, cached_samples)) {
    stop(sprintf(
      "Cached GatingSet for dataset '%s' does not match the current files on disk. Remove '%s' and rerender.",
      dataset_name,
      file.path(processed_root, sanitize_dataset_id(dataset_name), "gating_gs")
    ))
  }
}

dna_singlet_vec <- function(ff, a = "450-Violet C-A", h = "450-Violet C-H", nmads = 4) {
  xy <- exprs(ff)[, c(a, h), drop = FALSE]
  ok <- is.finite(xy[, 1]) & is.finite(xy[, 2]) & xy[, 1] > 0 & xy[, 2] > 0

  if (sum(ok) < 100) {
    return(rep(FALSE, nrow(xy)))
  }

  fit <- MASS::rlm(xy[ok, 2] ~ xy[ok, 1])
  pred <- coef(fit)[1] + coef(fit)[2] * xy[ok, 1]
  res <- xy[ok, 2] - pred

  spread <- mad(res, na.rm = TRUE)
  if (!is.finite(spread) || spread == 0) {
    spread <- sd(res, na.rm = TRUE)
  }
  if (!is.finite(spread) || spread == 0) {
    spread <- 1e-8
  }

  keep_ok <- abs(res - median(res, na.rm = TRUE)) <= nmads * spread
  keep <- rep(FALSE, nrow(xy))
  keep[which(ok)[keep_ok]] <- TRUE
  keep
}

find_density_peaks <- function(x, bw = "nrd0") {
  x <- x[is.finite(x)]
  if (length(x) < 2) {
    return(data.frame(x = numeric(0), y = numeric(0)))
  }

  d <- density(x, bw = bw)
  idx <- which(diff(sign(diff(d$y))) == -2) + 1L
  data.frame(x = d$x[idx], y = d$y[idx])
}

process_masks <- function(ff_raw, qc_channels, dna_channels, IT_limit = 0.65, dna_cofactor = 500, nmads = 4) {
  idx <- match(qc_channels, colnames(ff_raw))
  idx <- idx[!is.na(idx)]

  mr <- RemoveMargins(ff_raw, channels = idx, output = "full")
  ff_margin_raw <- first_flowFrame(mr)

  orig_id_margin <- as.integer(exprs(ff_margin_raw)[, "Original_ID"])
  margin_keep_global <- rep(FALSE, nrow(ff_raw))
  margin_keep_global[orig_id_margin] <- TRUE

  use_dna <- intersect(dna_channels, colnames(ff_margin_raw))
  if (length(use_dna) > 0) {
    tf <- transformList(use_dna, arcsinhTransform(a = 0, b = 1 / dna_cofactor, c = 0))
    ff_margin_proc <- transform(ff_margin_raw, tf)
  } else {
    ff_margin_proc <- ff_margin_raw
  }

  pq <- PeacoQC(
    ff = ff_margin_proc,
    channels = idx,
    determine_good_cells = "all",
    save_fcs = FALSE,
    plot = FALSE,
    report = FALSE,
    IT_limit = IT_limit
  )
  peacoqc_keep_local <- pq$GoodCells
  ff_peacoqc <- pq$FinalFF

  dna_keep_local <- dna_singlet_vec(
    ff_peacoqc,
    a = dna_channels[1],
    h = dna_channels[2],
    nmads = nmads
  )

  list(
    margin_keep = margin_keep_global,
    peacoqc_keep = peacoqc_keep_local,
    dna_keep = dna_keep_local
  )
}

build_gs <- function(flow_files, qc_channels, dna_channels, dataset_name, dna_cofactor = 500) {
  validate_dna_channels(dna_channels, dataset_name)

  fs <- read.flowSet(
    files = flow_files,
    transformation = FALSE,
    truncate_max_range = FALSE
  )
  validate_flowset_channels(fs, unique(c(qc_channels, dna_channels)), dataset_name)

  mask_list <- lapply(sampleNames(fs), function(sn) {
    process_masks(fs[[sn]], qc_channels = qc_channels, dna_channels = dna_channels, dna_cofactor = dna_cofactor)
  })
  names(mask_list) <- sampleNames(fs)

  margin_list <- setNames(lapply(mask_list, `[[`, "margin_keep"), sampleNames(fs))
  peacoqc_list <- setNames(lapply(mask_list, `[[`, "peacoqc_keep"), sampleNames(fs))
  dna_list <- setNames(lapply(mask_list, `[[`, "dna_keep"), sampleNames(fs))

  gs <- GatingSet(fs)

  dna_ch_exist <- intersect(dna_channels, colnames(gs))
  if (length(dna_ch_exist) > 0) {
    custom_arcsinh <- trans_new(
      name = "custom_arcsinh",
      transform = function(x) asinh(x / dna_cofactor),
      inverse = function(x) sinh(x) * dna_cofactor,
      breaks = extended_breaks(),
      format = scientific_format()
    )

    trans_list <- transformerList(
      dna_ch_exist,
      lapply(dna_ch_exist, function(x) custom_arcsinh)
    )
    gs <- transform(gs, trans_list)
  }

  gs_pop_add(gs, margin_list, name = "Margin_Clean", parent = "root")
  recompute(gs)

  gs_pop_add(gs, peacoqc_list, name = "PeacoQC_Clean", parent = "Margin_Clean")
  recompute(gs)

  gs_pop_add(gs, dna_list, name = "Singlets", parent = "PeacoQC_Clean")
  recompute(gs)

  gs
}

build_or_load_dataset <- function(cfg) {
  dir.create(cfg$cache_dir, recursive = TRUE, showWarnings = FALSE)
  dir.create(cfg$figure_dir, recursive = TRUE, showWarnings = FALSE)

  gs <- if (dir.exists(cfg$gs_cache_path)) {
    cached_gs <- load_gs(cfg$gs_cache_path)
    validate_cached_gs(cached_gs, cfg$flow_files, cfg$dataset_name)
    cached_gs
  } else {
    built_gs <- build_gs(
      flow_files = cfg$flow_files,
      qc_channels = cfg$qc_channels,
      dna_channels = cfg$dna_channels,
      dataset_name = cfg$dataset_name,
      dna_cofactor = cfg$dna_cofactor
    )
    save_gs(built_gs, path = cfg$gs_cache_path, backend_opt = "copy")
    built_gs
  }

  input_tbl <- tibble(
    dataset_id = cfg$dataset_id,
    dataset_name = cfg$dataset_name,
    sample_name = basename(cfg$flow_files),
    source_path = cfg$flow_files
  )

  gate_counts <- as_tibble(gs_pop_get_count_fast(gs)) %>%
    mutate(
      dataset_id = cfg$dataset_id,
      dataset_name = cfg$dataset_name,
      .before = 1
    )

  peak_table <- lapply(sampleNames(gs), function(sn) {
    ff <- gh_pop_get_data(gs[[sn]], "Singlets")
    pk <- find_density_peaks(exprs(ff)[, cfg$dna_channels[1]])
    pk <- pk[order(pk$x), , drop = FALSE]

    tibble(
      dataset_id = cfg$dataset_id,
      dataset_name = cfg$dataset_name,
      sample = sn,
      peak_x = pk$x,
      peak_density = pk$y
    )
  })
  peak_table <- bind_rows(peak_table)

  list(
    config = cfg,
    gs = gs,
    input_tbl = input_tbl,
    gate_counts = gate_counts,
    peak_table = peak_table
  )
}

should_print_table <- function(tbl, max_rows = 12, max_cols = 6) {
  nrow(tbl) <= max_rows && ncol(tbl) <= max_cols
}

dataset_input_summary <- function(input_tbl) {
  tibble(
    dataset_id = dplyr::first(input_tbl$dataset_id),
    dataset_name = dplyr::first(input_tbl$dataset_name),
    n_files = nrow(input_tbl),
    first_sample = dplyr::first(input_tbl$sample_name),
    last_sample = dplyr::last(input_tbl$sample_name)
  )
}

dataset_gate_summary <- function(gate_counts) {
  value_col <- intersect(c("count", "Count"), names(gate_counts))[1]
  pop_col <- intersect(c("pop", "Population", "population", "name"), names(gate_counts))[1]

  if (is.na(value_col) || is.na(pop_col)) {
    return(tibble(note = "Gate count schema not recognized for compact summary."))
  }

  gate_counts %>%
    group_by(.data[[pop_col]]) %>%
    summarise(
      n_samples = dplyr::n(),
      min_count = min(.data[[value_col]], na.rm = TRUE),
      median_count = median(.data[[value_col]], na.rm = TRUE),
      max_count = max(.data[[value_col]], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    rename(population = 1)
}

dataset_peak_summary <- function(peak_table) {
  if (nrow(peak_table) == 0) {
    return(tibble(note = "No peaks detected."))
  }

  peak_counts <- peak_table %>%
    count(sample, name = "n_peaks")

  tibble(
    n_samples = n_distinct(peak_table$sample),
    total_peaks = nrow(peak_table),
    min_peaks_per_sample = min(peak_counts$n_peaks),
    median_peaks_per_sample = median(peak_counts$n_peaks),
    max_peaks_per_sample = max(peak_counts$n_peaks),
    median_peak_x = median(peak_table$peak_x, na.rm = TRUE)
  )
}

render_or_include_plot <- function(plot_path, plot_builder, width, height, element_id, dpi = 300) {
  should_rebuild <- force_rebuild_figures || !reuse_existing_figures || !file.exists(plot_path)

  if (should_rebuild) {
    plot_obj <- plot_builder()
    ggplot2::ggsave(
      filename = plot_path,
      plot = plot_obj,
      width = width,
      height = height,
      dpi = dpi,
      units = "in",
      limitsize = FALSE
    )
  }

  plot_src <- gsub("\\\\", "/", plot_path)
  sprintf(
    '<img src="%s" id="%s" class="zoomable-plot" style="width:100%%;" />',
    plot_src,
    element_id
  )
}

build_fsc_ssc_plot <- function(gs) {
  p_fsc_raw <- ggcyto(gs, subset = "root", aes(x = `FSC-A`, y = `SSC-A`)) +
    geom_hex(bins = 60) +
    labs(title = "FSC vs SSC (Raw)") +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  p_fsc_filtered <- ggcyto(gs, subset = "Singlets", aes(x = `FSC-A`, y = `SSC-A`)) +
    geom_hex(bins = 60) +
    labs(title = "FSC vs SSC (Filtered)") +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  as.ggplot(p_fsc_raw) + as.ggplot(p_fsc_filtered)
}

build_dna_scatter_plot <- function(gs, dna_channels) {
  p_dna_raw <- ggcyto(gs, subset = "root", aes(x = .data[[dna_channels[1]]], y = .data[[dna_channels[2]]])) +
    geom_hex(bins = 60) +
    labs(title = "DNA Scatter (Raw)") +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  p_dna_filtered <- ggcyto(gs, subset = "Singlets", aes(x = .data[[dna_channels[1]]], y = .data[[dna_channels[2]]])) +
    geom_hex(bins = 60) +
    labs(title = "DNA Scatter (Filtered Singlets)") +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  as.ggplot(p_dna_raw) + as.ggplot(p_dna_filtered)
}

build_dna_density_plot <- function(gs, dna_channels) {
  p_density_raw <- ggcyto(gs, subset = "root", aes(x = .data[[dna_channels[1]]])) +
    geom_density(fill = "gray50", alpha = 0.5, linewidth = 0.5) +
    labs(title = "DNA Density (Raw)", y = "Density") +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  p_density_filtered <- ggcyto(gs, subset = "Singlets", aes(x = .data[[dna_channels[1]]])) +
    geom_density(fill = "gray50", alpha = 0.5, linewidth = 0.5) +
    labs(title = "DNA Density (Filtered Singlets)", y = "Density") +
    theme_bw() +
    theme(strip.text = element_text(size = 8))

  as.ggplot(p_density_raw) + as.ggplot(p_density_filtered)
}
```

# Build or Load Dataset Results

```{r dataset-results}
dataset_results <- lapply(dataset_configs, build_or_load_dataset)
dataset_counts <- bind_rows(lapply(dataset_results, `[[`, "gate_counts"))
dataset_peak_table <- bind_rows(lapply(dataset_results, `[[`, "peak_table"))
```

# Cross-Dataset Gate Counts

```{r gate-counts}
dataset_gate_overview <- bind_rows(lapply(dataset_results, function(res) {
  gate_summary <- dataset_gate_summary(res$gate_counts)
  gate_summary$dataset_id <- res$config$dataset_id
  gate_summary
})) %>%
  select(dataset_id, everything())

if (should_print_table(dataset_gate_overview)) {
  kable(dataset_gate_overview)
}
```

# Cross-Dataset DNA Peak Summary

```{r peak-summary}
dataset_peak_overview <- bind_rows(lapply(dataset_results, function(res) {
  peak_summary <- dataset_peak_summary(res$peak_table)
  peak_summary$dataset_id <- res$config$dataset_id
  peak_summary
})) %>%
  select(
    dataset_id,
    n_samples,
    total_peaks,
    min_peaks_per_sample,
    median_peaks_per_sample,
    median_peak_x
  )

if (should_print_table(dataset_peak_overview)) {
  kable(dataset_peak_overview)
}
```

# Dataset Details

```{r dataset-sections, results='asis'}
for (res in dataset_results) {
  cfg <- res$config
  gs <- res$gs

  cat(sprintf("## %s\n\n", cfg$dataset_name))
  cat(sprintf("- Dataset id: `%s`\n", cfg$dataset_id))
  cat(sprintf("- Source directory: `%s`\n", cfg$dataset_dir))
  cat(sprintf("- Cache path: `%s`\n", cfg$gs_cache_path))
  cat(sprintf("- Figure directory: `%s`\n\n", cfg$figure_dir))

  cat("### Input Summary\n\n")
  print(kable(dataset_input_summary(res$input_tbl)))
  cat("\n")

  gate_summary <- dataset_gate_summary(res$gate_counts)
  if (should_print_table(gate_summary)) {
    cat("### Gate Count Summary\n\n")
    print(kable(gate_summary))
    cat("\n")
  }

  peak_summary <- dataset_peak_summary(res$peak_table)
  if (should_print_table(peak_summary)) {
    cat("### DNA Peak Summary\n\n")
    print(kable(peak_summary))
    cat("\n")
  }

  if (should_print_table(res$input_tbl)) {
    cat("### Input Detail\n\n")
    print(kable(res$input_tbl))
    cat("\n")
  }

  if (should_print_table(res$gate_counts)) {
    cat("### Gate Count Detail\n\n")
    print(kable(res$gate_counts))
    cat("\n")
  }

  if (should_print_table(res$peak_table)) {
    cat("### DNA Peak Detail\n\n")
    print(kable(res$peak_table))
    cat("\n")
  }
  cat("\n")

  cat("### FSC vs SSC\n\n")
  cat(
    render_or_include_plot(
      plot_path = cfg$plot_paths$fsc_ssc,
      width = 21,
      height = 14,
      dpi = 300,
      element_id = sprintf("%s-fsc-ssc-plot", cfg$dataset_id),
      plot_builder = function() build_fsc_ssc_plot(gs)
    )
  )
  cat("\n\n")

  cat("### DNA Scatter\n\n")
  cat(
    render_or_include_plot(
      plot_path = cfg$plot_paths$dna_scatter,
      width = 21,
      height = 14,
      dpi = 300,
      element_id = sprintf("%s-dna-scatter-plot", cfg$dataset_id),
      plot_builder = function() build_dna_scatter_plot(gs, cfg$dna_channels)
    )
  )
  cat("\n\n")

  cat("### DNA Density\n\n")
  cat(
    render_or_include_plot(
      plot_path = cfg$plot_paths$dna_density,
      width = 21,
      height = 14,
      dpi = 300,
      element_id = sprintf("%s-dna-density-plot", cfg$dataset_id),
      plot_builder = function() build_dna_density_plot(gs, cfg$dna_channels)
    )
  )
  cat("\n\n")
}
```

```{r activate-zoom, results='asis'}
cat('
<script>
(function() {
  Array.prototype.forEach.call(document.querySelectorAll(".zoomable-plot"), function(el) {
    if (!el || typeof panzoom === "undefined") return;
    panzoom(el, { maxZoom: 10, minZoom: 1 });
  });
})();
</script>
')
```
