---
title: "Ploidy-Oxygen Workflow: Data Processing and Initial Modelling"
output:
  html_document:
    toc: true
    toc_float: true
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "./")
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%"
)
```

# Workflow Goal

This document turns the raw `.fcs` files into a first-pass set of quality-controlled
DNA-content summaries. The immediate objective is to identify the internal chicken
reference peak and the tumor G0/G1 peak in each sample, summarize their relative
positions, and test whether oxygen condition shifts the normalized DNA-content signal.

# Input Configuration

```{r libraries}
library(flowCore)
library(ggplot2)
library(knitr)
library(scales)
```

```{r input-config}
path <- "data/Hypoxia_SUM159"
stopifnot(dir.exists(path))

dna_channel <- "X450.Violet.C.A"
scatter_channels <- c("FSC.A", "SSC.A", "FSC.H")

ff <- list.files(path, pattern = "\\.fcs$", full.names = FALSE)
stopifnot(length(ff) > 0)
```

# Sample Metadata Parsing

```{r sample-metadata}
parse_sample_name <- function(filename) {
  core <- sub("^Sample_", "", filename)
  core <- sub("\\.fcs$", "", core)
  parts <- strsplit(core, "_", fixed = TRUE)[[1]]

  replicate_raw <- parts[4]
  replicate_id <- sub("_+$", "", replicate_raw)

  data.frame(
    file_name = filename,
    sample_id = core,
    cell_line = parts[1],
    starting_ploidy = parts[2],
    oxygen_condition = parts[3],
    replicate_id = replicate_id,
    stringsAsFactors = FALSE
  )
}

sample_meta <- do.call(
  rbind,
  lapply(ff, parse_sample_name)
)

sample_meta$starting_ploidy <- factor(sample_meta$starting_ploidy, levels = c("2N", "4N"))
sample_meta$oxygen_condition <- factor(sample_meta$oxygen_condition, levels = c("C", "O1", "O2"))

kable(sample_meta, caption = "Parsed sample metadata from FCS filenames.")
```

# Flow Import and Quality Control

```{r helper-functions}
read_flow_table <- function(file_name) {
  fcs <- read.FCS(file.path(path, file_name))
  dfi <- as.data.frame(exprs(fcs))
  colnames(dfi) <- make.names(colnames(dfi))
  dfi
}

apply_qc_gate <- function(dfi) {
  keep <- rep(TRUE, nrow(dfi))

  for (channel in c(dna_channel, scatter_channels)) {
    x <- dfi[[channel]]
    keep <- keep & is.finite(x)
  }

  for (channel in c(dna_channel, "FSC.A", "SSC.A")) {
    x <- dfi[[channel]]
    bounds <- quantile(x, probs = c(0.01, 0.99), na.rm = TRUE)
    keep <- keep & x >= bounds[1] & x <= bounds[2]
  }

  singlet_ratio <- dfi$FSC.H / pmax(dfi$FSC.A, 1)
  ratio_bounds <- quantile(singlet_ratio[is.finite(singlet_ratio)], probs = c(0.05, 0.95), na.rm = TRUE)
  keep <- keep & is.finite(singlet_ratio) & singlet_ratio >= ratio_bounds[1] & singlet_ratio <= ratio_bounds[2]

  dfi[keep, , drop = FALSE]
}

find_density_peaks <- function(x, n = 4) {
  x <- x[is.finite(x)]
  if (length(unique(x)) < 2) {
    return(data.frame(x = numeric(0), y = numeric(0)))
  }

  dens <- density(x, n = 2048, adjust = 1.2)
  idx <- which(diff(sign(diff(dens$y))) == -2) + 1
  peaks <- data.frame(x = dens$x[idx], y = dens$y[idx])
  peaks <- peaks[order(peaks$y, decreasing = TRUE), , drop = FALSE]

  if (nrow(peaks) == 0) {
    return(peaks)
  }

  head(peaks, n)
}

summarize_sample <- function(file_name) {
  raw_df <- read_flow_table(file_name)
  qc_df <- apply_qc_gate(raw_df)

  peaks <- find_density_peaks(qc_df[[dna_channel]], n = 4)
  peaks <- peaks[order(peaks$x), , drop = FALSE]

  chicken_peak <- if (nrow(peaks) >= 1) peaks$x[1] else NA_real_
  tumor_g1_peak <- if (nrow(peaks) >= 2) peaks$x[2] else NA_real_
  secondary_peak <- if (nrow(peaks) >= 3) peaks$x[3] else NA_real_
  normalized_ratio <- tumor_g1_peak / chicken_peak

  data.frame(
    file_name = file_name,
    events_raw = nrow(raw_df),
    events_qc = nrow(qc_df),
    qc_retention = nrow(qc_df) / nrow(raw_df),
    peak_count = nrow(peaks),
    chicken_peak = chicken_peak,
    tumor_g1_peak = tumor_g1_peak,
    secondary_peak = secondary_peak,
    tumor_to_chicken_ratio = normalized_ratio,
    qc_flag_low_events = nrow(qc_df) < 1000,
    qc_flag_missing_peaks = nrow(peaks) < 2,
    stringsAsFactors = FALSE
  )
}
```

```{r flow-import}
flow_raw <- setNames(lapply(ff, read_flow_table), ff)
flow_qc <- setNames(lapply(flow_raw, apply_qc_gate), ff)

qc_overview <- do.call(
  rbind,
  lapply(names(flow_raw), function(file_name) {
    data.frame(
      file_name = file_name,
      events_raw = nrow(flow_raw[[file_name]]),
      events_qc = nrow(flow_qc[[file_name]]),
      qc_retention = nrow(flow_qc[[file_name]]) / nrow(flow_raw[[file_name]])
    )
  })
)

kable(qc_overview, digits = 3, caption = "Event counts before and after a conservative exploratory QC gate.")
```

# Quality-Control Visualizations

```{r qc-example-plot}
example_file <- sample_meta$file_name[1]
example_raw <- flow_raw[[example_file]]
example_qc <- flow_qc[[example_file]]

ggplot(example_raw, aes(x = FSC.A, y = SSC.A)) +
  geom_point(alpha = 0.08, size = 0.3, color = "#9c6644") +
  geom_point(
    data = example_qc,
    aes(x = FSC.A, y = SSC.A),
    alpha = 0.08,
    size = 0.3,
    color = "#386641"
  ) +
  labs(
    title = paste("QC overlay for", example_file),
    subtitle = "Raw events in brown, retained events in green",
    x = "FSC-A",
    y = "SSC-A"
  ) +
  theme_minimal(base_size = 12)
```

# Peak Detection on DNA Channel

```{r peak-summary}
peak_summary <- do.call(
  rbind,
  lapply(ff, summarize_sample)
)

analysis_tbl <- merge(sample_meta, peak_summary, by = "file_name", sort = FALSE)
analysis_tbl <- analysis_tbl[order(analysis_tbl$starting_ploidy, analysis_tbl$oxygen_condition, analysis_tbl$replicate_id), ]

kable(
  analysis_tbl,
  digits = 3,
  caption = "Per-sample DNA peak summary after QC filtering and heuristic density-peak calling."
)
```

```{r example-density}
example_peaks <- find_density_peaks(example_qc[[dna_channel]], n = 4)

ggplot(example_qc, aes(x = .data[[dna_channel]])) +
  geom_density(fill = "#8ecae6", alpha = 0.5, color = "#023047", linewidth = 0.8) +
  geom_vline(
    data = example_peaks,
    aes(xintercept = x),
    color = "#d00000",
    linetype = "dashed",
    linewidth = 0.7
  ) +
  labs(
    title = paste("DNA density with called peaks for", example_file),
    subtitle = "Dashed lines mark the strongest local maxima",
    x = "450-Violet C-A",
    y = "Density"
  ) +
  theme_minimal(base_size = 12)
```

```{r all-sample-density-grid, fig.height=14, fig.width=16}
sample_order <- analysis_tbl$file_name

density_list <- lapply(sample_order, function(file_name) {
  dens <- density(flow_qc[[file_name]][[dna_channel]], n = 2048, adjust = 1.2)
  data.frame(
    file_name = file_name,
    dna_value = dens$x,
    density_value = dens$y
  )
})

peak_list <- lapply(sample_order, function(file_name) {
  peaks <- find_density_peaks(flow_qc[[file_name]][[dna_channel]], n = 4)
  if (nrow(peaks) == 0) {
    return(data.frame(
      file_name = character(0),
      peak_position = numeric(0),
      peak_height = numeric(0)
    ))
  }

  data.frame(
    file_name = file_name,
    peak_position = peaks$x,
    peak_height = peaks$y
  )
})

sample_density_tbl <- do.call(rbind, density_list)
sample_peak_tbl <- do.call(rbind, peak_list)

sample_density_tbl$file_name <- factor(sample_density_tbl$file_name, levels = sample_order)
sample_peak_tbl$file_name <- factor(sample_peak_tbl$file_name, levels = sample_order)

facet_labels <- setNames(analysis_tbl$sample_id, analysis_tbl$file_name)

ggplot(sample_density_tbl, aes(x = dna_value, y = density_value)) +
  geom_area(fill = "#8ecae6", alpha = 0.5) +
  geom_line(color = "#023047", linewidth = 0.5) +
  geom_vline(
    data = sample_peak_tbl,
    aes(xintercept = peak_position),
    color = "#d00000",
    linetype = "dashed",
    linewidth = 0.4
  ) +
  facet_wrap(
    ~file_name,
    ncol = 4,
    nrow = 5,
    scales = "free_y",
    labeller = as_labeller(facet_labels)
  ) +
  labs(
    title = "DNA intensity with called peaks for every flow sample",
    subtitle = "Each panel shows the QC-filtered DNA density; dashed lines mark called local maxima",
    x = "450-Violet C-A",
    y = "Density"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    strip.text = element_text(size = 8),
    panel.grid.minor = element_blank()
  )
```

# Summary Plots

```{r summary-plots}
ggplot(
  analysis_tbl,
  aes(x = oxygen_condition, y = tumor_to_chicken_ratio, color = starting_ploidy)
) +
  geom_point(position = position_jitter(width = 0.08, height = 0), size = 2.4, alpha = 0.9) +
  stat_summary(aes(group = starting_ploidy), fun = mean, geom = "line", linewidth = 0.8) +
  facet_wrap(~starting_ploidy, nrow = 1) +
  scale_color_manual(values = c("2N" = "#1d3557", "4N" = "#d62828")) +
  scale_y_continuous(labels = number_format(accuracy = 0.01)) +
  labs(
    title = "Normalized tumor DNA content across oxygen conditions",
    x = "Oxygen condition",
    y = "Tumor G0/G1 peak divided by chicken reference peak",
    color = "Starting ploidy"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

```{r summary-plot-peaks}
plot_tbl <- rbind(
  data.frame(
    sample_id = analysis_tbl$sample_id,
    starting_ploidy = analysis_tbl$starting_ploidy,
    oxygen_condition = analysis_tbl$oxygen_condition,
    peak_type = "Chicken reference",
    peak_value = analysis_tbl$chicken_peak
  ),
  data.frame(
    sample_id = analysis_tbl$sample_id,
    starting_ploidy = analysis_tbl$starting_ploidy,
    oxygen_condition = analysis_tbl$oxygen_condition,
    peak_type = "Tumor G0/G1",
    peak_value = analysis_tbl$tumor_g1_peak
  )
)

ggplot(plot_tbl, aes(x = oxygen_condition, y = peak_value, fill = peak_type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.75), size = 1.8, alpha = 0.9) +
  facet_wrap(~starting_ploidy, nrow = 1) +
  scale_fill_manual(values = c("Chicken reference" = "#f4a261", "Tumor G0/G1" = "#2a9d8f")) +
  labs(
    title = "Peak positions by oxygen condition",
    x = "Oxygen condition",
    y = "Peak position on DNA channel",
    fill = "Peak type"
  ) +
  theme_minimal(base_size = 12)
```

# Initial Model

```{r initial-model}
model_tbl <- subset(
  analysis_tbl,
  is.finite(tumor_to_chicken_ratio) & !qc_flag_missing_peaks
)

initial_model <- lm(
  tumor_to_chicken_ratio ~ starting_ploidy * oxygen_condition,
  data = model_tbl
)

model_coef <- data.frame(
  term = names(coef(initial_model)),
  estimate = unname(coef(initial_model)),
  stringsAsFactors = FALSE
)

kable(
  model_coef,
  digits = 4,
  caption = "Coefficient table from an exploratory linear model of normalized tumor DNA content."
)
```

```{r model-summary-text, results='asis'}
cat(
  sprintf(
    paste(
      "The exploratory model was fit on %s samples with usable peak calls.",
      "This should be treated as a workflow check rather than a final biological conclusion."
    ),
    nrow(model_tbl)
  )
)
```

# Next Steps

1. Replace the heuristic QC gate with explicit debris and singlet gates derived from the instrument channels and known control populations.
2. Validate the peak annotations against manual inspection for a subset of samples before using the ratios in downstream inference.
3. Extend the model to work on full distributions rather than just the dominant G0/G1 peak, especially if S-phase and 4N fractions become biologically important.
